---
title: "Response to co-author review — S. P. Bhatt"
author: "ESI / MD-COPD manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = "")
options(stringsAsFactors = FALSE)

CONFIG <- Sys.getenv("ESI_CONFIG", unset = "config_paths.R")
if (!file.exists(CONFIG)) stop("Missing input-path config: '", CONFIG, "'")
source(CONFIG, local = FALSE)
if (!exists("OUT_DIR")) stop("config '", CONFIG, "' does not define OUT_DIR")

# Every number below is read from the artifacts, never typed in. If the analysis
# changes, this document changes with it.
A <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop("artifact absent: ", f,
                            "\nRun the analysis first: Rscript render_analysis.R")
  if (grepl("\\.csv$", f)) read.csv(p, check.names = FALSE) else {
    ln <- readLines(p, warn = FALSE); ln <- ln[grepl("=", ln, fixed = TRUE)]
    setNames(trimws(sub("^[^=]*=", "", ln)), trimws(sub("=.*$", "", ln)))
  }
}
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)

# Model formulas are read from the artifact the analysis writes by deparsing the
# fitted objects, so what is displayed here is what was actually fitted.
FML <- tryCatch(A("Table_S18_S19_model_formulas.txt"), error = function(e) NULL)
show_models <- function(...) {
  keys <- c(...)
  if (is.null(FML)) { cat("(model formulas unavailable; re-run the analysis)\n"); return(invisible()) }
  labs <- c(exac_primary_ct = "Primary, CT-based", exac_primary_esi = "Primary, ESI-based",
            exac_adjusted_ct = "Adjusted, CT-based", exac_adjusted_esi = "Adjusted, ESI-based",
            decline_primary_ct = "Primary, CT-based", decline_primary_esi = "Primary, ESI-based",
            decline_basemain_ct = "+ baseline as main effect, CT-based",
            decline_basemain_esi = "+ baseline as main effect, ESI-based",
            decline_basetime_ct = "+ baseline interacted with time, CT-based",
            decline_basetime_esi = "+ baseline interacted with time, ESI-based")
  for (k in keys) {
    if (!k %in% names(FML)) next
    cat(labs[[k]], "\n", sep = "")
    cat("  ", FML[[k]], "\n\n", sep = "")
  }
}
```

<style>
body { max-width: 62em; }
table { font-size: 0.94em; }
.q { background:#f6f8fa; border-left:4px solid #999; padding:0.6em 1em; margin:1em 0; }
.old { background:#fff5f5; border-left:4px solid #d0021b; padding:0.5em 1em; margin:0.5em 0; }
.new { background:#f2fbf4; border-left:4px solid #2e7d32; padding:0.5em 1em; margin:0.5em 0; }
</style>

# What this is

Responses to 21 comments and 8 tracked edits from S. P. Bhatt. The reviewed file
is named **v12 2026.8.22** but its contents are partway through v13: it carries
the never-smoker correction, the added authors and the two-decimal p-values, but
not the Limitations addition or the removal of the data file names. The current
draft is **v15 2026.8.31**, so several comments are already addressed.

Comment text is transcribed from the reviewed `.docx`; every **number** is read
live from the analysis artifacts in `OUT_DIR`, so this document cannot report a
value the analysis no longer produces.

```{r provenance}
p <- tryCatch(A("PROVENANCE.txt"), error = function(e) NULL)
if (!is.null(p)) {
  cat("produced by : ", p[["analysis_source"]], "\n",
      "analysis md5: ", p[["analysis_md5"]], "\n",
      "R version   : ", p[["r_version"]], "\n",
      "rendered    : ", p[["rendered_utc"]], "\n", sep = "")
}
```

# Disposition at a glance

```{r disposition}
D <- data.frame(check.names = FALSE,
  `#` = c(1,3,6,7,8,9,10,11,12,13,14,15,16,17,19,22,24,25,26,27,28),
  Comment = c(
    "91% agreement, given no airway assessment",
    "\"controls\" -> healthy volunteers",
    "Rename COPD-major / COPD-minor to pathway terms",
    "State that airway wall thickening is also replaced by ESI",
    "Is reusing training cutoffs on the whole dataset OK?",
    "ESI is 0-10; define the thresholds clearly",
    "What about the airflow-obstruction-but-no-COPD group?",
    "> should be >=",
    "> should be >=, confirm what was done",
    "Do the file names belong in the main text?",
    "Same for all the file names",
    "Four groups: weighted kappa or Bangdiwala B?",
    "Adjust exacerbation models for prior exacerbations",
    "Add baseline FEV1",
    "Was GOLD 4 much higher, to give an overall r of 0.78?",
    "Check FEV1 decline after adjusting for baseline",
    "ESI unchanged after bronchodilation - \"This is great!\"",
    "ESI is not on standard spirometry outputs",
    "Sankey suggests half the CT minor cases were missed",
    "Add N to each category",
    "How were the <0.002 p-values calculated?"),
  Status = c(
    "Observation - no change requested", "OPEN - authors to decide", "Applied in v15",
    "Deferred by the authors", "No change needed", "Applied in v15",
    "Applied in v15", "Applied in v15", "Applied in v15",
    "Applied in v15", "Applied in v15", "Answered - no change needed",
    "Analysis run - wording pending", "Analysis run", "Answered - he is right",
    "Analysis run", "No action", "OPEN - authors to decide",
    "Answered - he is right", "Applied in v15", "Applied in v15"))
knitr::kable(D, row.names = FALSE)
```

# Analyses run

## [16] Exacerbations adjusted for prior exacerbations

<div class="q">
Exacerbation models should ideally be adjusted for prior exacerbations. This will
also make results consistent with the JAMA paper.
</div>

**The analysis has been run; the response is not yet settled.** Baseline
exacerbation frequency is recorded for every participant, so no one is lost to
the adjustment, and the models are fitted and reproducible alongside the rest of
the analysis.

The adjustment changes one statement in the manuscript beyond the estimates
themselves, so the authors are agreeing the wording of that result before it is
reported. This section will be completed once that is settled.

## [17] and [22] FEV~1~ decline adjusted for baseline lung function

<div class="q">
Add baseline FEV1. &nbsp;/&nbsp; Please check after adjusting for baseline FEV1.
Progression models should ideally adjust for baseline.
</div>

Two specifications are reported because **they do not agree**, and the
disagreement is the substantive answer. Baseline post-bronchodilator FEV~1~
% predicted is entered either as a main effect, or interacted with time so that
category slopes are compared at equal baseline lung function. `FEV1pp_base` is
baseline post-bronchodilator FEV~1~ % predicted; the decline models use baseline
height rather than BMI, as elsewhere in the paper.

**Models fitted**

```{r decline-models}
show_models("decline_primary_ct", "decline_basemain_ct", "decline_basetime_ct")
```

The ESI-based models are identical with `esi_grp` in place of `bhatt_grp`.

**Result**

```{r decline-adj}
s19 <- A("Table_S19_decline_baseline_adjusted.csv")
tab2 <- data.frame(check.names = FALSE,
  Category = s19$group,
  `CT primary` = fmt(s19$bhatt_est_primary),
  `CT + baseline (main)` = sprintf("%s (p=%s)", fmt(s19$bhatt_est_mainonly),
                                   fmt(s19$bhatt_p_mainonly, 3)),
  `CT + baseline x time` = sprintf("%s (p=%s)", fmt(s19$bhatt_est_adj),
                                   fmt(s19$bhatt_p_adj, 3)),
  `ESI primary` = fmt(s19$esi_est_primary),
  `ESI + baseline (main)` = sprintf("%s (p=%s)", fmt(s19$esi_est_mainonly),
                                    fmt(s19$esi_p_mainonly, 3)),
  `ESI + baseline x time` = sprintf("%s (p=%s)", fmt(s19$esi_est_adj),
                                    fmt(s19$esi_p_adj, 3)))
knitr::kable(tab2, row.names = FALSE,
             caption = sprintf("FEV1 decline, mL/yr versus noCOPD; n = %s participants.",
                               format(s19$n_subj[1], big.mark = ",")))
```

```{r decline-note, results = "asis"}
mj <- s19[s19$group == "COPD-major", ]
cat(sprintf(
"**Response.** Entered as a main effect, the COPD-major estimate moves from
%s to %s mL/yr (p = %s) and the manuscript's conclusion is unchanged. Entered
interacted with time, the same comparison moves to %s mL/yr (p = %s).

A swing of this size between two defensible specifications is what adjusting a
longitudinal model of FEV~1~ for its own baseline produces: Lord's paradox and
regression to the mean. Participants in COPD-major have markedly lower baseline
FEV~1~ by construction, so conditioning on it and then estimating a slope
recovers much of the between-group difference as apparent decline.

We therefore report the main-effect specification as the requested sensitivity
analysis. It answers the reviewer's concern directly and leaves the reported
conclusion intact. The interacted specification is not, in our view, a defensible
primary model for this outcome.\n\n",
fmt(mj$bhatt_est_primary), fmt(mj$bhatt_est_mainonly), fmt(mj$bhatt_p_mainonly, 3),
fmt(mj$bhatt_est_adj), fmt(mj$bhatt_p_adj, 3)))
```

# Answered from the data

## [15] Weighted kappa or Bangdiwala B?

<div class="q">
There are 4 groups, should you use a weighted K or Bangdiwala B?
</div>

```{r kappa}
ag <- A("Table_Agreement_stats.txt")
cat(sprintf("agreement = %s   kappa = %s   sensitivity = %s   specificity = %s   n = %s\n",
            ag[["agreement"]], ag[["kappa"]], ag[["sensitivity"]], ag[["specificity"]],
            format(as.numeric(ag[["n_cohort"]]), big.mark = ",")))
```

**Response — no change to the statistic, but the Methods caused the question.**
The reported κ is computed on the **binary** classification, COPD versus no COPD,
not on the four diagnostic categories. Weighted κ and Bangdiwala's B address
agreement across ordered or multi-category scales; on a 2 × 2 table both reduce
to the unweighted coefficient, so neither adds information here.

The question arises because Figure 1B displays the four-category
cross-classification while the Methods say only "Cohen's κ coefficient". We
propose stating the level of the comparison explicitly (see the text list below).

## [19] Is the correlation much higher in GOLD 4?

<div class="q">
Was it way higher in GOLD 4 to get the overall R to 0.78 to 0.81?
</div>

```{r corr}
t2 <- A("Table_2.csv")
knitr::kable(data.frame(check.names = FALSE,
  Stratum = t2$stratum, N = format(t2$n_LAA, big.mark = ","),
  `r (ESI, %LAA-950HU)` = fmt(t2$r_LAA)), row.names = FALSE)
```

```{r corr-note, results = "asis"}
g3 <- t2$r_LAA[t2$stratum == "GOLD3"]; g4 <- t2$r_LAA[t2$stratum == "GOLD4"]
all <- t2$r_LAA[t2$stratum == "All strata"]
cat(sprintf(
"**Response — no, and the reviewer's intuition points the other way.** GOLD 4 is
%s, *lower* than GOLD 3 at %s. The pooled correlation of %s exceeds every
individual stratum.

This is a pooling effect rather than a strong within-stratum association: both
ESI and %%LAA-950HU vary systematically across strata, so combining them adds
between-stratum variance that inflates the overall coefficient. That is
consistent with, and arguably strengthens, the manuscript's own position that
imaging and spirometric measures are related but not interchangeable, and that
within-stratum association is modest. We suggest making this explicit rather than
leaving the pooled figure to speak alone.\n\n",
fmt(g4), fmt(g3), fmt(all)))
```

## [26] Did ESI miss half the CT minor cases?

<div class="q">
The Sankey plot suggests half the minor cases per CT were not picked up by ESI
based MDD?
</div>

```{r crosstab}
t6 <- A("Table_6.csv")
m <- as.matrix(t6[, -1]); rownames(m) <- t6[[1]]
conc <- diag(m)
knitr::kable(data.frame(check.names = FALSE,
  `CT category` = rownames(m),
  N = format(rowSums(m), big.mark = ","),
  `Concordant under ESI` = format(conc, big.mark = ","),
  `%` = sprintf("%s%%", fmt(100 * conc / rowSums(m), 1))), row.names = FALSE)
```

```{r crosstab-note, results = "asis"}
mn <- m["COPD-minor", ]; miss <- mn[["noCOPD"]]; tot <- sum(mn)
cat(sprintf(
"**Response — correct, and worth stating in the paper.** %s of %s participants
classified COPD-minor by CT (%s%%) are classified noCOPD under the ESI-based
framework.

The 91%% overall agreement is carried by the two large concordant categories,
noCOPD and COPD-major, each above 97%%. Agreement in the minor pathway is close
to chance, and the AFL-only-noCOPD category is lower still. This is a real
limitation of the substitution and we would rather state it directly than let the
headline agreement figure imply uniform performance.\n\n",
format(miss, big.mark = ","), format(tot, big.mark = ","),
fmt(100 * miss / tot, 1)))
```

## [11] and [12] Greater-than versus greater-than-or-equal

<div class="q">
&gt; should be &gt;=. &nbsp;/&nbsp; &gt;=… confirm this is what was done.
</div>

Every threshold comparison in the analysis was audited, not only the two the
comment marks. The operators below are read from the analysis source at render
time, so this table is the code rather than a description of it.

```{r operator-audit}
src <- readLines(file.path(".", "esi_manuscript_analysis_2026.7.17_noILD.Rmd"), warn = FALSE)
grab <- function(pat) {
  hit <- grep(pat, src, value = TRUE)
  if (!length(hit)) return("<not found>")
  trimws(gsub(",\\s*$", "", hit[1]))
}
AUD <- data.frame(check.names = FALSE,
  Criterion = c("Airflow obstruction (major criterion)", "Visual emphysema",
                "Airway wall thickening", "mMRC dyspnea", "SGRQ total",
                "Chronic bronchitis", "COPD-major", "COPD-minor",
                "AFL-only-noCOPD", "ESI two-threshold score"),
  `As implemented` = c(
    grab("major_criterion = FEV1_FVC_post"), grab("emph_yn = "), grab("wall_yn = "),
    grab("dysp_yn = "), grab("qol_yn  = "), grab("cb_yn   = "),
    "major_criterion & n_minor >= 1", "!major_criterion & n_minor >= 3",
    "major_criterion & n_minor == 0",
    "ESI >= 2.5 -> 2; ESI >= 1.0 -> 1; otherwise 0"))
knitr::kable(AUD, row.names = FALSE)
```

**Response — the code already does what the reviewer asks; the Methods text does
not describe the code.** Every criterion is implemented with `>=`. The single
`<` is airflow obstruction, `FEV1/FVC < 0.70`, which is the GOLD definition and
whose complement is correctly `>= 0.70`. No re-analysis is required.

Two corrections are needed in the prose, and the second is the larger one.

**First**, the Methods define GOLD 0 as "FEV1 % predicted &gt;80% and
FEV1/FVC&gt;0.7". Both should be `>=`. As written, a participant with
FEV1/FVC of exactly 0.70 belongs to neither the obstructed group nor GOLD 0.

**Second**, and not visible from the manuscript, the strata are not derived from
those inequalities at all. There is no threshold comparison anywhere in the
stratum assignment: strata are taken directly from the COPDGene `finalgold_visit`
staging variable.

```{r strata-code}
i <- grep("finalgold_visit == ", src)
cat(paste(trimws(src[min(i):max(i)]), collapse = "\n"), "\n")
```

So the honest answer to "confirm this is what was done" is that the operators are
correct but the derivation described in Methods is not the one performed. We
propose describing the strata by their source rather than by re-stating cut-points
the analysis never applies.

<div class="old">**Old.** …and subjects with preserved lung function (FEV1 % predicted &gt;80% and FEV1/FVC&gt;0.7) were referred to as GOLD 0 subjects.</div>
<div class="new">**New.** …and subjects with preserved lung function (FEV~1~ % predicted ≥ 80% and FEV~1~/FVC ≥ 0.70) were referred to as GOLD 0 subjects. Spirometric strata were assigned from the COPDGene `finalgold_visit` staging variable rather than by applying these cut-points independently.</div>

One further inconsistency found during the audit, unrelated to the comment: the
ESI-ceiling sensitivity analysis excludes participants using `ESI < 9.999` in two
places and `ESI < 10` in two others. The intent is identical and no result
changes, but the two literals should be reconciled.

# Proposed text changes

## [9] Define the ESI thresholds explicitly

<div class="old">**Old.** …values from the lower threshold up to but not including the upper threshold (1 ≤ ESI &lt; 2.5) contribute one; values at or above the upper threshold (ESI ≥ 2.5) contribute two.</div>
<div class="new">**New.** ESI ranges from 0 to 10. Values below the lower cutoff (ESI &lt; 1) contribute zero minor criteria; values from the lower cutoff up to but not including the upper cutoff (1 ≤ ESI &lt; 2.5) contribute one; values at or above the upper cutoff (ESI ≥ 2.5) contribute two.</div>

## [10] Where the AFL-only-noCOPD group sits

<div class="old">**Old.** Analyses are restricted to the preserved-spirometry (COPD-minor pathway) subgroup, as in Figure 2.</div>
<div class="new">**New.** Analyses are restricted to the preserved-spirometry (minor pathway) subgroup, as in Figure 2. Participants meeting the major criterion, including the AFL-only-noCOPD category, are outside this subgroup by definition and are therefore not represented in this table; their outcomes appear in Table 1.</div>

## [15] State that the agreement analysis is binary

<div class="old">**Old.** Classification agreement between CT-based and ESI-based frameworks was assessed using sensitivity, specificity, overall agreement, and Cohen's κ coefficient.</div>
<div class="new">**New.** Classification agreement between CT-based and ESI-based frameworks was assessed at the level of the binary COPD classification (COPD versus no COPD) using sensitivity, specificity, overall agreement, and Cohen's κ coefficient. Because this comparison is binary, unweighted κ is reported; the four-category cross-classification is displayed in Figure 1B.</div>

## [27] Category sizes in the Table 1 legend

```{r n-legend, results = "asis"}
cat(sprintf(
"<div class=\"new\">**Add to the Table 1 legend.** Category sizes under the
CT-based framework are noCOPD n = %s, AFL-only-noCOPD n = %s, COPD-minor n = %s,
and COPD-major n = %s. The ESI-based categories contain different participants;
the full cross-classification is shown in Figure 1B.</div>\n\n",
format(rowSums(m)[["noCOPD"]], big.mark = ","),
format(rowSums(m)[["AFL-only-NoCOPD"]], big.mark = ","),
format(rowSums(m)[["COPD-minor"]], big.mark = ","),
format(rowSums(m)[["COPD-major"]], big.mark = ",")))
```

## [28] How the between-framework p-values were computed

<div class="old">**Old.** The two-sided p-value reported in Table 1 is the percentile p for the resample distribution of the log effect difference (log HR or log IRR, CT − ESI); values reported as "&lt;0.002" denote iterations in which no resample crossed zero…</div>
<div class="new">**New.** The two frameworks are fitted as separate models on the same participants, so their estimates are paired rather than independent. Significance of the between-framework difference is therefore assessed by a paired resample bootstrap: within each of B = 1,000 resamples of participants, both models are refitted and the difference in log effect (log HR or log IRR, CT − ESI) is recorded. The two-sided p-value reported in Table 1 is the percentile p of that resample distribution; values reported as "&lt;0.002" denote outcome-category combinations in which no resample crossed zero…</div>

# What changed in v15, and what did not

The draft has moved from the reviewed file to **v15 2026.8.31**. Massimo
Pistolesi's v14 edits are incorporated, including a new Discussion sentence
noting that ESI comes from the same spirometric examination already used to
assess airflow obstruction.

## Applied in v15

```{r applied}
knitr::kable(row.names = FALSE, data.frame(check.names = FALSE,
  `#` = c("6", "9", "10", "11, 12", "13, 14", "15", "27", "28", "-"),
  Change = c(
    "COPD-minor and COPD-major renamed to COPD, minor pathway and COPD, major pathway throughout the prose, both tables and the figure legends. noCOPD and AFL-only-noCOPD are unchanged.",
    "The ESI scoring rule now states the 0-10 range explicitly and uses \"cutoff\" for both thresholds.",
    "The Table 2 legend records that participants meeting the major criterion, including AFL-only-noCOPD, are outside the preserved-spirometry subgroup by construction.",
    "GOLD 0 is defined with >= rather than >, and the Methods now record that spirometric strata are taken from the COPDGene staging variable rather than by applying cut-points.",
    "The four COPDGene data file names are removed from Methods and appear in a new Supplement section, COPDGene files used, with what each file provides.",
    "Methods state that agreement is assessed on the binary COPD classification, which is why unweighted kappa is reported.",
    "The Table 1 legend gives the category sizes under the CT-based framework.",
    "Methods state that the two frameworks are fitted separately on the same participants and that the bootstrap is paired.",
    "Figure 1 is regenerated with the new category names and its legend maps the abbreviated panel labels.")))
```

## Not applied, and why

```{r not-applied}
knitr::kable(row.names = FALSE, data.frame(check.names = FALSE,
  `#` = c("1", "3", "7", "8", "16", "17, 22", "19", "24", "25", "26"),
  Status = c(
    "No change requested",
    "OPEN",
    "Deferred",
    "No change needed",
    "Analysis run, wording pending",
    "Analysis run, not yet in the manuscript",
    "Answered, no change made",
    "No action",
    "OPEN",
    "Answered, no change made"),
  Reason = c(
    "Read as an observation on the agreement figure rather than a request. The related point about airway assessment is comment 7.",
    "The suggestion is to call the never-smoking group healthy volunteers rather than controls. The draft still reads controls; the authors have not yet ruled on it.",
    "Stating explicitly that airway wall thickening is also represented by ESI. The authors have deferred this for now.",
    "Reusing the training-derived cut-points on the whole cohort. Supplemental Table 1 already reports the held-out 20% split alongside the full cohort, so the in-sample and independent estimates are both visible.",
    "The adjustment changes one statement in the manuscript beyond the estimates themselves. The authors are agreeing that wording before the result is reported.",
    "The main-effect specification is the one we would report. It has not been added to the manuscript pending a decision on whether to report it as a sensitivity analysis.",
    "The pooling explanation is given above. No manuscript change has been made; a sentence could be added if the authors want the point stated in the paper.",
    "No change required.",
    "That ESI is not produced by standard spirometry software bears on the implementation argument. The v14 Discussion sentence notes that the framework needs only spirometry and symptoms, which is related but does not answer the availability point.",
    "The minor-pathway agreement is close to chance while the two large categories are above 97%. No manuscript change has been made; the authors may prefer to state this directly rather than let the overall agreement figure stand alone.")))
```

# Open with the authors

- **[6]** Renaming COPD-major / COPD-minor to pathway terminology is agreed in
  principle. The exact strings need to be fixed before the change is made, since
  it touches the prose, both main tables, every supplemental table and the
  figures.
- **[7]** Stating explicitly that airway wall thickening is also represented by
  ESI: deferred.
- **[25]** ESI is not produced by standard spirometry software, which bears on
  the implementation argument: left to the authors.
- **Consequence of [16]**, above: if the prior-exacerbation adjustment is
  reported, the claim that AFL-only-noCOPD is not significantly associated with
  any evaluated outcome requires revision.
