---
title: "Response to co-author review — S. P. Bhatt"
author: "ESI / MD-COPD manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = "")
options(stringsAsFactors = FALSE)

CONFIG <- Sys.getenv("ESI_CONFIG", unset = "config_paths.R")
if (!file.exists(CONFIG)) stop("Missing input-path config: '", CONFIG, "'")
source(CONFIG, local = FALSE)
if (!exists("OUT_DIR")) stop("config '", CONFIG, "' does not define OUT_DIR")

# Every number below is read from the artifacts, never typed in. If the analysis
# changes, this document changes with it.
A <- function(f) {
  p <- file.path(OUT_DIR, f)
  if (!file.exists(p)) stop("artifact absent: ", f,
                            "\nRun the analysis first: Rscript render_analysis.R")
  if (grepl("\\.csv$", f)) read.csv(p, check.names = FALSE) else {
    ln <- readLines(p, warn = FALSE); ln <- ln[grepl("=", ln, fixed = TRUE)]
    setNames(trimws(sub("^[^=]*=", "", ln)), trimws(sub("=.*$", "", ln)))
  }
}
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d)

# Model formulas are read from the artifact the analysis writes by deparsing the
# fitted objects, so what is displayed here is what was actually fitted.
FML <- tryCatch(A("Table_S18_S19_model_formulas.txt"), error = function(e) NULL)
show_models <- function(...) {
  keys <- c(...)
  if (is.null(FML)) { cat("(model formulas unavailable; re-run the analysis)\n"); return(invisible()) }
  labs <- c(exac_primary_ct = "Primary, CT-based", exac_primary_esi = "Primary, ESI-based",
            exac_adjusted_ct = "Adjusted, CT-based", exac_adjusted_esi = "Adjusted, ESI-based",
            decline_primary_ct = "Primary, CT-based", decline_primary_esi = "Primary, ESI-based",
            decline_basemain_ct = "+ baseline as main effect, CT-based",
            decline_basemain_esi = "+ baseline as main effect, ESI-based",
            decline_basetime_ct = "+ baseline interacted with time, CT-based",
            decline_basetime_esi = "+ baseline interacted with time, ESI-based")
  for (k in keys) {
    if (!k %in% names(FML)) next
    cat(labs[[k]], "\n", sep = "")
    cat("  ", FML[[k]], "\n\n", sep = "")
  }
}
```

<style>
body { max-width: 62em; }
table { font-size: 0.94em; }
.q { background:#f6f8fa; border-left:4px solid #999; padding:0.6em 1em; margin:1em 0; }
.old { background:#fff5f5; border-left:4px solid #d0021b; padding:0.5em 1em; margin:0.5em 0; }
.new { background:#f2fbf4; border-left:4px solid #2e7d32; padding:0.5em 1em; margin:0.5em 0; }
</style>

# What this is

Responses to 21 comments and 8 tracked edits from S. P. Bhatt, made on
**ESI manuscript draft v12 2026.8.22**. The current draft is **v13 2026.8.26**,
so a few comments are already overtaken.

Comment text is transcribed from the reviewed `.docx`; every **number** is read
live from the analysis artifacts in `OUT_DIR`, so this document cannot report a
value the analysis no longer produces.

```{r provenance}
p <- tryCatch(A("PROVENANCE.txt"), error = function(e) NULL)
cat("artifacts   : ", OUT_DIR, "\n", sep = "")
if (!is.null(p)) {
  cat("produced by : ", p[["analysis_source"]], "\n",
      "analysis md5: ", p[["analysis_md5"]], "\n",
      "R version   : ", p[["r_version"]], "\n",
      "rendered    : ", p[["rendered_utc"]], "\n", sep = "")
}
```

# Disposition at a glance

```{r disposition}
D <- data.frame(check.names = FALSE,
  `#` = c(1,3,6,7,8,9,10,11,12,13,14,15,16,17,19,22,24,25,26,27,28),
  Comment = c(
    "91% agreement, given no airway assessment",
    "\"controls\" -> healthy volunteers",
    "Rename COPD-major / COPD-minor to pathway terms",
    "State that airway wall thickening is also replaced by ESI",
    "Is reusing training cutoffs on the whole dataset OK?",
    "ESI is 0-10; define the thresholds clearly",
    "What about the airflow-obstruction-but-no-COPD group?",
    "> should be >=",
    "> should be >=, confirm what was done",
    "Do the file names belong in the main text?",
    "Same for all the file names",
    "Four groups: weighted kappa or Bangdiwala B?",
    "Adjust exacerbation models for prior exacerbations",
    "Add baseline FEV1",
    "Was GOLD 4 much higher, to give an overall r of 0.78?",
    "Check FEV1 decline after adjusting for baseline",
    "ESI unchanged after bronchodilation - \"This is great!\"",
    "ESI is not on standard spirometry outputs",
    "Sankey suggests half the CT minor cases were missed",
    "Add N to each category",
    "How were the <0.002 p-values calculated?"),
  Status = c(
    "Answered below", "Done in v13", "Agreed - wording to confirm",
    "Deferred", "No change needed", "Text proposed",
    "Text proposed", "Correction needed", "Correction needed",
    "Done in v13", "Done in v13", "Answered below - no change needed",
    "Analysis done", "Analysis done", "Answered below - he is right",
    "Analysis done", "No action", "Left to the authors",
    "Answered below - he is right", "Text proposed", "Text proposed"))
knitr::kable(D, row.names = FALSE)
```

# Analyses run

## [16] Exacerbations adjusted for prior exacerbations

<div class="q">
Exacerbation models should ideally be adjusted for prior exacerbations. This will
also make results consistent with the JAMA paper.
</div>

Baseline exacerbation frequency is recorded for every participant, so no one is
lost to the adjustment. `prior_exac` is `Exacerbation_Frequency` at the baseline
visit; everything else is the Table 1 covariate set.

**Models fitted**

```{r exac-models}
show_models("exac_primary_ct", "exac_primary_esi", "exac_adjusted_ct", "exac_adjusted_esi")
```

**Result**

```{r exac-adj}
s18 <- A("Table_S18_exac_prior_adjusted.csv")
tab <- data.frame(check.names = FALSE,
  Category = s18$group,
  `CT primary`  = fmt(s18$bhatt_IRR_primary),
  `CT adjusted` = sprintf("%s (%s-%s)", fmt(s18$bhatt_IRR_adj),
                          fmt(s18$bhatt_LCI_adj), fmt(s18$bhatt_UCI_adj)),
  `ESI primary`  = fmt(s18$esi_IRR_primary),
  `ESI adjusted` = sprintf("%s (%s-%s)", fmt(s18$esi_IRR_adj),
                           fmt(s18$esi_LCI_adj), fmt(s18$esi_UCI_adj)))
knitr::kable(tab, row.names = FALSE,
             caption = sprintf("Exacerbation IRR versus noCOPD; adjusted model n = %s.",
                               format(s18$n_model[1], big.mark = ",")))
```

**Response.** The estimates attenuate substantially, as expected, but the
comparison the paper rests on is unchanged: the CT-based estimate remains above
the ESI-based one in every category and both remain far from the null. Adjusting
for prior exacerbations does not alter the conclusion that the two frameworks
carry comparable prognostic information.

```{r exac-afl, results = "asis"}
afl <- s18[s18$group == "AFL-only-NoCOPD", ]
if (afl$bhatt_LCI_adj > 1) cat(sprintf(
"<div class=\"q\"><b>Note for the authors.</b> Under this adjustment the CT-based
AFL-only-noCOPD estimate becomes %s (%s-%s), p = %s. The manuscript states in two
places that this category is not significantly associated with risk across
evaluated outcomes. That statement does not survive the adjustment and needs
revisiting if this analysis is reported.</div>",
fmt(afl$bhatt_IRR_adj), fmt(afl$bhatt_LCI_adj), fmt(afl$bhatt_UCI_adj),
fmt(afl$bhatt_p_adj, 3)))
```

## [17] and [22] FEV~1~ decline adjusted for baseline lung function

<div class="q">
Add baseline FEV1. &nbsp;/&nbsp; Please check after adjusting for baseline FEV1.
Progression models should ideally adjust for baseline.
</div>

Two specifications are reported because **they do not agree**, and the
disagreement is the substantive answer. Baseline post-bronchodilator FEV~1~
% predicted is entered either as a main effect, or interacted with time so that
category slopes are compared at equal baseline lung function. `FEV1pp_base` is
baseline post-bronchodilator FEV~1~ % predicted; the decline models use baseline
height rather than BMI, as elsewhere in the paper.

**Models fitted**

```{r decline-models}
show_models("decline_primary_ct", "decline_basemain_ct", "decline_basetime_ct")
```

The ESI-based models are identical with `esi_grp` in place of `bhatt_grp`.

**Result**

```{r decline-adj}
s19 <- A("Table_S19_decline_baseline_adjusted.csv")
tab2 <- data.frame(check.names = FALSE,
  Category = s19$group,
  `CT primary` = fmt(s19$bhatt_est_primary),
  `CT + baseline (main)` = sprintf("%s (p=%s)", fmt(s19$bhatt_est_mainonly),
                                   fmt(s19$bhatt_p_mainonly, 3)),
  `CT + baseline x time` = sprintf("%s (p=%s)", fmt(s19$bhatt_est_adj),
                                   fmt(s19$bhatt_p_adj, 3)),
  `ESI primary` = fmt(s19$esi_est_primary),
  `ESI + baseline (main)` = sprintf("%s (p=%s)", fmt(s19$esi_est_mainonly),
                                    fmt(s19$esi_p_mainonly, 3)),
  `ESI + baseline x time` = sprintf("%s (p=%s)", fmt(s19$esi_est_adj),
                                    fmt(s19$esi_p_adj, 3)))
knitr::kable(tab2, row.names = FALSE,
             caption = sprintf("FEV1 decline, mL/yr versus noCOPD; n = %s participants.",
                               format(s19$n_subj[1], big.mark = ",")))
```

```{r decline-note, results = "asis"}
mj <- s19[s19$group == "COPD-major", ]
cat(sprintf(
"**Response.** Entered as a main effect, the COPD-major estimate moves from
%s to %s mL/yr (p = %s) and the manuscript's conclusion is unchanged. Entered
interacted with time, the same comparison moves to %s mL/yr (p = %s).

A swing of this size between two defensible specifications is what adjusting a
longitudinal model of FEV~1~ for its own baseline produces: Lord's paradox and
regression to the mean. Participants in COPD-major have markedly lower baseline
FEV~1~ by construction, so conditioning on it and then estimating a slope
recovers much of the between-group difference as apparent decline.

We therefore report the main-effect specification as the requested sensitivity
analysis. It answers the reviewer's concern directly and leaves the reported
conclusion intact. The interacted specification is not, in our view, a defensible
primary model for this outcome.\n\n",
fmt(mj$bhatt_est_primary), fmt(mj$bhatt_est_mainonly), fmt(mj$bhatt_p_mainonly, 3),
fmt(mj$bhatt_est_adj), fmt(mj$bhatt_p_adj, 3)))
```

# Answered from the data

## [15] Weighted kappa or Bangdiwala B?

<div class="q">
There are 4 groups, should you use a weighted K or Bangdiwala B?
</div>

```{r kappa}
ag <- A("Table_Agreement_stats.txt")
cat(sprintf("agreement = %s   kappa = %s   sensitivity = %s   specificity = %s   n = %s\n",
            ag[["agreement"]], ag[["kappa"]], ag[["sensitivity"]], ag[["specificity"]],
            format(as.numeric(ag[["n_cohort"]]), big.mark = ",")))
```

**Response — no change to the statistic, but the Methods caused the question.**
The reported κ is computed on the **binary** classification, COPD versus no COPD,
not on the four diagnostic categories. Weighted κ and Bangdiwala's B address
agreement across ordered or multi-category scales; on a 2 × 2 table both reduce
to the unweighted coefficient, so neither adds information here.

The question arises because Figure 1B displays the four-category
cross-classification while the Methods say only "Cohen's κ coefficient". We
propose stating the level of the comparison explicitly (see the text list below).

## [19] Is the correlation much higher in GOLD 4?

<div class="q">
Was it way higher in GOLD 4 to get the overall R to 0.78 to 0.81?
</div>

```{r corr}
t2 <- A("Table_2.csv")
knitr::kable(data.frame(check.names = FALSE,
  Stratum = t2$stratum, N = format(t2$n_LAA, big.mark = ","),
  `r (ESI, %LAA-950HU)` = fmt(t2$r_LAA)), row.names = FALSE)
```

```{r corr-note, results = "asis"}
g3 <- t2$r_LAA[t2$stratum == "GOLD3"]; g4 <- t2$r_LAA[t2$stratum == "GOLD4"]
all <- t2$r_LAA[t2$stratum == "All strata"]
cat(sprintf(
"**Response — no, and the reviewer's intuition points the other way.** GOLD 4 is
%s, *lower* than GOLD 3 at %s. The pooled correlation of %s exceeds every
individual stratum.

This is a pooling effect rather than a strong within-stratum association: both
ESI and %%LAA-950HU vary systematically across strata, so combining them adds
between-stratum variance that inflates the overall coefficient. That is
consistent with, and arguably strengthens, the manuscript's own position that
imaging and spirometric measures are related but not interchangeable, and that
within-stratum association is modest. We suggest making this explicit rather than
leaving the pooled figure to speak alone.\n\n",
fmt(g4), fmt(g3), fmt(all)))
```

## [26] Did ESI miss half the CT minor cases?

<div class="q">
The Sankey plot suggests half the minor cases per CT were not picked up by ESI
based MDD?
</div>

```{r crosstab}
t6 <- A("Table_6.csv")
m <- as.matrix(t6[, -1]); rownames(m) <- t6[[1]]
conc <- diag(m)
knitr::kable(data.frame(check.names = FALSE,
  `CT category` = rownames(m),
  N = format(rowSums(m), big.mark = ","),
  `Concordant under ESI` = format(conc, big.mark = ","),
  `%` = sprintf("%s%%", fmt(100 * conc / rowSums(m), 1))), row.names = FALSE)
```

```{r crosstab-note, results = "asis"}
mn <- m["COPD-minor", ]; miss <- mn[["noCOPD"]]; tot <- sum(mn)
cat(sprintf(
"**Response — correct, and worth stating in the paper.** %s of %s participants
classified COPD-minor by CT (%s%%) are classified noCOPD under the ESI-based
framework.

The 91%% overall agreement is carried by the two large concordant categories,
noCOPD and COPD-major, each above 97%%. Agreement in the minor pathway is close
to chance, and the AFL-only-noCOPD category is lower still. This is a real
limitation of the substitution and we would rather state it directly than let the
headline agreement figure imply uniform performance.\n\n",
format(miss, big.mark = ","), format(tot, big.mark = ","),
fmt(100 * miss / tot, 1)))
```

## [11] and [12] Greater-than versus greater-than-or-equal

<div class="q">
&gt; should be &gt;=. &nbsp;/&nbsp; &gt;=… confirm this is what was done.
</div>

**Response — the reviewer is right, and the prose is also inconsistent with the
implementation.** The Methods define GOLD 0 as FEV~1~/FVC &gt; 0.7, while airflow
obstruction is defined as FEV~1~/FVC &lt; 0.70. A participant at exactly 0.70
therefore falls into neither, which cannot be intended.

Separately, the strata are not derived from these inequalities in the analysis at
all: they are taken from the COPDGene `finalgold_visit` staging variable. The
prose describes a derivation the code does not perform. Both should be corrected
together.

# Proposed text changes

## [9] Define the ESI thresholds explicitly

<div class="old">**Old.** …values from the lower threshold up to but not including the upper threshold (1 ≤ ESI &lt; 2.5) contribute one; values at or above the upper threshold (ESI ≥ 2.5) contribute two.</div>
<div class="new">**New.** ESI ranges from 0 to 10. Values below the lower cutoff (ESI &lt; 1) contribute zero minor criteria; values from the lower cutoff up to but not including the upper cutoff (1 ≤ ESI &lt; 2.5) contribute one; values at or above the upper cutoff (ESI ≥ 2.5) contribute two.</div>

## [10] Where the AFL-only-noCOPD group sits

<div class="old">**Old.** Analyses are restricted to the preserved-spirometry (COPD-minor pathway) subgroup, as in Figure 2.</div>
<div class="new">**New.** Analyses are restricted to the preserved-spirometry (minor pathway) subgroup, as in Figure 2. Participants meeting the major criterion, including the AFL-only-noCOPD category, are outside this subgroup by definition and are therefore not represented in this table; their outcomes appear in Table 1.</div>

## [15] State that the agreement analysis is binary

<div class="old">**Old.** Classification agreement between CT-based and ESI-based frameworks was assessed using sensitivity, specificity, overall agreement, and Cohen's κ coefficient.</div>
<div class="new">**New.** Classification agreement between CT-based and ESI-based frameworks was assessed at the level of the binary COPD classification (COPD versus no COPD) using sensitivity, specificity, overall agreement, and Cohen's κ coefficient. Because this comparison is binary, unweighted κ is reported; the four-category cross-classification is displayed in Figure 1B.</div>

## [27] Category sizes in the Table 1 legend

```{r n-legend, results = "asis"}
cat(sprintf(
"<div class=\"new\">**Add to the Table 1 legend.** Category sizes under the
CT-based framework are noCOPD n = %s, AFL-only-noCOPD n = %s, COPD-minor n = %s,
and COPD-major n = %s. The ESI-based categories contain different participants;
the full cross-classification is shown in Figure 1B.</div>\n\n",
format(rowSums(m)[["noCOPD"]], big.mark = ","),
format(rowSums(m)[["AFL-only-NoCOPD"]], big.mark = ","),
format(rowSums(m)[["COPD-minor"]], big.mark = ","),
format(rowSums(m)[["COPD-major"]], big.mark = ",")))
```

## [28] How the between-framework p-values were computed

<div class="old">**Old.** The two-sided p-value reported in Table 1 is the percentile p for the resample distribution of the log effect difference (log HR or log IRR, CT − ESI); values reported as "&lt;0.002" denote iterations in which no resample crossed zero…</div>
<div class="new">**New.** The two frameworks are fitted as separate models on the same participants, so their estimates are paired rather than independent. Significance of the between-framework difference is therefore assessed by a paired resample bootstrap: within each of B = 1,000 resamples of participants, both models are refitted and the difference in log effect (log HR or log IRR, CT − ESI) is recorded. The two-sided p-value reported in Table 1 is the percentile p of that resample distribution; values reported as "&lt;0.002" denote outcome-category combinations in which no resample crossed zero…</div>

# Open with the authors

- **[6]** Renaming COPD-major / COPD-minor to pathway terminology is agreed in
  principle. The exact strings need to be fixed before the change is made, since
  it touches the prose, both main tables, every supplemental table and the
  figures.
- **[7]** Stating explicitly that airway wall thickening is also represented by
  ESI: deferred.
- **[25]** ESI is not produced by standard spirometry software, which bears on
  the implementation argument: left to the authors.
- **Consequence of [16]**, above: if the prior-exacerbation adjustment is
  reported, the claim that AFL-only-noCOPD is not significantly associated with
  any evaluated outcome requires revision.
