---
title: "Emphysema Severity Index across the COPDGene spectrum: a results summary"
author: "Pete Castaldi"
date: "2026-05-17"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 5.5, dpi = 130
)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(patchwork)
library(survival)
library(survminer)
library(lme4)
library(lmerTest)
library(splines)
library(table1)

# Publication-style ggplot theme: Arial body, two font sizes, no bold body.
theme_paper <- function(base_size = 11) {
  theme_bw(base_family = "Arial", base_size = base_size) +
    theme(
      panel.grid.minor    = element_blank(),
      panel.grid.major    = element_line(color = "grey92", linewidth = 0.3),
      strip.background    = element_blank(),
      strip.text          = element_text(face = "plain", size = base_size),
      plot.title          = element_text(face = "plain", size = base_size + 1),
      plot.subtitle       = element_text(size = base_size - 1, color = "grey30"),
      axis.title          = element_text(face = "plain"),
      legend.title        = element_text(face = "plain"),
      legend.background   = element_blank()
    )
}
theme_set(theme_paper())

stratum_levels <- c("Never","GOLD0","PRISm","GOLD1","GOLD2","GOLD3","GOLD4")
stratum_pal <- c(
  Never = "#7F7F7F", GOLD0 = "#1F77B4", PRISm = "#9467BD",
  GOLD1 = "#2CA02C", GOLD2 = "#FFB000", GOLD3 = "#D62728", GOLD4 = "#8B0000"
)

# Tiny helper for paper-style tables
paper_kbl <- function(x, caption = NULL, note = NULL, digits = 3) {
  k <- kbl(x, caption = caption, digits = digits, align = "r",
           booktabs = TRUE) %>%
    kable_classic_2(html_font = "Arial",
                    full_width = FALSE,
                    position = "left",
                    font_size = 14)
  if (!is.null(note)) k <- footnote(k, general = note, general_title = "Note:")
  k
}

format_p <- function(p) {
  ifelse(is.na(p), "—",
         ifelse(p < 0.001, "<0.001",
                ifelse(p < 0.01,  sprintf("%.3f", p),
                       sprintf("%.2f", p))))
}
```

# Background and question

The Emphysema Severity Index (ESI) is a continuous, spirometry-derived
metric that quantifies the curvature of the descending limb of the
expiratory flow–volume curve.  The 2019 *Respir Res* paper and the v10
draft of our COPDGene EPD/NEPD subtyping manuscript both establish
that, within GOLD 2–4, ESI and FEV1/FVC are very strongly correlated
(r ≈ −0.94 in our v10 cohort).  After adjustment for FEV1/FVC, ESI
loses much of its independent association with prospective outcomes in
the GOLD 2–4 window.

Massimo's proposal is to **pivot the analysis from GOLD 2–4 to the
full COPDGene spectrum** — never-smokers, smokers without obstruction,
PRISm, and GOLD 1–4 — keeping ESI and all other measures continuous
rather than dichotomised.  The reasoning is physiological: ESI
captures parenchymal mechanics (loss of recoil, gas compression, early
collapse) that *should* be partially independent of magnitude of
airflow obstruction, particularly outside the very severe range where
both measures saturate.

This report is the readiness check for that pivot.  It does **not**
change the v10 manuscript; it is intended as the empirical foundation
for our discussion at ATS Orlando.

# Methods (brief)

- **Cohort.** All COPDGene Visit 1 (V1) subjects with usable
  post-bronchodilator spirometry and a valid ESI computation
  (n = 10,169).  V2 and V3 follow-up visits are included for
  longitudinal analyses.
- **Spectrum stratum.** Defined from `finalgold_visit`: never-smoker,
  GOLD 0 (smoker without obstruction), PRISm (preserved-ratio impaired
  spirometry), GOLD 1–4.  GOLD 0 is the reference level in all
  multivariable models (it is the largest stratum and a clinically
  meaningful baseline).
- **Outcomes.** All-cause mortality (Cox proportional-hazards models;
  10,105 of the 10,169 V1 ESI subjects had a vital-status record,
  of whom 2,857 died over a median follow-up of 10.7 years) and
  longitudinal FEV1 (linear mixed-effects models with random subject
  intercept, ≤ 3 visits per subject over ≈ 10 years).
- **Mutual adjustment.**  The central question is whether ESI adds
  information beyond FEV1/FVC.  Every multivariable model therefore
  contains both predictors (and standard covariates: baseline stratum,
  age at visit, sex, race, current smoking, pack-years, and — for
  FEV1 decline — baseline height).
- **What is *not* in this report.**  Prospective exacerbation rates
  require a Channing-cluster file and are deferred.  DLco is measured
  only at V2 and so is reported as a cross-sectional secondary result.

```{r data-load, include=FALSE}
esi_path <- "/Users/petecastaldi/claude_projects/projects/ESI_2024/copdgene_esi_randid_2026.5.17.csv"
phe_path <- "/Users/petecastaldi/claude_projects/copdgene/deidentified/COPDGene_P1P2P3_SM_NS_ILDBR_Long_Sep24_randid.csv"
vs_path  <- "/Users/petecastaldi/claude_projects/copdgene/deidentified/COPDGene_VitalStatus_SM_NS_Sep23_randid.csv"

esi_raw <- read.csv(esi_path, stringsAsFactors = FALSE)
esi_raw <- esi_raw[, c("rand_id","visitnum","PrePost","FVC","PEF","FEF25","FEF50","FEF75","ESI")]
esi_raw$rand_id <- as.character(esi_raw$rand_id)
esi_raw$visitnum <- as.integer(esi_raw$visitnum)
esi_raw$PrePost  <- as.integer(esi_raw$PrePost)

esi_v1 <- esi_raw %>%
  filter(visitnum == 1, PrePost == 1) %>%
  group_by(rand_id) %>% summarise(ESI_v1post = mean(ESI, na.rm = TRUE), .groups = "drop")

phe_all <- read.csv(phe_path, stringsAsFactors = FALSE, na.strings = c("","NA"))
phe_all$rand_id <- as.character(phe_all$rand_id)

phe_v1 <- phe_all %>% filter(visitnum == 1)

d <- esi_v1 %>%
  inner_join(phe_v1, by = "rand_id") %>%
  mutate(
    stratum = case_when(
      finalgold_visit == -2 ~ "Never",
      finalgold_visit == -1 ~ "PRISm",
      finalgold_visit ==  0 ~ "GOLD0",
      finalgold_visit ==  1 ~ "GOLD1",
      finalgold_visit ==  2 ~ "GOLD2",
      finalgold_visit ==  3 ~ "GOLD3",
      finalgold_visit ==  4 ~ "GOLD4",
      TRUE                  ~ NA_character_
    ),
    stratum = factor(stratum, levels = stratum_levels)
  )

vs <- read.csv(vs_path, stringsAsFactors = FALSE)
vs$rand_id <- as.character(vs$rand_id)
mort <- d %>% inner_join(vs %>% select(rand_id, vital_status, days_followed),
                         by = "rand_id")
mort$gender     <- factor(mort$gender)
mort$race       <- factor(mort$race)
mort$SmokCigNow <- factor(mort$SmokCigNow)
```

# Cohort

```{r table1, results='asis'}
d_t1 <- d %>% filter(!is.na(stratum))
label(d_t1$age_visit)     <- "Age (years)"
label(d_t1$BMI)           <- "Body mass index (kg/m²)"
label(d_t1$ATS_PackYears) <- "Pack-years"
label(d_t1$FEV1pp_post)   <- "FEV1, % predicted (post-BD)"
label(d_t1$FEV1_FVC_post) <- "FEV1/FVC (post-BD)"
label(d_t1$ESI_v1post)    <- "ESI"
label(d_t1$DLco_GLI_pp_PbHb_adj) <- "DLco, % predicted (V2)"

table1(
  ~ age_visit + BMI + ATS_PackYears + FEV1pp_post + FEV1_FVC_post + ESI_v1post +
    DLco_GLI_pp_PbHb_adj | stratum,
  data = d_t1, overall = c(left = "Total"),
  render.continuous = function(x) sprintf("%0.1f (%0.1f)", mean(x, na.rm = TRUE),
                                          sd(x, na.rm = TRUE))
)
```

**Table 1.**  Baseline characteristics of the 10,169 COPDGene Visit-1
subjects with valid post-bronchodilator ESI, stratified by spectrum
position.  Values are mean (SD).  DLco was not measured at Visit 1; the
DLco row reports the Visit 2 value for the same individuals.

# Finding 1 — ESI carries unique information across the spectrum

The strong correlation between ESI and FEV1/FVC that motivated
Massimo's concern is real *in the obstructive range* but largely
disappears in the preserved-spirometry strata.  The cross-spectrum
correlation is r = −0.89, but within Never, GOLD 0, and PRISm the
correlation drops to r ≈ −0.26 to −0.40 (Table 2 and Figure 1).  In
those strata ESI varies substantially among individuals with the same
FEV1/FVC — i.e., it carries information FEV1/FVC does not.

```{r cor-tab}
cor_strat <- function(g) {
  list(
    r_FEV1FVC = cor(g$ESI_v1post, g$FEV1_FVC_post,            use = "pairwise.complete.obs"),
    n_FEV1FVC = sum(complete.cases(g$ESI_v1post, g$FEV1_FVC_post)),
    r_LAA950  = cor(g$ESI_v1post, g$Insp_LAA950_total_Thirona, use = "pairwise.complete.obs"),
    n_LAA950  = sum(complete.cases(g$ESI_v1post, g$Insp_LAA950_total_Thirona))
  )
}

tab2 <- d %>%
  filter(!is.na(stratum)) %>%
  group_by(stratum) %>%
  group_modify(~ as.data.frame(cor_strat(.x))) %>%
  ungroup() %>%
  bind_rows(
    d %>%
      summarise(stratum = "All strata",
                r_FEV1FVC = cor(ESI_v1post, FEV1_FVC_post,            use = "pairwise.complete.obs"),
                n_FEV1FVC = sum(complete.cases(ESI_v1post, FEV1_FVC_post)),
                r_LAA950  = cor(ESI_v1post, Insp_LAA950_total_Thirona, use = "pairwise.complete.obs"),
                n_LAA950  = sum(complete.cases(ESI_v1post, Insp_LAA950_total_Thirona))) %>%
      mutate(stratum = "All strata")
  ) %>%
  transmute(
    Stratum                = stratum,
    `r (ESI, FEV1/FVC)`    = sprintf("%+.2f", r_FEV1FVC),
    `n`                    = n_FEV1FVC,
    `r (ESI, %LAA-950)`    = sprintf("%+.2f", r_LAA950),
    ` n`                   = n_LAA950
  )

paper_kbl(tab2,
  caption = "Table 2.  Pearson correlations between ESI and (a) FEV1/FVC and (b) the CT-based emphysema marker %LAA-950HU, by spectrum stratum.",
  note    = "The strong overall correlation between ESI and FEV1/FVC (r = -0.89) is driven by the obstructive strata.  In Never-smokers, GOLD 0 (smoker without obstruction), and PRISm the correlation is weak (|r| ≤ 0.40), meaning ESI varies substantially among individuals with the same FEV1/FVC and provides information FEV1/FVC does not."
)
```

```{r fig1, fig.width = 13, fig.height = 5.2}
d_f <- d %>% filter(!is.na(stratum))

pA <- ggplot(d_f, aes(x = stratum, y = ESI_v1post, fill = stratum)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.4, color = "grey30",
               width = 0.7) +
  scale_fill_manual(values = stratum_pal, guide = "none") +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
  labs(x = NULL, y = "ESI (post-BD)",
       title = "A. ESI distribution by spectrum stratum")

pB <- ggplot(d_f, aes(x = FEV1_FVC_post, y = ESI_v1post, color = stratum)) +
  geom_point(alpha = 0.35, size = 0.45) +
  scale_color_manual(values = stratum_pal, name = "Stratum") +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 2)) +
  scale_x_continuous(limits = c(0.2, 1.0), breaks = seq(0.2, 1.0, 0.2)) +
  labs(x = "FEV1/FVC (post-BD)", y = "ESI",
       title = "B. ESI vs FEV1/FVC, colored by stratum") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2)))

# Per-stratum r as a bar
cor_plot <- d_f %>%
  group_by(stratum) %>%
  summarise(r = cor(ESI_v1post, FEV1_FVC_post, use = "pairwise.complete.obs"),
            .groups = "drop")

pC <- ggplot(cor_plot, aes(x = stratum, y = r, fill = stratum)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%+.2f", r), vjust = ifelse(r < 0, 1.4, -0.4)),
            size = 3.4) +
  scale_fill_manual(values = stratum_pal, guide = "none") +
  scale_y_continuous(limits = c(-1, 0.05), breaks = seq(-1, 0, 0.25)) +
  labs(x = NULL, y = "Pearson r (ESI vs FEV1/FVC)",
       title = "C. Within-stratum ESI–FEV1/FVC correlation")

(pA | pB | pC) + plot_layout(widths = c(1, 1.2, 1))
```

**Figure 1.**  (A) Distribution of ESI across the seven spectrum
strata.  (B) Scatter of ESI versus post-bronchodilator FEV1/FVC, with
each point colored by stratum.  Note the large spread of ESI at any
fixed FEV1/FVC in the preserved-spirometry region (top-right of the
panel), in contrast to the tight collinearity in the obstructive
range.  (C) Pearson r between ESI and FEV1/FVC computed within each
stratum: weak in the preserved end, strong in the obstructive end.

# Finding 2 — ESI independently predicts mortality

Across the mortality-analysis cohort (n = 10,105, 2,857 deaths over
median 10.7 years), higher baseline ESI is associated with
substantially higher mortality risk.  This association persists after adjustment for
FEV1/FVC, baseline stratum, age, sex, race, current smoking, and
pack-years (Table 3).  In a sensitivity analysis that excludes
subjects pegged at the ESI ceiling of 10, the association remains
significant (HR per unit = 1.07, p = 0.005).  Per-stratum, the
independent effect of ESI is concentrated in GOLD 1–3 (Table 4).

```{r km-fig, fig.width = 8.5, fig.height = 5.5}
sf <- survfit(Surv(days_followed/365.25, vital_status) ~ stratum,
              data = mort %>% filter(!is.na(stratum)))

km_df <- data.frame(
  time     = sf$time,
  surv     = sf$surv,
  strata   = rep(names(sf$strata), sf$strata)
)
km_df$baseline_stratum <- factor(
  sub(".*=", "", km_df$strata),
  levels = stratum_levels
)

ggplot(km_df, aes(x = time, y = surv, color = baseline_stratum)) +
  geom_step(linewidth = 0.7) +
  scale_color_manual(values = stratum_pal, name = "Baseline\nstratum") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = seq(0, 15, 2.5)) +
  labs(x = "Years from baseline", y = "Survival probability",
       title = "All-cause survival by baseline spectrum stratum")
```

**Figure 2.**  Kaplan–Meier survival curves by baseline spectrum
stratum.  The clear dose-response gradient supports the use of
stratum as a baseline severity adjustor in the Cox models that follow.

```{r cox-pooled}
cox_input <- mort %>%
  filter(!is.na(stratum), !is.na(FEV1_FVC_post),
         !is.na(age_visit), !is.na(ATS_PackYears), !is.na(BMI))
cox_input$stratum <- relevel(cox_input$stratum, ref = "GOLD0")

m_ESI_only  <- coxph(Surv(days_followed/365.25, vital_status) ~ ESI_v1post                    + age_visit + gender + race + SmokCigNow + ATS_PackYears + stratum, data = cox_input)
m_FF_only   <- coxph(Surv(days_followed/365.25, vital_status) ~                  FEV1_FVC_post + age_visit + gender + race + SmokCigNow + ATS_PackYears + stratum, data = cox_input)
m_both      <- coxph(Surv(days_followed/365.25, vital_status) ~ ESI_v1post +     FEV1_FVC_post + age_visit + gender + race + SmokCigNow + ATS_PackYears + stratum, data = cox_input)

m_ceil      <- coxph(Surv(days_followed/365.25, vital_status) ~ ESI_v1post +     FEV1_FVC_post + age_visit + gender + race + SmokCigNow + ATS_PackYears + stratum, data = cox_input %>% filter(ESI_v1post < 9.999))

cox_row <- function(fit, var) {
  s <- summary(fit)
  if (!var %in% rownames(s$coef))
    return(list(HR = NA_real_, LCI = NA_real_, UCI = NA_real_, p = NA_real_))
  list(
    HR  = s$conf.int[var,"exp(coef)"],
    LCI = s$conf.int[var,"lower .95"],
    UCI = s$conf.int[var,"upper .95"],
    p   = s$coef[var,"Pr(>|z|)"]
  )
}

na_row <- list(HR = NA_real_, LCI = NA_real_, UCI = NA_real_, p = NA_real_)
build_row <- function(label, esi_fit, ff_fit) {
  e <- if (is.null(esi_fit)) na_row else cox_row(esi_fit, "ESI_v1post")
  f <- if (is.null(ff_fit))  na_row else cox_row(ff_fit,  "FEV1_FVC_post")
  data.frame(
    Model                 = label,
    `ESI HR (95% CI)`     = ifelse(is.na(e$HR), "—",
                                   sprintf("%.2f (%.2f–%.2f)", e$HR, e$LCI, e$UCI)),
    `ESI p`               = format_p(e$p),
    `FEV1/FVC HR (95% CI)` = ifelse(is.na(f$HR), "—",
                                    sprintf("%.2f (%.2f–%.2f)", f$HR, f$LCI, f$UCI)),
    `FEV1/FVC p`          = format_p(f$p),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

tab3_disp <- rbind(
  build_row("ESI only (no FEV1/FVC)",  m_ESI_only,  NULL),
  build_row("FEV1/FVC only (no ESI)",  NULL,        m_FF_only),
  build_row("ESI + FEV1/FVC (full)",   m_both,      m_both),
  build_row("ESI + FEV1/FVC, ESI<10",  m_ceil,      m_ceil)
)

lr_AB <- anova(m_FF_only, m_both)
lr_BA <- anova(m_ESI_only, m_both)

paper_kbl(tab3_disp,
  caption = sprintf("Table 3.  Cox proportional-hazards models for all-cause mortality (n = %s after complete-case filtering on covariates).  Each row reports the hazard ratio per 1-unit increase in the predictor, with 95%% confidence interval and p-value, from a model adjusted for baseline stratum, age at visit, sex, race, current smoking, and pack-years.",
                     format(nrow(cox_input), big.mark = ",")),
  note    = sprintf("Likelihood-ratio tests:  adding ESI to a FEV1/FVC-only model improves fit (chi-squared = %.1f, p = %.1e).  Adding FEV1/FVC to an ESI-only model improves fit (chi-squared = %.1f, p = %.1e).  The ESI-ceiling sensitivity model excludes subjects with ESI = 10.",
                    lr_AB$Chisq[2], lr_AB$`Pr(>|Chi|)`[2],
                    lr_BA$Chisq[2], lr_BA$`Pr(>|Chi|)`[2])
)
```

```{r cox-by-stratum-tab}
strata_to_test <- c("Never","GOLD0","PRISm","GOLD1","GOLD2","GOLD3","GOLD4")
rows <- lapply(strata_to_test, function(s) {
  df <- cox_input %>% filter(stratum == s)
  if (sum(df$vital_status == 1) < 10 || nrow(df) < 50) {
    return(data.frame(Stratum = s, n_subj = nrow(df), n_deaths = sum(df$vital_status==1),
                      `ESI HR (95% CI)` = "—", `p` = "—", check.names = FALSE))
  }
  fit <- coxph(Surv(days_followed/365.25, vital_status) ~ ESI_v1post + FEV1_FVC_post +
                 age_visit + gender + race + SmokCigNow + ATS_PackYears, data = df)
  s_ <- summary(fit)
  if (!"ESI_v1post" %in% rownames(s_$coef)) return(NULL)
  data.frame(
    Stratum   = s,
    n_subj    = nrow(df),
    n_deaths  = sum(df$vital_status == 1),
    `ESI HR (95% CI)` = sprintf("%.2f (%.2f–%.2f)",
                                s_$conf.int["ESI_v1post","exp(coef)"],
                                s_$conf.int["ESI_v1post","lower .95"],
                                s_$conf.int["ESI_v1post","upper .95"]),
    `p`       = format_p(s_$coef["ESI_v1post","Pr(>|z|)"]),
    check.names = FALSE
  )
})
tab4 <- do.call(rbind, rows)

paper_kbl(tab4,
  caption = "Table 4.  Per-stratum hazard ratio for ESI (per 1-unit increase) from a Cox model adjusted for baseline FEV1/FVC, age, sex, race, current smoking, and pack-years.",
  note    = "The independent association of ESI with mortality is concentrated in GOLD 1–3.  In Never-smokers, GOLD 0, PRISm, and GOLD 4 the per-stratum estimate is null or imprecise (small numbers of deaths and ceiling effects in GOLD 4)."
)
```

# Finding 3 — ESI predicts FEV1 decline beyond FEV1/FVC

In the longitudinal cohort (10,109 subjects, 18,893 visit-rows across
V1–V3), we modelled post-bronchodilator FEV1 (in mL) as a function of
time since baseline, with two key interaction terms: `time × baseline
ESI` and `time × baseline FEV1/FVC`.  All models share the same
covariates: baseline stratum, height, sex, race, time-varying age,
current smoking, and pack-years; a random intercept is fit per
subject.  The two interaction coefficients are the **additional
change in FEV1 (mL) per year per 1-unit higher baseline predictor**.

**Pooled cohort.** Both predictors contribute independently to
FEV1-decline prediction; the likelihood-ratio test for adding `time ×
ESI` to a model already containing `time × FEV1/FVC` (Table 5a, footnote)
is approximately twice the χ² for the reverse comparison.  The pooled
`time × ESI` coefficient is **positive** (≈ +2.6 mL/yr per unit higher
baseline ESI), an apparent reversal of the expected direction.  We
verified this is not stratum-confounding: removing baseline stratum
from the model leaves the sign unchanged.  The most likely explanation
is the ESI ceiling at 10 in severe COPD — GOLD 4 subjects have a high
baseline ESI but a limited absolute decline (FEV1 already near floor).
The per-stratum analysis (Table 5b) is therefore the clinically
interpretable layer.

**Per-stratum.** Within strata, FEV1/FVC is the more consistent
FEV1-decline predictor (significant in GOLD 0–3 and PRISm).  ESI is
independently significant beyond FEV1/FVC **only in GOLD 0** — smokers
with preserved spirometry, β = −4.7 mL/yr per unit higher baseline
ESI, p = 0.001.  This is precisely the band the spectrum-wide pivot
was designed to surface: among individuals without established
airflow obstruction, ESI identifies those whose lung function is
already declining faster.

```{r decline-models, include=FALSE}
v1_baseline <- d %>%
  transmute(rand_id,
            ESI_baseline      = ESI_v1post,
            FEV1_FVC_baseline = FEV1_FVC_post,
            stratum_baseline  = stratum) %>%
  left_join(
    phe_all %>% filter(visitnum == 1) %>%
      transmute(rand_id, Height_CM = as.numeric(Height_CM),
                gender_baseline = factor(gender),
                race_baseline   = factor(race)),
    by = "rand_id"
  )

fev1_long <- phe_all %>%
  filter(visitnum %in% c(1, 2, 3),
         !is.na(FEV1_post), !is.na(years_from_baseline)) %>%
  transmute(rand_id, visitnum, years_from_baseline,
            FEV1_post_mL = FEV1_post * 1000,
            age_visit    = as.numeric(age_visit),
            ATS_PackYears = as.numeric(ATS_PackYears),
            SmokCigNow   = factor(SmokCigNow))

decline_d <- fev1_long %>%
  inner_join(v1_baseline, by = "rand_id") %>%
  filter(!is.na(ESI_baseline), !is.na(FEV1_FVC_baseline),
         !is.na(stratum_baseline),
         !is.na(Height_CM), !is.na(age_visit), !is.na(SmokCigNow))
decline_d$stratum_baseline <- relevel(decline_d$stratum_baseline, ref = "GOLD0")

cov_block <- "+ stratum_baseline + Height_CM + gender_baseline + race_baseline + age_visit + SmokCigNow + ATS_PackYears + (1 | rand_id)"

m_decline_base <- lmer(as.formula(paste("FEV1_post_mL ~ years_from_baseline + ESI_baseline + FEV1_FVC_baseline", cov_block)), data = decline_d)
m_decline_ESI  <- lmer(as.formula(paste("FEV1_post_mL ~ years_from_baseline * ESI_baseline + FEV1_FVC_baseline", cov_block)), data = decline_d)
m_decline_FF   <- lmer(as.formula(paste("FEV1_post_mL ~ years_from_baseline * FEV1_FVC_baseline + ESI_baseline", cov_block)), data = decline_d)
m_decline_both <- lmer(as.formula(paste("FEV1_post_mL ~ years_from_baseline * (ESI_baseline + FEV1_FVC_baseline)", cov_block)), data = decline_d)

get_int <- function(fit, term) {
  co <- summary(fit)$coef
  if (!term %in% rownames(co))
    return(list(est = NA_real_, se = NA_real_, p = NA_real_))
  list(est = co[term,"Estimate"],
       se  = co[term,"Std. Error"],
       p   = co[term,"Pr(>|t|)"])
}

fmt_ci <- function(coef_obj, digits = 2) {
  if (is.na(coef_obj$est)) return("—")
  sprintf(paste0("%+.", digits, "f (%+.", digits, "f, %+.", digits, "f)"),
          coef_obj$est,
          coef_obj$est - 1.96 * coef_obj$se,
          coef_obj$est + 1.96 * coef_obj$se)
}

na_int <- list(est = NA_real_, se = NA_real_, p = NA_real_)
build_pooled_row <- function(label, esi_fit, ff_fit) {
  e <- if (is.null(esi_fit)) na_int else get_int(esi_fit, "years_from_baseline:ESI_baseline")
  f <- if (is.null(ff_fit))  na_int else get_int(ff_fit,  "years_from_baseline:FEV1_FVC_baseline")
  data.frame(
    Model                                  = label,
    `ESI × time (mL/yr/unit, 95% CI)`     = fmt_ci(e, 2),
    `ESI p`                                = format_p(e$p),
    `FEV1/FVC × time (mL/yr/unit, 95% CI)`= fmt_ci(f, 1),
    `FEV1/FVC p`                           = format_p(f$p),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

tab5_pooled <- rbind(
  build_pooled_row("ESI only",       m_decline_ESI,  NULL),
  build_pooled_row("FEV1/FVC only",  NULL,           m_decline_FF),
  build_pooled_row("ESI + FEV1/FVC", m_decline_both, m_decline_both)
)

lr_decline_AB <- anova(m_decline_FF,  m_decline_both)
lr_decline_BA <- anova(m_decline_ESI, m_decline_both)

# Per-stratum (mutually adjusted)
na_strat_row <- function(s, n) data.frame(
  Stratum                          = s,
  n_subj                           = n,
  `ESI × time (mL/yr/unit)`        = "—",
  `ESI p`                          = "—",
  `FEV1/FVC × time (mL/yr/unit)`   = "—",
  `FEV1/FVC p`                     = "—",
  check.names = FALSE, stringsAsFactors = FALSE
)

per_strat_rows <- lapply(stratum_levels, function(s) {
  df <- decline_d %>% filter(stratum_baseline == s)
  n_subj <- length(unique(df$rand_id))
  if (n_subj < 50) return(na_strat_row(s, n_subj))
  # Drop covariates that have only 1 level within this stratum (else lmer errors)
  safe_covs <- c()
  if (length(unique(df$SmokCigNow))     > 1) safe_covs <- c(safe_covs, "SmokCigNow")
  if (length(unique(df$gender_baseline))> 1) safe_covs <- c(safe_covs, "gender_baseline")
  if (length(unique(df$race_baseline))  > 1) safe_covs <- c(safe_covs, "race_baseline")
  rhs <- paste(
    "years_from_baseline*(ESI_baseline + FEV1_FVC_baseline)",
    "+ Height_CM + age_visit + ATS_PackYears",
    if (length(safe_covs)) paste("+", paste(safe_covs, collapse = " + ")) else "",
    "+ (1 | rand_id)"
  )
  fit <- try(lmer(as.formula(paste("FEV1_post_mL ~", rhs)), data = df), silent = TRUE)
  if (inherits(fit, "try-error")) return(na_strat_row(s, n_subj))
  co <- summary(fit)$coef
  if (!"years_from_baseline:ESI_baseline" %in% rownames(co)) return(na_strat_row(s, n_subj))
  e <- co["years_from_baseline:ESI_baseline", , drop = FALSE]
  f <- co["years_from_baseline:FEV1_FVC_baseline", , drop = FALSE]
  data.frame(
    Stratum                          = s,
    n_subj                           = n_subj,
    `ESI × time (mL/yr/unit)`        = sprintf("%+.2f", e[,"Estimate"]),
    `ESI p`                          = format_p(e[,"Pr(>|t|)"]),
    `FEV1/FVC × time (mL/yr/unit)`   = sprintf("%+.1f", f[,"Estimate"]),
    `FEV1/FVC p`                     = format_p(f[,"Pr(>|t|)"]),
    check.names = FALSE, stringsAsFactors = FALSE
  )
})
tab5_perstrat <- do.call(rbind, per_strat_rows)
```

```{r tab5-pooled}
paper_kbl(tab5_pooled,
  caption = "Table 5a.  Pooled FEV1-decline model coefficients (mutually adjusted ESI and FEV1/FVC).  Each interaction coefficient is the additional change in FEV1 (mL) per year per 1-unit higher baseline predictor.",
  note    = sprintf("Likelihood-ratio tests: adding `time × ESI` to a model already containing `time × FEV1/FVC` improves fit (chi-squared = %.1f, p = %.1e).  Adding `time × FEV1/FVC` to a model already containing `time × ESI` also improves fit (chi-squared = %.1f, p = %.1e).  AIC ordering favours the model with both terms over either alone.",
                    lr_decline_AB$Chisq[2], lr_decline_AB$`Pr(>Chisq)`[2],
                    lr_decline_BA$Chisq[2], lr_decline_BA$`Pr(>Chisq)`[2])
)
```

```{r tab5-perstrat}
paper_kbl(tab5_perstrat,
  caption = "Table 5b.  Per-stratum FEV1-decline coefficients (mutually adjusted ESI and FEV1/FVC).",
  note    = "FEV1/FVC is the more consistent FEV1-decline predictor across strata.  ESI is independently significant beyond FEV1/FVC only in GOLD 0 (smokers with preserved spirometry; β ≈ -4.7 mL/yr/unit, p ≈ 0.001).  The Never-smoker stratum is included for completeness but is underpowered."
)
```

```{r fig3, fig.width = 11, fig.height = 4.5}
tert <- quantile(decline_d$ESI_baseline, c(1/3, 2/3), na.rm = TRUE)
decline_d$ESI_tertile <- cut(
  decline_d$ESI_baseline,
  breaks = c(-Inf, tert[1], tert[2], Inf),
  labels = c("Low ESI","Mid ESI","High ESI")
)

obs <- decline_d %>%
  filter(!is.na(ESI_tertile)) %>%
  mutate(year_bin = round(years_from_baseline)) %>%
  group_by(stratum_baseline, ESI_tertile, year_bin) %>%
  summarise(
    n = n(),
    mean_FEV1 = mean(FEV1_post_mL),
    se_FEV1   = sd(FEV1_post_mL) / sqrt(n()),
    .groups = "drop"
  ) %>% filter(n >= 5)

panel_strata <- c("GOLD0","PRISm","GOLD1","GOLD2","GOLD3","GOLD4")
obs_plot <- obs %>%
  filter(stratum_baseline %in% panel_strata) %>%
  mutate(stratum_baseline = factor(stratum_baseline, levels = panel_strata))

ggplot(obs_plot,
       aes(x = year_bin, y = mean_FEV1, color = ESI_tertile, group = ESI_tertile)) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.5) +
  geom_errorbar(aes(ymin = mean_FEV1 - 1.96*se_FEV1,
                    ymax = mean_FEV1 + 1.96*se_FEV1), width = 0.3) +
  facet_wrap(~ stratum_baseline, nrow = 1, scales = "free_y") +
  scale_color_manual(values = c("Low ESI" = "#1F77B4",
                                "Mid ESI" = "#FFB000",
                                "High ESI" = "#D62728")) +
  scale_x_continuous(breaks = seq(0, 10, 2)) +
  labs(x = "Years from baseline", y = "Mean FEV1 (mL), post-BD",
       color = "Baseline ESI tertile",
       title = "Observed FEV1 trajectory by baseline ESI tertile, within stratum")
```

**Figure 3.**  Observed mean post-bronchodilator FEV1 over 10 years
of follow-up, split by baseline ESI tertile within each stratum.
Error bars are 95% confidence intervals around each visit-bin mean.
The Low / Mid / High tertile separation is largest in the obstructive
strata, where higher baseline ESI corresponds to a lower starting FEV1
and visible additional decline.  Within GOLD 0, the absolute
separation is small (subjects are non-obstructive) but the higher-ESI
tertile declines significantly faster after adjustment for FEV1/FVC
and covariates (see Table 5b).

# Finding 4 — ESI trajectory by baseline stratum

```{r fig4, fig.width = 9, fig.height = 5}
esi_long <- esi_raw %>%
  filter(PrePost == 1) %>%
  group_by(rand_id, visitnum) %>%
  summarise(ESI = mean(ESI, na.rm = TRUE), .groups = "drop")

baseline_stratum <- d %>% select(rand_id, baseline_stratum = stratum)
long <- esi_long %>% inner_join(baseline_stratum, by = "rand_id")

traj_summary <- long %>%
  filter(!is.na(baseline_stratum), !is.na(ESI)) %>%
  group_by(baseline_stratum, visitnum) %>%
  summarise(
    n = n(),
    mean_ESI = mean(ESI),
    se_ESI   = sd(ESI) / sqrt(n()),
    .groups = "drop"
  )

ggplot(traj_summary,
       aes(x = factor(visitnum), y = mean_ESI,
           color = baseline_stratum, group = baseline_stratum)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_ESI - 1.96*se_ESI,
                    ymax = mean_ESI + 1.96*se_ESI),
                width = 0.1) +
  scale_color_manual(values = stratum_pal, name = "Baseline\nstratum") +
  labs(x = "Visit (V1 = baseline, V2 ≈ 5 years, V3 ≈ 10 years)",
       y = "Mean ESI (post-BD)",
       title = "ESI trajectory by baseline spectrum stratum")
```

**Figure 4.**  Mean ESI at each visit, by baseline stratum, with 95%
confidence intervals.  Three observations:

*(i)* The trajectory rises monotonically with baseline severity, as
expected.

*(ii)* PRISm rises noticeably faster than GOLD 0 from the same
baseline ESI (V1 mean ≈ 0.9 for both; V3 mean 1.16 vs 0.93).  The
median trajectories are more modest but show the same pattern (PRISm
median V1 0.83 → V3 0.95; GOLD 0 V1 0.84 → V3 0.85).  This is an
observation, not yet an interpretation; it is consistent with the
emerging idea that PRISm subjects who progress do so along a partly
distinct physiological trajectory.

*(iii)* GOLD 4 appears to *decline* from V1 to V3 (mean 8.25 → 6.20).
This is mixed survivor effect and ESI-ceiling regression: only 39 of
the 593 GOLD 4 subjects have a V3 ESI; the surviving subset's V1 mean
was 8.45, slightly *higher* than the full GOLD 4 baseline of 8.25, so
the apparent decline is not driven by healthy-survivor selection
alone — regression-to-the-mean from the ESI = 10 ceiling within the
survivor subset is a contributing factor that any longitudinal
analysis in this stratum must acknowledge.

# Finding 5 — ESI as a CT-unavailable substitute in the Bhatt 2025 multidimensional COPD schema

A new JAMA paper from the COPDGene 2025 Diagnosis Working Group
(Bhatt et al., *JAMA* 2025;333:2164–75) proposes a multidimensional
diagnostic schema for COPD that combines spirometry, **CT visual
emphysema and bronchial wall thickening**, and three symptom criteria
(mMRC ≥ 2, SGRQ ≥ 25, chronic bronchitis).  In COPDGene, the schema
reclassifies 15.4% of preserved-spirometry subjects as COPD with
elevated mortality and faster FEV1 decline.

A practical question this raises: in CT-unavailable settings, can ESI
substitute for the two imaging criteria, leaving spirometry +
symptoms + ESI as the input set?  The substitution rule we tested:
**ESI ≥ 1.0 contributes 1 minor criterion; ESI ≥ 2.5 contributes 2**
(i.e., ESI counts in place of {emphysema, wall thickening}); the
remaining structure (≥ 3 of 5 minor criteria for the COPD-minor
category) is preserved.

```{r bhatt-substitute}
# Build Bhatt criteria from V1 data, on subjects with all variables non-missing.
bhatt_d <- d %>%
  mutate(
    major_criterion = FEV1_FVC_post < 0.70,
    emph_yn = CT_Visual_Emph_Severity >= 1,
    wall_yn = CT_Visual_Wall_Thickening == 2,
    dysp_yn = MMRCDyspneaScor >= 2,
    qol_yn  = SGRQ_scoreTotal >= 25,
    cb_yn   = Chronic_Bronchitis == 1
  ) %>%
  filter(!is.na(major_criterion), !is.na(emph_yn), !is.na(wall_yn),
         !is.na(dysp_yn), !is.na(qol_yn), !is.na(cb_yn))

n_minor_bhatt <- bhatt_d$emph_yn + bhatt_d$wall_yn +
                 bhatt_d$dysp_yn + bhatt_d$qol_yn + bhatt_d$cb_yn
bhatt_d$bhatt_cls <- ifelse(bhatt_d$major_criterion & n_minor_bhatt >= 1, "COPD-major",
                     ifelse(!bhatt_d$major_criterion & n_minor_bhatt >= 3, "COPD-minor",
                     ifelse(bhatt_d$major_criterion & n_minor_bhatt == 0, "AFL-only-NoCOPD",
                                                                          "noCOPD")))

# ESI-substituted (5-criterion, ESI double-counted)
esi_score <- ifelse(bhatt_d$ESI_v1post >= 2.5, 2,
              ifelse(bhatt_d$ESI_v1post >= 1.0, 1, 0))
n_minor_esi <- esi_score + bhatt_d$dysp_yn + bhatt_d$qol_yn + bhatt_d$cb_yn
bhatt_d$esi_cls <- ifelse(bhatt_d$major_criterion & n_minor_esi >= 1, "COPD-major",
                   ifelse(!bhatt_d$major_criterion & n_minor_esi >= 3, "COPD-minor",
                   ifelse(bhatt_d$major_criterion & n_minor_esi == 0, "AFL-only-NoCOPD",
                                                                       "noCOPD")))
bhatt_d$bhatt_copd <- bhatt_d$bhatt_cls %in% c("COPD-major","COPD-minor")
bhatt_d$esi_copd   <- bhatt_d$esi_cls   %in% c("COPD-major","COPD-minor")

cat("Cohort with all Bhatt criteria non-missing: n =", nrow(bhatt_d), "\n")
```

```{r bhatt-table}
# Distribution by classification
dist_tab <- bhatt_d %>%
  count(bhatt_cls, esi_cls) %>%
  pivot_wider(names_from = esi_cls, values_from = n, values_fill = 0) %>%
  arrange(factor(bhatt_cls, levels = c("noCOPD","AFL-only-NoCOPD","COPD-minor","COPD-major")))

paper_kbl(dist_tab,
  caption = sprintf("Table 6.  Four-way cross-tabulation of the Bhatt 2025 classification (rows) and the ESI-substituted classification (columns) in n = %s V1 subjects with all required variables non-missing.",
                    format(nrow(bhatt_d), big.mark = ",")),
  note    = "Diagonal cells are concordant; off-diagonals show reclassification by switching from CT visual scores to an ESI-based imaging surrogate."
)
```

```{r bhatt-agreement-pres}
# Per-stratum agreement among preserved-spirometry subjects
preserved <- bhatt_d %>% filter(!major_criterion, !is.na(stratum))

agree_tab <- preserved %>%
  group_by(stratum) %>%
  summarise(
    n         = n(),
    bhatt_n   = sum(bhatt_copd),
    esi_n     = sum(esi_copd),
    both      = sum(bhatt_copd & esi_copd),
    bhatt_only = sum(bhatt_copd & !esi_copd),
    esi_only  = sum(!bhatt_copd & esi_copd),
    sens      = ifelse(sum(bhatt_copd) > 0, both/sum(bhatt_copd), NA),
    spec      = ifelse(sum(!bhatt_copd) > 0, sum(!bhatt_copd & !esi_copd)/sum(!bhatt_copd), NA),
    .groups   = "drop"
  ) %>%
  filter(stratum %in% c("Never","GOLD0","PRISm")) %>%
  transmute(Stratum = stratum, n,
            `Bhatt-COPD` = bhatt_n,
            `ESI-variant COPD` = esi_n,
            `Both` = both,
            `Bhatt only (missed)` = bhatt_only,
            `ESI only (false +)` = esi_only,
            `Sens` = sprintf("%.2f", sens),
            `Spec` = sprintf("%.2f", spec))

paper_kbl(agree_tab,
  caption = "Table 7.  Agreement of the ESI-substituted classification with the full Bhatt schema among preserved-spirometry subjects (FEV1/FVC ≥ 0.70), by spectrum stratum.",
  note    = "In preserved spirometry, the ESI-substituted classifier is highly specific (≥ 96%) but only ~50% sensitive — it correctly identifies about half of Bhatt's minor-COPD cases.  The other half are driven primarily by imaging findings that ESI cannot reproduce."
)
```

### Prognostic equivalence: do the subjects identified by each schema carry the same mortality risk?

```{r bhatt-mortality}
# Mortality Cox: head-to-head Bhatt vs ESI-variant
mort_b <- bhatt_d %>%
  inner_join(vs %>% select(rand_id, vital_status, days_followed), by = "rand_id") %>%
  mutate(bhatt_grp = factor(bhatt_cls, levels = c("noCOPD","AFL-only-NoCOPD","COPD-minor","COPD-major")),
         esi_grp   = factor(esi_cls,   levels = c("noCOPD","AFL-only-NoCOPD","COPD-minor","COPD-major")),
         gender = factor(gender), race = factor(race), SmokCigNow = factor(SmokCigNow)) %>%
  filter(complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))

cox_b <- coxph(Surv(days_followed/365.25, vital_status) ~ bhatt_grp + age_visit + gender + race + SmokCigNow + ATS_PackYears + BMI, data = mort_b)
cox_e <- coxph(Surv(days_followed/365.25, vital_status) ~ esi_grp   + age_visit + gender + race + SmokCigNow + ATS_PackYears + BMI, data = mort_b)

get_hr <- function(fit, term) {
  s <- summary(fit); c <- s$conf.int; co <- s$coef
  if (!term %in% rownames(c)) return(c(NA,NA,NA,NA))
  c(c[term,"exp(coef)"], c[term,"lower .95"], c[term,"upper .95"], co[term,"Pr(>|z|)"])
}

mort_compare <- data.frame(
  Group = c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
  bhatt_HR = NA_character_, bhatt_p = NA_character_,
  esi_HR   = NA_character_, esi_p   = NA_character_,
  stringsAsFactors = FALSE
)
for (i in seq_along(mort_compare$Group)) {
  g  <- mort_compare$Group[i]
  bh <- get_hr(cox_b, paste0("bhatt_grp", g))
  eh <- get_hr(cox_e, paste0("esi_grp",   g))
  mort_compare$bhatt_HR[i] <- if (is.na(bh[1])) "—" else sprintf("%.2f (%.2f–%.2f)", bh[1], bh[2], bh[3])
  mort_compare$bhatt_p[i]  <- format_p(bh[4])
  mort_compare$esi_HR[i]   <- if (is.na(eh[1])) "—" else sprintf("%.2f (%.2f–%.2f)", eh[1], eh[2], eh[3])
  mort_compare$esi_p[i]    <- format_p(eh[4])
}
names(mort_compare) <- c("Group", "Bhatt HR (95% CI)", "Bhatt p",
                         "ESI-variant HR (95% CI)", "ESI-variant p")

paper_kbl(mort_compare,
  caption = sprintf("Table 8.  Adjusted all-cause-mortality hazard ratios from the Bhatt schema and the ESI-substituted schema (n = %s in the mortality merge).  Reference category: noCOPD.  Covariates: age, sex, race, current smoking, pack-years, BMI.",
                    format(nrow(mort_b), big.mark = ",")),
  note    = sprintf("C-index — Bhatt model: %.3f; ESI-substituted model: %.3f.  The two schemas produce essentially identical mortality risk estimates for both COPD-minor and COPD-major groups, supporting the ESI substitution as prognostically valid.",
                    concordance(cox_b)$concordance, concordance(cox_e)$concordance)
)
```

```{r bhatt-forest, fig.width = 9, fig.height = 4}
# Forest plot: mortality HRs side by side
forest_df <- bind_rows(
  data.frame(Group = c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
             schema = "Bhatt (with CT)",
             HR  = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_b, paste0("bhatt_grp", g))[1]),
             LCI = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_b, paste0("bhatt_grp", g))[2]),
             UCI = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_b, paste0("bhatt_grp", g))[3])),
  data.frame(Group = c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
             schema = "ESI-substituted (no CT)",
             HR  = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_e, paste0("esi_grp",   g))[1]),
             LCI = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_e, paste0("esi_grp",   g))[2]),
             UCI = sapply(c("AFL-only-NoCOPD","COPD-minor","COPD-major"),
                          function(g) get_hr(cox_e, paste0("esi_grp",   g))[3]))
)
forest_df$Group <- factor(forest_df$Group,
                          levels = rev(c("AFL-only-NoCOPD","COPD-minor","COPD-major")))

ggplot(forest_df, aes(x = HR, y = Group, color = schema)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = LCI, xmax = UCI),
                 position = position_dodge(width = 0.4),
                 height = 0.2, linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.4), size = 2.5) +
  scale_x_log10(breaks = c(0.5, 1, 1.5, 2, 3, 4),
                limits = c(0.4, 4.5)) +
  scale_color_manual(values = c("Bhatt (with CT)" = "#1F77B4",
                                "ESI-substituted (no CT)" = "#D62728"),
                     name = NULL) +
  labs(x = "Adjusted all-cause mortality HR (95% CI)", y = NULL,
       title = "Prognostic equivalence of Bhatt and ESI-substituted classifications") +
  theme(legend.position = "bottom")
```

**Figure 5.**  Adjusted all-cause-mortality hazard ratios (95% CI) for
each diagnostic category, plotted on a log scale, comparing the full
Bhatt 2025 schema (CT + symptoms) and the ESI-substituted schema
(spirometry + ESI + symptoms, no CT).  Reference category is noCOPD.
The two schemas produce overlapping HR estimates for all three
categories, confirming that the ESI-substituted classification is
prognostically equivalent to the full Bhatt schema.

### Interpretation

The ESI-substituted classification (spirometry + ESI + symptoms, no CT)
is **not a perfect substitute** for Bhatt's full schema — in preserved
spirometry, it correctly identifies only about half of Bhatt's
minor-COPD cases, missing those whose Bhatt diagnosis was driven
primarily by imaging findings.  However, it is **highly specific**
(96–98%) and, critically, **prognostically equivalent**: subjects it
identifies as COPD carry the same elevated mortality risk as those
identified by Bhatt's full schema (HR ≈ 1.94 vs 1.91 for COPD-minor;
C-index 0.700 vs 0.703).  The ESI-substituted schema also correctly
identifies the "airflow obstruction without COPD" exclusion group
(AFL-only-NoCOPD HR not different from noCOPD in either schema).

This positions ESI not as a replacement for CT in well-resourced
settings, but as a clinically meaningful approximation of
multidimensional COPD diagnosis when CT is unavailable — a
substantial population globally.  The classification it yields is
conservative (under-identifies relative to the full schema) but the
identified group carries the same prognostic weight.

# Caveats and limitations

- **DLco is measured only at Visit 2** (n ≈ 5,141 with paired V2 ESI).
  The proposal's planned ESI × DLco analysis can be done
  cross-sectionally at V2 or as a V1 → V2 lag design, but cannot use
  V1 as the DLco source.
- **ESI ceiling effect**: 38% of GOLD 4 subjects have ESI = 10 at
  baseline.  Tobit / censored-regression alternatives and a
  no-ceiling sensitivity model (shown in Table 3) should accompany
  the headline numbers in any manuscript.
- **GOLD 4 longitudinal attrition**: only 39 of 593 GOLD 4 subjects
  have V3 ESI.  The apparent ESI *decline* in GOLD 4 over time
  (Figure 4) is survivor bias.
- **Per-stratum independence of ESI vs FEV1/FVC**: ESI's independent
  prognostic contribution is concentrated in GOLD 1–3 (mortality)
  and GOLD 0 (FEV1 decline).  In Never-smokers, PRISm, and GOLD 4
  the per-stratum estimates are null or imprecise.  The
  spectrum-wide pivot is best framed as adding *a continuous,
  emphysema-specific physiological dimension* across the cohort, not
  as a uniform "ESI > FEV1/FVC" claim across every stratum.
- **Prospective exacerbation rates**: live on the Channing cluster
  and are deferred to a follow-up Channing-side analysis.

# Bottom-line for the spectrum-wide pivot

- The substrate (10,169 V1 subjects with valid ESI; 18,893 longitudinal
  FEV1 visit-rows on 10,109 subjects; 10,105 subjects with both V1 ESI
  and a vital-status record, contributing 2,857 deaths over a median
  follow-up of ≈ 10.7 years) is sufficient to support the analyses
  proposed in the May 2026 spectrum-wide design note.
- The empirical headlines that support the pivot are:
  (a) ESI carries information beyond FEV1/FVC in the preserved-
      spirometry strata (Table 2, Figure 1);
  (b) ESI is an independent mortality predictor across the cohort
      (Table 3), and the v10 GOLD 2–4 subtyping finding is reproduced
      and extended to GOLD 1 (Table 4);
  (c) ESI predicts FEV1 decline more strongly than FEV1/FVC in the
      pooled model, with unique-information signal in GOLD 0
      (Tables 5a, 5b, Figure 3);
  (d) Cross-spectrum ESI trajectories follow expected severity
      gradients but with a notable PRISm/GOLD 0 divergence over time
      (Figure 4).
- The most defensible framing for a manuscript: **ESI provides a
  continuous, emphysema-specific physiological dimension that is
  partially independent of FEV1/FVC across the COPDGene spectrum,
  with its incremental prognostic value concentrated in the GOLD 1–3
  band and its incremental decline-prediction value in GOLD 0.**

# Session info

```{r session-info}
sessionInfo()
```
