---
title: "ESI manuscript — verification report"
author: "Reconciliation of manuscript claims against computed artifacts"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      comment = "")   # no "##" gutter on printed output
options(stringsAsFactors = FALSE)
```

# What this report is

Every number printed in the manuscript is declared here as a **registry entry**
naming the artifact and the field that produces it. Each entry is compared
against the artifact as it exists on disk *now*. A number that has drifted —
because the analysis changed and the prose did not, or the prose was edited by
hand — fails loudly.

The manuscript is a `.docx` and the Changit snapshot ships no `manuscript/`
directory, so the report cannot read the document on the cluster. The `claim`
and `expected` columns are therefore a **transcription** of the printed
document, made once, by hand, from:

- `manuscript/ESI manuscript draft v12 2026.8.22_PJC.docx`

The transcription is laid out below to mirror the printed tables so a reader can
check it against the document by eye. Everything downstream of the transcription
is mechanical.

**Scope.** The main manuscript only. The supplement is machine-built from these
same artifacts by `build_supplement.py` (see `build_log_v9_supp.txt`), so its
numbers cannot drift from them. The main document's prose is hand-maintained —
`manifest.py`'s `MANUSCRIPT_OUT` still points at *v11* while the live document is
*v12* — which is where drift actually lives.

```{r config, include = FALSE}
CONFIG <- Sys.getenv("ESI_CONFIG", unset = "config_paths.R")
if (!file.exists(CONFIG)) {
  stop("Missing input-path config: '", CONFIG, "'. ",
       "Copy config_paths.R.example to config_paths.R.")
}
source(CONFIG, local = FALSE)
if (!exists("OUT_DIR")) stop("config '", CONFIG, "' does not define OUT_DIR")
if (!dir.exists(OUT_DIR)) stop("OUT_DIR does not exist: ", OUT_DIR)

MANUSCRIPT_VERSION <- "ESI manuscript draft v12 2026.8.22_PJC.docx"
MANUSCRIPT_PATH    <- file.path("manuscript", MANUSCRIPT_VERSION)
ANALYSIS_SRC       <- "esi_manuscript_analysis_2026.7.17_noILD.Rmd"

# The transcription's trust root. `expected` throughout this registry is a hand
# transcription of the document named above, made once. If the document changes,
# the transcription may no longer describe it and the registry must be re-read
# against it -- so the document is pinned by hash. Editing the manuscript without
# revisiting this file fails here. The Changit snapshot ships no manuscript/
# directory, so on the cluster this degrades to NOT-AVAILABLE rather than
# pretending to check something that is not there.
MANUSCRIPT_MD5 <- "da957e51ef630e97b7fa568f25b12549"

cat("OUT_DIR:  ", OUT_DIR, "\n",
    "Registry transcribed from: ", MANUSCRIPT_VERSION, "\n", sep = "")
```

```{r artifact-loader, include = FALSE}
# --- artifact access ---------------------------------------------------------
# Artifacts are loaded once and cached. `x` inside a registry `field` expression
# is the loaded artifact: a data.frame for .csv, a named character vector for
# the key=value .txt sidecars.
.cache <- new.env(parent = emptyenv())

read_kv <- function(path) {
  ln <- readLines(path, warn = FALSE)
  ln <- ln[nzchar(trimws(ln)) & grepl("=", ln, fixed = TRUE)]
  k  <- trimws(sub("=.*$", "", ln))
  v  <- trimws(sub("^[^=]*=", "", ln))
  setNames(v, k)
}

art <- function(name) {
  if (!exists(name, envir = .cache)) {
    p <- file.path(OUT_DIR, name)
    if (!file.exists(p)) stop("artifact absent: ", name, call. = FALSE)
    obj <- if (grepl("\\.csv$", name)) {
      read.csv(p, check.names = FALSE)
    } else if (grepl("\\.txt$", name)) {
      read_kv(p)
    } else {
      stop("no reader for artifact type: ", name, call. = FALSE)
    }
    assign(name, obj, envir = .cache)
  }
  get(name, envir = .cache)
}

# Pull the point estimate / CI bounds out of a "1.01 (0.96-1.06)" cell.
parse_est <- function(s, which = c("est", "lo", "hi")) {
  which <- match.arg(which)
  # Guard the row selection here rather than downstream: the caller's length-1
  # check sees only what this returns, and the "lo"/"hi" paths take the first
  # element of a strsplit, so an ambiguous selection would collapse to a
  # passing scalar instead of raising.
  if (length(s) != 1L) stop("parse_est: selection returned ", length(s), " rows")
  s <- trimws(as.character(s))
  est <- as.numeric(sub("^([0-9.eE+-]+).*$", "\\1", s))
  ci  <- sub("^[^(]*\\(", "", sub("\\).*$", "", s))
  bd  <- as.numeric(strsplit(ci, "-", fixed = TRUE)[[1]])
  switch(which, est = est, lo = bd[1], hi = bd[2])
}

# The analysis writes p-values through fmt_p(), which emits "<0.001" at the
# floor; strip the "<" so a numeric comparison is possible where the manuscript
# also prints a bounded value.
num_p <- function(s) as.numeric(sub("^<", "", trimws(as.character(s))))
```

```{r registry-builder, include = FALSE}
# --- registry ----------------------------------------------------------------
REG <- list()

# Entries are grouped for display by the section of the manuscript they check.
# Each registry chunk sets .SECTION before calling reg().
.SECTION <- "unassigned"

reg <- function(id, location, claim, expected, artifact, field,
                tolerance, local_only = FALSE, note = "") {
  REG[[length(REG) + 1]] <<- list(
    id = id, section = .SECTION,
    location = location, claim = claim, expected = expected,
    artifact = artifact, field = field, tolerance = tolerance,
    local_only = local_only, note = note)
  invisible(NULL)
}

# Tolerances. Two independent machines computing their own bootstraps agree to
# 1e-8..1e-15 on the headline tables, so the binding constraint is almost always
# the manuscript's own printed precision, not numerical reproducibility.
TOL_2DP  <- 0.005    # a value printed to 2 decimal places
TOL_1DP  <- 0.05     # a value printed to 1 decimal place
TOL_3DP  <- 0.0005   # a value printed to 3 decimal places
# Bootstrap percentile p. This tolerance is set from the Monte Carlo error of
# the statistic, not from a single observed cross-machine difference.
#
# The percentile p is an estimate from B_effective resamples, so its SE is
# sqrt(p(1-p)/B). Two independent runs differ with SE sqrt(2p(1-p)/B): at
# p ~ 0.85 and B = 1,000 that is 0.016, so a ~3 SE spread is ~0.05.
#
# Measured on two machines, laptop vs Channing:
#   all-cause   COPD-minor    848/1000 vs 878/1000   diff 0.030
#   respiratory COPD-minor    292/416  vs 300/412    diff 0.026
#   respiratory COPD-major     61/208  vs  57/206    diff 0.017
#
# The all-cause row used all 1,000 resamples on both machines with none
# dropped, so differing effective resamples cannot explain it; it is sampling
# variation in the resample draw. An earlier ±0.01 estimate was below the noise
# floor of the statistic and failed all three of these legitimately.
TOL_BOOT <- 0.05
TOL_EXACT <- 0
```

```{r reg-cohort, include = FALSE}
.SECTION <- "Cohort and strata"
reg("COH-01", "Results P1; Figure 1 legend", "analytic cohort of 9,402 participants",
    9402, "Table_Agreement_stats.txt", 'as.numeric(x[["n_cohort"]])', TOL_EXACT)

for (s in list(list("COH-02", "Never",  "106 never-smokers",                 106),
               list("COH-03", "GOLD0",  "4,054 smokers without obstruction", 4054),
               list("COH-04", "PRISm",  "1,129 with PRISm",                  1129),
               list("COH-05", "GOLD1-4","4,113 with GOLD 1-4 obstruction",   4113))) {
  fld <- if (s[[2]] == "GOLD1-4") {
    'sum(x$n[x$stratum %in% c("GOLD1","GOLD2","GOLD3","GOLD4")])'
  } else sprintf('x$n[x$stratum == "%s"]', s[[2]])
  reg(s[[1]], "Results P1", s[[3]], s[[4]], "Table_S2_stratum_counts.csv", fld, TOL_EXACT)
}

reg("COH-06", "Results P1", "mean ESI 0.83 in never-smokers",
    0.83, "Table_S2_stratum_counts.csv", 'x$ESI_mean[x$stratum == "Never"]', TOL_2DP)
reg("COH-07", "Results P1", "mean ESI 8.30 in GOLD 4",
    8.30, "Table_S2_stratum_counts.csv", 'x$ESI_mean[x$stratum == "GOLD4"]', TOL_2DP)
```

```{r reg-correlations, include = FALSE}
.SECTION <- "ESI versus quantitative CT"
reg("COR-01", "Results P2", "r = 0.78 with %LAA-950HU, whole cohort",
    0.78, "Table_2.csv", 'x$r_LAA[x$stratum == "All strata"]', TOL_2DP)
reg("COR-02", "Results P2", "r = 0.81 with PRM emphysema, whole cohort",
    0.81, "Table_2_PRMversion.csv", 'x$r_PRM[x$stratum == "All strata"]', TOL_2DP)
reg("COR-03", "Results P2", "r = 0.08 in GOLD 0",
    0.08, "Table_2.csv", 'x$r_LAA[x$stratum == "GOLD0"]', TOL_2DP)
reg("COR-04", "Results P2", "r = 0.58 in GOLD 3",
    0.58, "Table_2.csv", 'x$r_LAA[x$stratum == "GOLD3"]', TOL_2DP)
```

```{r reg-agreement, include = FALSE}
.SECTION <- "Classification agreement"
agree <- list(
  list("AGR-01", "91% overall agreement", 0.91,   "agreement"),
  list("AGR-02", "kappa = 0.82",          0.82,   "kappa"),
  list("AGR-03", "88% sensitivity",       0.88,   "sensitivity"),
  list("AGR-04", "94% specificity",       0.94,   "specificity"))
for (a in agree) {
  reg(a[[1]], "Abstract; Results P3; Figure 1 legend", a[[2]], a[[3]],
      "Table_Agreement_stats.txt", sprintf('as.numeric(x[["%s"]])', a[[4]]), TOL_2DP)
}
```

```{r reg-table1, include = FALSE}
.SECTION <- "Table 1"
T1 <- data.frame(
  outcome  = c("all-cause",  "all-cause",  "respiratory", "respiratory", "exacerbations", "exacerbations"),
  category = c("COPD-minor", "COPD-major", "COPD-minor",  "COPD-major",  "COPD-minor",    "COPD-major"),
  ct_est = c( 1.91,  2.59,  4.80, 36.27, 2.72, 5.05),
  ct_lo  = c( 1.64,  2.35,  2.14, 20.85, 2.36, 4.59),
  ct_hi  = c( 2.22,  2.86, 10.76, 63.07, 3.14, 5.56),
  esi_est= c( 1.94,  2.39,  5.68, 30.95, 2.77, 4.47),
  esi_lo = c( 1.63,  2.18,  2.43, 18.76, 2.33, 4.07),
  esi_hi = c( 2.32,  2.63, 13.32, 51.06, 3.30, 4.90),
  p      = c("0.85", "<0.002", "0.70", "0.29", "0.80", "<0.002"))

t1_src <- c("all-cause"     = "Table_8_allcause.csv",
            "respiratory"   = "Table_8_resp.csv",
            "exacerbations" = "Table_Exacerbations_Bhatt.csv")
t1_pfx <- c("all-cause" = "bhatt_HR", "respiratory" = "bhatt_HR",
            "exacerbations" = "bhatt_IRR")

for (i in seq_len(nrow(T1))) {
  r   <- T1[i, ]
  src <- t1_src[[r$outcome]]
  # column stems: Table_8_* use bhatt_HR/esi_HR; the exacerbation table uses IRR
  stem <- if (r$outcome == "exacerbations") "IRR" else "HR"
  for (fw in c("bhatt", "esi")) {
    for (part in c("est", "lo", "hi")) {
      col <- if (part == "est") paste0(fw, "_", stem)
             else paste0(fw, "_", c(lo = "LCI", hi = "UCI")[[part]])
      exp_col <- paste0(if (fw == "bhatt") "ct_" else "esi_", part)
      reg(sprintf("T1-%s-%s-%s-%s", r$outcome, r$category,
                  if (fw == "bhatt") "CT" else "ESI", part),
          "Table 1",
          sprintf("%s, %s, %s-based %s", r$outcome, r$category,
                  if (fw == "bhatt") "CT" else "ESI", part),
          r[[exp_col]], src,
          sprintf('x[["%s"]][x$group == "%s"]', col, r$category),
          TOL_2DP)
    }
  }
}

# The CT vs ESI p column comes from the bootstrap difference tables. Compare the
# QUANTITY, not the formatted string: `two_sided_p_reported` is a formatting
# decision, so a string comparison would verify only that the formatter still
# formats, and it would also reject a legitimate cluster re-run that landed on
# 0.71 instead of 0.702 -- exactly the ±0.01 slack TOL_BOOT exists to allow.
#
# The two "<0.002" cells are a different claim. There `two_sided_p` is NA,
# because no resample crossed zero, which is what "<0.002" is asserting. So
# check that assertion directly: the percentile interval must not span zero.
t1p_src <- c("all-cause"     = "Supp_Table_HR_Difference_Bootstrap.csv",
             "respiratory"   = "Supp_Table_HR_Difference_Bootstrap.csv",
             "exacerbations" = "Supp_Table_IRR_Difference_Bootstrap.csv")
for (i in seq_len(nrow(T1))) {
  r   <- T1[i, ]
  sel <- sprintf('x$outcome == "%s" & x$category == "%s"', r$outcome, r$category)
  stem <- if (r$outcome == "exacerbations") "logIRR" else "logHR"
  if (grepl("^<", r$p)) {
    reg(sprintf("T1-p-%s-%s", r$outcome, r$category), "Table 1",
        sprintf('CT vs ESI p "%s" (no resample crossed zero), %s, %s',
                r$p, r$outcome, r$category),
        TRUE, t1p_src[[r$outcome]],
        sprintf('{ i <- which(%s); is.na(x$two_sided_p[i]) && x$ci_lo_%s[i] * x$ci_hi_%s[i] > 0 }',
                sel, stem, stem),
        TOL_EXACT)
  } else {
    reg(sprintf("T1-p-%s-%s", r$outcome, r$category), "Table 1",
        sprintf("CT vs ESI p, %s, %s", r$outcome, r$category),
        as.numeric(r$p), t1p_src[[r$outcome]],
        sprintf('x$two_sided_p[%s]', sel), TOL_BOOT)
  }
}
```

```{r reg-table2, include = FALSE}
.SECTION <- "Table 2"
T2 <- data.frame(
  group = c("CT-only-COPD", "ESI-only-COPD", "Both-COPD"),
  src   = c("CT-only-COPD (ESI missed)", "ESI-only-COPD (Bhatt missed)", "Both-COPD"),
  all_est = c(1.53, 1.26, 2.02), all_lo = c(1.25, 0.76, 1.66), all_hi = c( 1.88,  2.07,  2.46),
  rsp_est = c(2.90, 4.86, 6.10), rsp_lo = c(0.91, 0.62, 2.19), rsp_hi = c( 9.28, 38.02, 16.96),
  exa_est = c(2.00, 1.62, 3.15), exa_lo = c(1.60, 0.97, 2.52), exa_hi = c( 2.51,  2.68,  3.93))

for (i in seq_len(nrow(T2))) {
  r <- T2[i, ]
  # all-cause and respiratory mortality live in one artifact, exacerbations in another
  for (blk in list(list("all", "Table_BhattOnly_vs_Both.csv", c(est="all_HR",  lo="all_LCI",  hi="all_UCI"),  "all-cause mortality HR"),
                   list("rsp", "Table_BhattOnly_vs_Both.csv", c(est="resp_HR", lo="resp_LCI", hi="resp_UCI"), "respiratory mortality HR"),
                   list("exa", "Table_Exacerbations_Discordance.csv", c(est="IRR", lo="LCI", hi="UCI"), "exacerbation IRR"))) {
    for (part in c("est", "lo", "hi")) {
      reg(sprintf("T2-%s-%s-%s", r$group, blk[[1]], part), "Table 2",
          sprintf("%s, %s, %s", r$group, blk[[4]], part),
          r[[paste0(blk[[1]], "_", part)]], blk[[2]],
          sprintf('x[["%s"]][x$group == "%s"]', blk[[3]][[part]], r$src),
          TOL_2DP)
    }
  }
}

# Table 2 legend: pairwise adjusted p-values among the three COPD-positive groups.
reg("T2-p-both-vs-ct", "Table 2 legend",
    "Both-COPD vs CT-only exacerbation IRR, p = 0.005", 0.005,
    "Table_2_pairwise_contrasts.csv",
    'x$p_adj[x$outcome == "exacerbations" & x$contrast == "Both-COPD - CT-only-COPD"]',
    TOL_3DP)
reg("T2-p-both-vs-esi", "Table 2 legend",
    "Both-COPD vs ESI-only exacerbation IRR, p = 0.037", 0.037,
    "Table_2_pairwise_contrasts.csv",
    'x$p_adj[x$outcome == "exacerbations" & x$contrast == "Both-COPD - ESI-only-COPD"]',
    TOL_3DP)

# Table 2 legend: model sample sizes. These are model Ns, taken off the fitted
# objects by the `cross-class-exac` chunk. They are NOT the d_b-level subgroup
# size, which is why the cross-check below is registered separately.
reg("T2-n-mortality", "Table 2 legend", "mortality models fit on 5,289 participants",
    5289, "Table_2_model_Ns.txt", 'as.numeric(x[["n_mortality_model"]])', TOL_EXACT)
reg("T2-n-exacerbation", "Table 2 legend", "exacerbation model fit on 4,635 participants",
    4635, "Table_2_model_Ns.txt", 'as.numeric(x[["n_exacerbation_model"]])', TOL_EXACT)

# Cross-check: the preserved-spirometry subgroup at the d_b level is the noCOPD
# and COPD-minor rows of the cross-tab. If this equals the mortality model N,
# no participant was lost to the vital-status join or covariate completeness.
reg("T2-n-mortality-xcheck", "Table 2 legend (cross-check)",
    "preserved-spirometry subgroup size from the cross-tab equals the mortality model N",
    5289, "Table_6.csv",
    'sum(as.matrix(x[x$Bhatt %in% c("noCOPD","COPD-minor"), -1]))', TOL_EXACT)
```

```{r reg-prose, include = FALSE}
.SECTION <- "Results and Discussion prose"
reg("PRO-01", "Results P5", "AFL-only-noCOPD FEV1 +10.1 mL/yr under the ESI framework",
    10.1, "Table_BhattDecline.csv", 'x$esi_est[x$group == "AFL-only-NoCOPD"]', TOL_1DP)
reg("PRO-02", "Results P5", "AFL-only-noCOPD FEV1 decline p = 0.02",
    0.02, "Table_BhattDecline.csv", 'x$esi_p[x$group == "AFL-only-NoCOPD"]', TOL_2DP)

# The artifact stores the slope as a signed estimate (-4.73); the manuscript
# states it as a decline of 4.73 mL/yr. Compare magnitudes and check the sign
# separately so a sign flip cannot pass silently.
reg("PRO-03", "Results P6", "GOLD 0: 4.73 mL/yr additional FEV1 decline per 1-unit ESI",
    4.73, "Supp_Table_Continuous_FEV1Decline_GOLD0.txt",
    'abs(as.numeric(x[["gold0_esi_est"]]))', TOL_2DP)
reg("PRO-04", "Results P6", "that GOLD 0 slope is negative, i.e. a decline",
    -1, "Supp_Table_Continuous_FEV1Decline_GOLD0.txt",
    'sign(as.numeric(x[["gold0_esi_est"]]))', TOL_EXACT)
reg("PRO-05", "Results P6", "GOLD 0 FEV1-decline P = 0.001",
    0.001, "Supp_Table_Continuous_FEV1Decline_GOLD0.txt",
    'as.numeric(x[["gold0_esi_p"]])', TOL_3DP)

reg("PRO-06", "Results P6; Discussion P3", "mean bronchodilator dESI = -0.09",
    -0.09, "Supp_Bronchodilator_deltaESI.txt", 'as.numeric(x[["mean_delta"]])', TOL_2DP)
reg("PRO-07", "Discussion P3", "n = 10,160 paired pre/post-bronchodilator measurements",
    10160, "Supp_Bronchodilator_deltaESI.txt", 'as.numeric(x[["n_paired"]])', TOL_EXACT,
    local_only = TRUE,
    note = paste("Local-only. 10,160 locally, 10,397 on the cluster: the ESI extracts",
                 "differ OUTSIDE the analytic cohort. d_b is 9,402 in both places, so",
                 "no analytic-cohort claim is affected. Not reproducible across sites."))

reg("PRO-08", "Discussion P8", "continuous ESI adjusted for FEV1/FVC, IRR = 1.01",
    1.01, "Table_S6b_continuous_exacerbations.csv",
    'parse_est(x$ESI_IRR[x$model == "ESI + FEV1/FVC"], "est")', TOL_2DP)
reg("PRO-09", "Discussion P8", "continuous ESI adjusted for FEV1/FVC, p = 0.727",
    0.727, "Table_S6b_continuous_exacerbations.csv",
    'num_p(x$ESI_p[x$model == "ESI + FEV1/FVC"])', TOL_3DP)

reg("PRO-10", "Discussion, limitations", "ESI-only-COPD subgroup n = 94",
    94, "Table_BhattOnly_vs_Both.csv",
    'x$n[x$group == "ESI-only-COPD (Bhatt missed)"]', TOL_EXACT)
reg("PRO-11", "Discussion, limitations", "AFL-only-noCOPD subgroup n = 170",
    170, "Table_6.csv",
    'sum(as.numeric(x[x$Bhatt == "AFL-only-NoCOPD", -1]))', TOL_EXACT)
```

```{r reg-significance, include = FALSE}
.SECTION <- "Significance claims"
for (a in list(list("SIG-01", "all-cause mortality", "Table_8_allcause.csv",   c("bhatt_p", "esi_p")),
               list("SIG-02", "respiratory mortality", "Table_8_resp.csv",     c("bhatt_p", "esi_p")),
               list("SIG-03", "exacerbations", "Table_Exacerbations_Bhatt.csv", c("bhatt_p", "esi_p")))) {
  reg(a[[1]], "Results P4; Discussion P7",
      sprintf("AFL-only-noCOPD not significantly associated with %s under either framework", a[[2]]),
      TRUE, a[[3]],
      sprintf('all(c(x$%s, x$%s)[x$group == "AFL-only-NoCOPD"] > 0.05)', a[[4]][1], a[[4]][2]),
      TOL_EXACT)
}
reg("SIG-04", "Results P5",
    "AFL-only-noCOPD FEV1 increase is significant under ESI but NOT under the CT framework",
    TRUE, "Table_BhattDecline.csv",
    '{ i <- which(x$group == "AFL-only-NoCOPD"); x$esi_p[i] < 0.05 && x$bhatt_p[i] >= 0.05 }',
    TOL_EXACT)
```

```{r reg-methods, include = FALSE}
.SECTION <- "Methods constants"
reg("MET-01", "Methods, equivalence", "C-index equivalence margin +/-0.02, all-cause",
    TRUE, "Table_1_Cindex_Equivalence.csv",
    'abs(x$delta_c[x$outcome == "all-cause"]) < 0.02 && x$equivalent_within_margin[x$outcome == "all-cause"]',
    TOL_EXACT)
reg("MET-02", "Methods, equivalence", "C-index equivalence margin +/-0.02, respiratory",
    TRUE, "Table_1_Cindex_Equivalence.csv",
    'abs(x$delta_c[x$outcome == "respiratory"]) < 0.02 && x$equivalent_within_margin[x$outcome == "respiratory"]',
    TOL_EXACT)
```

Three Methods claims are constants in the analysis source rather than values in
an artifact: the two ESI thresholds and the bootstrap resample count. They are
checked by reading the source.

```{r reg-constants, include = FALSE}
.SECTION <- "Methods constants"
src_lines <- readLines(ANALYSIS_SRC, warn = FALSE)
grab_const <- function(pattern) {
  hit <- grep(pattern, src_lines, value = TRUE)
  if (!length(hit)) return(NA_real_)
  as.numeric(sub("^.*<-\\s*([0-9.]+).*$", "\\1", hit[1]))
}
CONSTS <- c(ESI_T_LOW   = grab_const("^ESI_T_LOW\\s*<-"),
            ESI_T_HIGH  = grab_const("^ESI_T_HIGH\\s*<-"),
            B_BOOTSTRAP = grab_const("^B_BOOTSTRAP\\s*<-"),
            RPART_SEED  = grab_const("^RPART_SEED\\s*<-"))

# The training fraction is an argument to sample(), not a named constant.
train_line  <- grep("train_idx\\s*<-\\s*sample\\(", src_lines, value = TRUE)
CONSTS[["TRAIN_FRAC"]] <- if (length(train_line))
  as.numeric(sub("^.*size\\s*=\\s*round\\(([0-9.]+)\\s*\\*.*$", "\\1", train_line[1])) else NA_real_

const_reg <- list(
  list("MET-03", "Methods, ESI thresholds", "lower threshold ESI = 1",   1.0,  "ESI_T_LOW"),
  list("MET-04", "Methods, ESI thresholds", "upper threshold ESI = 2.5", 2.5,  "ESI_T_HIGH"),
  list("MET-05", "Methods, equivalence",    "paired bootstrap B = 1,000", 1000, "B_BOOTSTRAP"),
  list("MET-07", "Methods, ESI thresholds",
       "thresholds derived in a randomly selected 80% training cohort", 0.8, "TRAIN_FRAC"))
for (c1 in const_reg) {
  reg(c1[[1]], c1[[2]], c1[[3]], c1[[4]], "<analysis source>",
      sprintf('CONSTS[["%s"]]', c1[[5]]), TOL_EXACT)
}

# The MD-COPD criterion constants define the framework the whole paper is about.
# They are literals in the analysis, and if one moved, every artifact would move
# with it -- the tables would stay self-consistent and nothing else in this
# registry would notice. So they are checked against the source, like the ESI
# thresholds.
crit <- list(
  list("MET-08", "post-bronchodilator FEV1/FVC below 0.70",       0.70,
       "major_criterion\\s*=\\s*FEV1_FVC_post\\s*<\\s*([0-9.]+)"),
  list("MET-09", "visual emphysema Fleischner at least mild (>= 1)", 1,
       "emph_yn\\s*=\\s*CT_Visual_Emph_Severity\\s*>=\\s*([0-9.]+)"),
  list("MET-10", "airway wall thickening Fleischner definite (== 2)", 2,
       "wall_yn\\s*=\\s*CT_Visual_Wall_Thickening\\s*==\\s*([0-9.]+)"),
  list("MET-11", "mMRC dyspnea score of 2 or greater",            2,
       "dysp_yn\\s*=\\s*MMRCDyspneaScor\\s*>=\\s*([0-9.]+)"),
  list("MET-12", "SGRQ total score of 25 or greater",             25,
       "qol_yn\\s*=\\s*SGRQ_scoreTotal\\s*>=\\s*([0-9.]+)"),
  list("MET-13", "COPD-major: major criterion plus at least 1 of 5 minor", 1,
       "major_criterion\\s*&\\s*n_minor_b\\s*>=\\s*([0-9.]+)"),
  list("MET-14", "COPD-minor: at least 3 of 5 minor without the major", 3,
       "!d_b\\$major_criterion\\s*&\\s*n_minor_b\\s*>=\\s*([0-9.]+)"))
for (cc in crit) {
  hit <- grep(cc[[4]], src_lines, value = TRUE)
  CONSTS[[cc[[1]]]] <- if (length(hit))
    as.numeric(sub(paste0("^.*", cc[[4]], ".*$"), "\\1", hit[1])) else NA_real_
  reg(cc[[1]], "Methods, MD-COPD framework", cc[[2]], cc[[3]],
      "<analysis source>", sprintf('CONSTS[["%s"]]', cc[[1]]), TOL_EXACT)
}

# --- covariate sets ----------------------------------------------------------
# Both table legends and Methods state which covariates each model adjusts for.
# Nothing else in this registry would notice if a model formula changed: every
# artifact would move together, the tables would stay internally consistent, and
# the prose would silently stop describing the code. Same failure shape as the
# criterion constants above, so it is checked the same way, against the source.
COVARIATES <- c("age_visit", "gender", "gender_baseline", "race", "race_baseline",
                "SmokCigNow", "ATS_PackYears", "BMI", "Height_CM", "stratum")

covars_of <- function(anchor) {
  i <- grep(anchor, src_lines)
  if (!length(i)) return("<model not found>")
  # A formula may wrap over several lines; accumulate until parentheses balance.
  txt <- src_lines[i[1]]
  j <- i[1]
  while (nchar(gsub("[^(]", "", txt)) != nchar(gsub("[^)]", "", txt)) &&
         j < length(src_lines)) {
    j <- j + 1
    txt <- paste(txt, src_lines[j])
  }
  rhs <- sub("^[^~]*~", "", txt)
  toks <- unlist(strsplit(rhs, "[+*(),|]"))
  toks <- trimws(gsub("[^A-Za-z0-9_]", "", toks))
  # radix, not the default: sort() is locale-collated, so the same formula
  # would render in a different order on a machine with a different locale and
  # the comparison would fail for no reason. radix is C-order everywhere.
  paste(sort(unique(intersect(toks, COVARIATES)), method = "radix"),
        collapse = ", ")
}

PRIMARY_SET <- "ATS_PackYears, BMI, SmokCigNow, age_visit, gender, race"
DECLINE_SET <- "ATS_PackYears, Height_CM, SmokCigNow, age_visit, gender_baseline, race_baseline"
CONT_SET    <- "ATS_PackYears, SmokCigNow, age_visit, gender, race, stratum"

cov_reg <- list(
  list("COV-01", "Table 1 legend", "all-cause mortality, CT-based",  PRIMARY_SET, "^cox_bhatt_all\\s*<-"),
  list("COV-02", "Table 1 legend", "all-cause mortality, ESI-based", PRIMARY_SET, "^cox_esi_all\\s*<-"),
  list("COV-03", "Table 1 legend", "respiratory mortality, CT-based", PRIMARY_SET, "^cox_bhatt_resp\\s*<-"),
  list("COV-04", "Table 1 legend", "exacerbations, CT-based",        PRIMARY_SET, "^nbB\\s*<-"),
  list("COV-05", "Table 1 legend", "exacerbations, ESI-based",       PRIMARY_SET, "^nbE\\s*<-"),
  list("COV-06", "Table 2 legend", "cross-classification mortality", PRIMARY_SET, "^cox_disc_all\\s*<-"),
  list("COV-07", "Table 2 legend", "cross-classification exacerbations", PRIMARY_SET, "^nb_disc\\s*<-"),
  list("COV-08", "Methods, adjustment sets", "FEV1 decline, CT-based: height replaces BMI",
       DECLINE_SET, "^lmm_bhatt\\s*<-"),
  list("COV-09", "Methods, adjustment sets", "FEV1 decline, ESI-based: height replaces BMI",
       DECLINE_SET, "^lmm_esi\\s*<-"),
  list("COV-10", "Methods, adjustment sets", "GOLD 0 continuous-ESI decline: height replaces BMI",
       DECLINE_SET, "^lmm_g0\\s*<-"),
  list("COV-11", "Methods, adjustment sets", "continuous-ESI mortality omits BMI",
       CONT_SET, "^fit_cox_esi\\s*<-"),
  list("COV-12", "Methods, adjustment sets", "continuous-ESI exacerbations omits BMI",
       CONT_SET, "^nb_esi_ex\\s*<-"))

# Resolve here rather than embedding the anchor regex in the field expression:
# a "\\s" inside generated R source is not a valid string escape.
COVSET <- vapply(cov_reg, function(cv) covars_of(cv[[5]]), character(1))
names(COVSET) <- vapply(cov_reg, function(cv) cv[[1]], character(1))

for (cv in cov_reg) {
  reg(cv[[1]], cv[[2]],
      sprintf("adjustment covariates, %s", cv[[3]]),
      cv[[4]], "<analysis source>",
      sprintf('COVSET[["%s"]]', cv[[1]]), TOL_EXACT)
}

# The manuscript's R version is a property of the run that produced the
# artifacts, not of whatever interpreter renders this report, so read it off the
# provenance stamp the analysis writes.
#
# Waived like PRO-07 rather than hard-failed. The claim is about the canonical
# run that produced the published numbers; a reproduction elsewhere legitimately
# uses a different R (the cluster is 4.3.2, and the README records that the
# results reproduce there) and that does not falsify the manuscript. This stays
# visible rather than silent: verify.R names every waived entry on each run.
reg("MET-06", "Methods, statistical analysis", "all analyses performed in R version 4.5.2",
    "4.5.2", "PROVENANCE.txt", 'x[["r_version"]]', TOL_EXACT,
    local_only = TRUE,
    note = paste("Waived off the canonical machine. The manuscript's claim is",
                 "about the run that produced the published numbers. A",
                 "reproduction on another R shows up here as LOCAL-ONLY with",
                 "the actual version in the computed column - read it."))
```

# Verdict

```{r evaluate, include = FALSE}
rows <- lapply(REG, function(e) {
  # Fields are evaluated in an environment whose parent is baseenv(), NOT the
  # global environment. If globals were reachable a field could name T1 or T2 --
  # the transcription itself -- and check the answer key against itself.
  computed <- tryCatch({
    env <- new.env(parent = baseenv())
    env$parse_est <- parse_est; env$num_p <- num_p
    if (e$artifact == "<analysis source>") {
      env$CONSTS <- CONSTS; env$COVSET <- COVSET
    } else {
      env$x <- art(e$artifact)
    }
    eval(parse(text = e$field), env)
  }, error = function(err) structure(NA, msg = conditionMessage(err)))

  msg <- attr(computed, "msg")
  status <-
    if (!is.null(msg)) {
      "ERROR"
    } else if (length(computed) != 1L) {
      "ERROR"
    } else if (is.logical(e$expected)) {
      # Both sides must be logical. Accepting a number here would let a field
      # that lost its comparison -- `abs(delta_c)` where `abs(delta_c) < 0.02`
      # was meant -- report PASS on the strength of being non-zero.
      if (!is.logical(computed)) "ERROR"
      else if (isTRUE(computed) == isTRUE(e$expected)) "PASS" else "FAIL"
    } else if (is.logical(computed)) {
      "ERROR"
    } else if (is.character(e$expected)) {
      if (identical(trimws(as.character(computed)), trimws(e$expected))) "PASS" else "FAIL"
    } else {
      if (is.na(suppressWarnings(as.numeric(computed)))) "ERROR"
      else if (abs(as.numeric(computed) - e$expected) <= e$tolerance) "PASS" else "FAIL"
    }

  # local_only waives the NUMERIC COMPARISON only. An entry that cannot be
  # evaluated at all is not "verified with a caveat", it is unverified, and it
  # must stay red.
  if (e$local_only && status == "FAIL") status <- "LOCAL-ONLY"

  data.frame(
    id = e$id, section = e$section, location = e$location, claim = e$claim,
    expected  = as.character(e$expected),
    artifact  = e$artifact, field = e$field,
    tolerance = e$tolerance,
    computed  = if (!is.null(msg)) paste0("<", msg, ">")
                else if (length(computed) != 1L) paste0("<returned ", length(computed), " values>")
                else as.character(computed),
    status    = status,
    note      = e$note)
})
V <- do.call(rbind, rows)

# --- provenance -------------------------------------------------------------
# Two trust-root checks, appended as ordinary registry rows so that they reach
# the gate. File mtimes are not usable for this: a git clone on the cluster
# stamps every file with checkout time in arbitrary order, so an mtime
# comparison there is noise. Content hashes are.
prov_row <- function(id, claim, expected, computed, status, note = "") {
  data.frame(id = id, section = "Provenance", location = "provenance", claim = claim,
             expected = expected, artifact = "PROVENANCE.txt", field = "-",
             tolerance = 0, computed = computed, status = status, note = note)
}

src_md5 <- unname(tools::md5sum(ANALYSIS_SRC)[[1]])
P <- tryCatch(art("PROVENANCE.txt"), error = function(e) NULL)
V <- rbind(V, if (is.null(P)) {
  prov_row("PRV-01", "artifacts were produced by the current analysis source",
           src_md5, "<PROVENANCE.txt absent>", "ERROR",
           "The analysis has not been run since the provenance stamp was added.")
} else if (identical(unname(P[["analysis_md5"]]), src_md5)) {
  prov_row("PRV-01", "artifacts were produced by the current analysis source",
           src_md5, P[["analysis_md5"]], "PASS")
} else {
  prov_row("PRV-01", "artifacts were produced by the current analysis source",
           src_md5, P[["analysis_md5"]], "FAIL",
           paste("These artifacts came from a different version of",
                 ANALYSIS_SRC, "- every number below was reconciled against a",
                 "stale run. Re-render the analysis."))
})

V <- rbind(V, if (!file.exists(MANUSCRIPT_PATH)) {
  prov_row("PRV-02", "registry was transcribed from the pinned manuscript",
           MANUSCRIPT_MD5, "<manuscript absent>", "NOT-AVAILABLE",
           "No manuscript/ directory here, so the transcription's trust root cannot be checked.")
} else {
  m <- unname(tools::md5sum(MANUSCRIPT_PATH)[[1]])
  prov_row("PRV-02", "registry was transcribed from the pinned manuscript",
           MANUSCRIPT_MD5, m,
           if (identical(m, MANUSCRIPT_MD5)) "PASS" else "FAIL",
           if (identical(m, MANUSCRIPT_MD5)) "" else
             paste("The manuscript has changed since the registry was",
                   "transcribed from it. Re-read the document against the",
                   "registry, then update MANUSCRIPT_MD5."))
})
V$artifact[V$id == "PRV-02"] <- MANUSCRIPT_VERSION

write.csv(V, file.path(OUT_DIR, "VERIFICATION.csv"), row.names = FALSE)
```

```{r verdict-banner}
n_fail <- sum(V$status %in% c("FAIL", "ERROR"))
n_pass <- sum(V$status == "PASS")
cat(if (n_fail == 0)
      sprintf("VERIFICATION PASSED - all %d manuscript claims reconcile.\n", nrow(V))
    else
      sprintf("VERIFICATION FAILED - %d of %d manuscript claims do not reconcile.\n",
              n_fail, nrow(V)))
cat(strrep("-", 62), "\n")
cat(sprintf("  manuscript   %s\n", MANUSCRIPT_VERSION))
cat(sprintf("  analysis     %s\n", ANALYSIS_SRC))
cat(sprintf("  artifacts    %s\n", OUT_DIR))
cat(sprintf("  checked at   %s\n", format(Sys.time(), "%Y-%m-%d %H:%M")))
cat(strrep("-", 62), "\n")
tally <- table(factor(V$status,
                      levels = c("PASS", "FAIL", "ERROR", "LOCAL-ONLY", "NOT-AVAILABLE")))
for (k in names(tally)) if (tally[[k]] > 0) cat(sprintf("  %-14s %3d\n", k, tally[[k]]))
```

Anything not marked PASS is listed first, then every claim is shown grouped by
the part of the manuscript it comes from. The machine-readable record, including
the exact field expression behind each number, is written to
`VERIFICATION.csv` in the artifact directory.

## Anything that did not pass

```{r failures}
bad <- V[V$status != "PASS", ]
if (!nrow(bad)) {
  cat("Nothing. Every claim reconciled.\n")
} else {
  knitr::kable(bad[, c("id", "status", "location", "claim", "expected", "computed")],
               row.names = FALSE,
               col.names = c("ID", "Status", "Where", "Claim", "Manuscript says", "Artifact says"))
}
```

```{r failure-notes, results = "asis"}
if (nrow(bad)) for (i in seq_len(nrow(bad))) if (nzchar(bad$note[i]))
  cat("\n**", bad$id[i], "** ", bad$note[i], "\n", sep = "")
```

## Every claim, by manuscript section

```{r by-section, results = "asis"}
show <- function(d) {
  print(knitr::kable(d[, c("id", "claim", "expected", "computed", "tolerance", "status")],
                     row.names = FALSE,
                     col.names = c("ID", "Claim", "Manuscript says", "Artifact says",
                                   "Tolerance", "Status")))
}
for (sec in unique(V$section)) {
  d <- V[V$section == sec, ]
  cat(sprintf("\n### %s (%d of %d passing)\n\n",
              sec, sum(d$status == "PASS"), nrow(d)))
  show(d)
  cat("\n")
}
```

## The transcription, as printed

`Manuscript says` above is a hand transcription of the document, made once.
The two tables are reproduced here in the shape they appear in the paper so
that the transcription itself can be checked by eye against the page.

```{r transcription}
knitr::kable(T1, row.names = FALSE,
             caption = "Table 1 as printed in the manuscript.")
knitr::kable(T2[, setdiff(names(T2), "src")], row.names = FALSE,
             caption = "Table 2 as printed in the manuscript.")
```

## Appendix: how each number was obtained

The `field` column is the expression evaluated against the named artifact. This
is the audit trail: it is what makes each claim traceable back to the code that
produced it.

```{r appendix}
knitr::kable(V[, c("id", "artifact", "field", "computed", "status")],
             row.names = FALSE,
             col.names = c("ID", "Artifact", "Field expression", "Computed", "Status"))
```

## How to run this

```
Rscript verify.R
```

`verify.R` renders this document and then exits non-zero if any claim did not
reconcile. The gate lives there rather than in this file so that a failing run
still leaves a readable report behind.
