---
title: "If the CT criteria are removed from MD-COPD, what should replace them? A head-to-head comparison of candidate replacement criteria (2026-09-07)"
author: "Peter J. Castaldi"
date: "2026-09-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

<!--
CO-AUTHOR REPORT. Side analysis, not part of the manuscript pipeline: nothing
here is read by ESI_ctfree_analysis_2026.9.3.Rmd, by CLAIMS.md, by verify.R or
by the document builders, and no number here appears in the manuscript.

It exists to answer one question a reader of the CT-free paper will ask, and to
answer it in the open rather than in a reviewer response: the paper removes the
two CT criteria from MD-COPD and replaces them with ESI. Is ESI the right
replacement, and by what standard?

Written against manuscript draft v1 (ctfree/manuscript/), at Pete's direction.

Recomputes, from source, everything in:
  side_analyses/replacement_criteria_comparison.R
  side_analyses/fevfvc_as_criterion.R
and checks itself against the committed
  side_analyses/replacement_criteria_comparison.csv
in Section 8, so drift between this report and those scripts fails loudly.
-->

<style>
body { font-size: 18px; line-height: 1.65; }
table { font-size: 16px; }
pre code { font-size: 15px; }
#TOC { font-size: 15px; }
</style>

# Overview {-}

The CT-free manuscript takes Bhatt et al.'s multidimensional COPD framework
(MD-COPD), removes its two CT criteria, and puts the Emphysema Severity Index
(ESI) in their place. This report asks whether ESI is the right thing to put
there, by fitting every plausible replacement the same way and judging them
all on the same three yardsticks.

**The three findings, stated up front.**

1. **Macro-averaged F1 and "recovers the people CT was needed for" are
   different objectives, and the paper optimizes the first while arguing the
   second.** A second FEV1/FVC threshold beats ESI on held-out macro-F1 while
   recovering fewer of the participants who reached COPD-major only through
   CT. Quantitative CT itself, the actual structural measurement, scores
   nearly worst on macro-F1 while recovering the most. Two of the candidates
   rank in opposite orders on the two yardsticks, which is close to a proof
   that macro-F1 is the wrong lens for this particular question.

2. **Every spirometric replacement repairs the AFL-only label, not just ESI.**
   With no replacement, AFL-only carries a large excess of respiratory
   mortality and the label is false. Every candidate tested brings that
   interval back across 1. The paper's claim that removing CT breaks the
   AFL-only label and a spirometric criterion repairs it is robust. A claim
   that ESI specifically is *required* to repair it is not supported by
   anything here.

3. **Reach into preserved spirometry is where ESI differs from a second
   FEV1/FVC threshold specifically** -- not from spirometric alternatives in
   general. A second FEV1/FVC threshold must sit below 0.70 or it fires for
   everyone with airflow limitation, so it reaches nobody with preserved
   spirometry. ESI reaches a substantial group. But FEV1 percent predicted
   reaches a larger one, so "ESI is the only CT-free criterion with reach"
   would be an overstatement.

Sections 3 to 5 give each of these its data, its method and its numbers.
Section 6 states what the paper may and may not claim on this evidence.
Section 7 records the one question this analysis does not settle.

# PATHS -- supplied at run time

Same six COPDGene inputs the manuscript uses. Copy `config_paths.R.example`
to `config_paths.R` at the repository root, fill in the local file locations,
and knit.

```{r paths, fig.keep = "none"}
# Locate this report's own directory so it runs from any working directory.
# knitr sets the working directory to the .Rmd's location during rendering, so
# getwd() is the primary signal; the --file= fallback covers a plain
# `Rscript -e rmarkdown::render(...)` invocation from elsewhere.
.b <- grep("^--file=", commandArgs(trailingOnly = FALSE), value = TRUE)
SELF <- "replacement_criteria_2026.9.7.Rmd"
HERE <- {
  if (file.exists(SELF)) normalizePath(".")
  else if (length(.b) && file.exists(file.path(
             dirname(normalizePath(sub("^--file=", "", .b[1]))), SELF)))
    dirname(normalizePath(sub("^--file=", "", .b[1])))
  else if (file.exists(file.path("ctfree", "side_analyses", SELF)))
    normalizePath(file.path("ctfree", "side_analyses"))
  else stop("cannot locate this report's directory; knit from ",
            "ctfree/side_analyses/ or from the repository root")
}
CTFREE <- dirname(HERE)
ROOT   <- dirname(CTFREE)
CONFIG <- file.path(ROOT, "config_paths.R")
if (!file.exists(CONFIG))
  stop("cannot find ", CONFIG, "; copy config_paths.R.example to ",
       "config_paths.R and fill in the local file paths")
source(CONFIG)

# Figure styling comes from the ESI project's single source of truth. No font
# size or palette entry is retyped here; a change in style.R reaches this
# report on the next render.
source(file.path(ROOT, "figures", "style.R"))

FIGDIR <- file.path(HERE, "figures")
dir.create(FIGDIR, recursive = TRUE, showWarnings = FALSE)

# ---- Analytic constants, matching replacement_criteria_comparison.R -------
CV_REPEATS <- 3L
CV_FOLDS   <- 5L
CV_SEED    <- 20260901L
```

```{r setup, include = FALSE, fig.keep = "none"}
# Warnings printed as they happen rather than buffered to the end of the knit,
# so a warning is attributable to the chunk that raised it.
options(warn = 1, stringsAsFactors = FALSE)
suppressPackageStartupMessages({
  library(dplyr); library(survival); library(ggplot2)
  library(ggrepel); library(knitr)
})
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = NATIVE_W, fig.align = "center", dpi = 150)

# Number formatting used throughout the prose, so no value is ever typed.
n_fmt  <- function(x) formatC(x, format = "d", big.mark = ",")
f3     <- function(x) sprintf("%.3f", x)
f2     <- function(x) sprintf("%.2f", x)
ci_fmt <- function(e, lo, hi) sprintf("%.2f (%.2f to %.2f)", e, lo, hi)
```

# Section 1 -- The question, the cohort, and the yardsticks

## Section 1.1 -- The question

MD-COPD classifies a participant on one major criterion (post-bronchodilator
FEV~1~/FVC < 0.70, here called **airflow limitation**, AFL) and a count of
minor criteria. In the published framework the minor criteria are three
symptom-based items and two visual CT findings (emphysema and airway wall
thickening). The four categories are:

| Category | Definition |
|---|---|
| **noCOPD** | no airflow limitation, and fewer than *k* minor criteria met |
| **AFL-only** | airflow limitation, and no minor criterion met |
| **COPD-minor** | no airflow limitation, but at least *k* minor criteria met |
| **COPD-major** | airflow limitation, and at least one minor criterion met |

Remove the two CT criteria and two things happen. Participants who reached
COPD-major only through a CT finding fall back to AFL-only, and AFL-only --
which in the published framework is a genuinely low-risk category -- fills up
with people who are not low risk. The manuscript's answer is to put ESI in the
place the CT criteria vacated. **This report asks what else could go there,
and how the candidates compare when each is given the same freedom to fit.**

## Section 1.2 -- Cohort

Identical construction to the manuscript: COPDGene visit 1, no exclusionary
disease, ever-smokers, non-missing post-bronchodilator FEV~1~/FVC, present in
Bhatt et al.'s per-subject MD-COPD classification file, and with a
post-bronchodilator baseline ESI value.

```{r cohort}
read_any <- function(p, ...) {
  h <- readLines(p, n = 1, warn = FALSE)
  read.delim(p, sep = if (grepl("\t", h)) "\t" else ",",
             stringsAsFactors = FALSE, ...)
}
# Participant identifier column name differs between site copies of these
# files; config_paths.R names it, and a merge-suffixed ".x" variant is
# detected automatically.
bp <- function(d) {
  d$pid <- as.character(d[[intersect(c(ID_COL, paste0(ID_COL, ".x")),
                                     names(d))[1]]])
  d
}

phe <- bp(read_any(PHE_PATH, na.strings = c("", "NA")))
mdc <- bp(read_any(MDCOPD_PATH)); esi <- bp(read_any(ESI_PATH))
vs  <- bp(read_any(VS_PATH));     cod <- bp(read_any(COD_PATH))

# Cause of death: TORCH Basic underlying cause, group 1 = respiratory.
u <- suppressWarnings(as.integer(cod$Torch_Group_Basic))
cod$UCD_Resp <- as.integer(!is.na(u) & u == 1L)

v1 <- phe[phe$visitnum == 1, ]
b  <- v1[is.na(v1$ExclusionaryDisease), ]
b  <- b[trimws(b$cohort) != "Never smoked", ]
b  <- b[!is.na(b$FEV1_FVC_post), ]
b  <- merge(b, mdc[!is.na(mdc$MultiDim_COPD_score),
                   c("pid", "MultiDim_COPD_score")], by = "pid")
# Post-bronchodilator ESI at visit 1, one value per participant.
ev <- esi %>% filter(visitnum == 1, PrePost == 1) %>%
  group_by(pid) %>% summarise(ESI = mean(ESI, na.rm = TRUE), .groups = "drop")
d <- merge(b, ev, by = "pid"); d <- d[!is.na(d$ESI), ]

# The three symptom minor criteria. A missing item counts as not met, which is
# what the published framework does.
f <- function(x) { x[is.na(x)] <- FALSE; x }
d$om  <- f(d$MMRCDyspneaScor >= 2) + f(d$SGRQ_scoreTotal >= 25) +
         f(d$Chronic_Bronchitis == 1)
d$afl <- d$FEV1_FVC_post < 0.70

O <- c("noCOPD", "AFL-only", "COPD-minor", "COPD-major")
# The reference classification is taken from Bhatt et al.'s own file, not
# reconstructed: score 1 = COPD-major, 2 = COPD-minor, 0 = neither, and the
# last is split into AFL-only vs noCOPD by the major criterion.
d$ref <- factor(ifelse(d$MultiDim_COPD_score == 1, "COPD-major",
                ifelse(d$MultiDim_COPD_score == 2, "COPD-minor",
                ifelse(d$afl, "AFL-only", "noCOPD"))), levels = O)

# The group a CT-free rule has to recover: COPD-major in the reference with
# zero symptom criteria met, so their COPD-major status rested entirely on a
# CT finding.
d$ct_only <- d$ref == "COPD-major" & d$om == 0
N_COHORT <- nrow(d); N_CTONLY <- sum(d$ct_only)
N_PRESERVED <- sum(!d$afl)

kable(as.data.frame(table(Category = d$ref)) %>%
        rename(n = Freq) %>% mutate(`% of cohort` = sprintf("%.1f", 100 * n / N_COHORT)),
      caption = sprintf(
        "Reference MD-COPD categories in the analytic cohort (n = %s participants).",
        n_fmt(N_COHORT)))
```

The cohort is **`r n_fmt(N_COHORT)` participants**. Of these,
**`r n_fmt(N_CTONLY)`** are COPD-major in the reference *with no symptom
criterion met*: their COPD-major status rested entirely on a CT finding. That
group of `r n_fmt(N_CTONLY)` is the thing a CT-free replacement has to recover,
and it is the denominator behind every "recovered" count in this report.
`r n_fmt(N_PRESERVED)` participants have preserved spirometry
(FEV~1~/FVC >= 0.70).

## Section 1.3 -- The three yardsticks

Each candidate is judged on three things. They measure different objectives
and, as Section 3 shows, they do not agree.

```{r yardsticks, echo = FALSE}
kable(data.frame(
  Yardstick = c("Macro-averaged F1 (in-sample)",
                "Macro-averaged F1 (held-out)",
                "CT-only COPD-major recovered",
                "AFL-only hazard ratio",
                "Reach in preserved spirometry"),
  `What it measures` = c(
    "agreement with the reference labels, averaged over the four categories",
    "the same, with thresholds refitted inside every training fold",
    "how many of the participants who reached COPD-major only through CT are put back in COPD-major",
    "mortality in the rule's own AFL-only group, against the rule's own noCOPD group",
    "participants with FEV1/FVC >= 0.70 in whom the replacement criterion fires at all"),
  Unit = c("proportion, 0 to 1", "proportion, 0 to 1",
           sprintf("participants, out of %s", n_fmt(N_CTONLY)),
           "hazard ratio, adjusted", sprintf("participants, out of %s", n_fmt(N_PRESERVED))),
  `Zero point / null` = c(
    "1 = exact reproduction of the reference; the reference row is 1 by construction",
    "no fixed maximum; the reference cannot be scored, having no free parameters",
    "833 = full recovery, 0 = none",
    "1 = the AFL-only label is true (that group is no worse off than noCOPD)",
    "0 = the criterion can never act outside airflow limitation"),
  check.names = FALSE),
  caption = "The yardsticks, with units and null values.")
```

**The distinction the rest of this report turns on.** Macro-F1 is symmetric
across the four categories: it rewards putting the right number of people in
each, and it penalizes over-calling as much as under-calling. Recovery of the
CT-only group is asymmetric: it counts only one direction of one category. A
rule can score well on the first by getting noCOPD, AFL-only and COPD-minor
right while doing a mediocre job on the specific people CT was needed for.

# Section 2 -- The candidates, and how each was fitted

Six candidates, plus the reference. Every one is given the same freedom: its
own threshold is searched over its own grid, jointly with the same minor-count
threshold *k* in 1 to 3, and the winner is the combination maximizing
macro-averaged F1 against the reference labels. **No candidate is handed a
threshold from the manuscript or from anywhere else** -- otherwise the
comparison would be between a fitted rule and un-fitted ones.

```{r candidates-fit}
# The classifier: a candidate contributes a 0/1 score, added to the symptom
# count, and the four categories follow the MD-COPD rules.
mk <- function(dd, s, k) {
  tot <- s + dd$om
  factor(ifelse(dd$afl & tot >  0, "COPD-major",
         ifelse(dd$afl & tot == 0, "AFL-only",
         ifelse(!dd$afl & tot >= k, "COPD-minor", "noCOPD"))), levels = O)
}
# Macro-averaged F1: the unweighted mean of the four per-category F1 values.
mF1 <- function(ref, est) mean(vapply(O, function(g) {
  tp <- sum(ref == g & est == g); if (tp == 0) return(0)
  2 * tp / (sum(est == g) + sum(ref == g)) }, numeric(1)))

CAND <- list(
  none       = list(var = NULL,                        grid = NA),
  `FEV1/FVC` = list(var = "FEV1_FVC_post", dir = "le", grid = seq(0.50, 0.69, by = 0.01)),
  FEV1pp     = list(var = "FEV1pp_utah",   dir = "le", grid = seq(30, 100, by = 2.5)),
  ESI        = list(var = "ESI",           dir = "ge", grid = seq(0.50, 4.00, by = 0.25)),
  `LAA-950`  = list(var = "Insp_LAA950_total_Thirona", dir = "ge",
                    grid = seq(1, 20, by = 0.5)))

score_of <- function(dd, cn, t) {
  cd <- CAND[[cn]]
  if (is.null(cd$var)) return(rep(0L, nrow(dd)))
  v <- dd[[cd$var]]
  # A missing value cannot meet the criterion, in either direction.
  v[is.na(v)] <- if (cd$dir == "ge") -Inf else Inf
  as.integer(if (cd$dir == "ge") v >= t else v <= t)
}
fit_one <- function(dd, cn) {
  g <- if (is.null(CAND[[cn]]$var)) NA else CAND[[cn]]$grid
  best <- list(f1 = -Inf)
  for (k in 1:3) for (t in g) {
    v <- mF1(dd$ref, mk(dd, score_of(dd, cn, t), k))
    if (v > best$f1) best <- list(f1 = v, k = k, t = t)
  }
  best
}
# The combined rule gets a two-dimensional search, so it is not handicapped
# relative to the single-criterion rules.
fit_both <- function(dd) {
  best <- list(f1 = -Inf)
  for (k in 1:3) for (tf in seq(0.50, 0.69, by = 0.01))
                  for (te in seq(0.50, 4.00, by = 0.25)) {
    s <- as.integer(dd$FEV1_FVC_post <= tf) + as.integer(dd$ESI >= te)
    v <- mF1(dd$ref, mk(dd, s, k))
    if (v > best$f1) best <- list(f1 = v, k = k, tf = tf, te = te)
  }
  best
}
lab_both <- function(dd, p) mk(dd, as.integer(dd$FEV1_FVC_post <= p$tf) +
                                   as.integer(dd$ESI >= p$te), p$k)

FITS <- lapply(names(CAND), function(cn) fit_one(d, cn)); names(FITS) <- names(CAND)
FB   <- fit_both(d)
LAB  <- lapply(names(CAND), function(cn)
          mk(d, score_of(d, cn, FITS[[cn]]$t), FITS[[cn]]$k))
names(LAB) <- names(CAND)
LAB[["FEV1/FVC+ESI"]] <- lab_both(d, FB)
```

```{r candidate-table, echo = FALSE}
kable(data.frame(
  Candidate = c("none", "FEV1/FVC", "FEV1pp", "ESI", "LAA-950", "FEV1/FVC+ESI"),
  `What it is` = c(
    "no replacement -- the NoCT-MD-COPD classification as published in the manuscript",
    "a second, lower threshold on the same variable that defines the major criterion",
    "FEV1 percent predicted, the conventional severity measure",
    "the Emphysema Severity Index, the manuscript's choice",
    "quantitative CT emphysema (Thirona, inspiratory, % voxels < -950 HU)",
    "both FEV1/FVC and ESI, each with its own fitted threshold"),
  `CT-free?` = c("yes", "yes", "yes", "yes", "**no**", "yes"),
  `Grid searched` = c("--", "0.50 to 0.69 by 0.01", "30 to 100 by 2.5",
                      "0.50 to 4.00 by 0.25", "1 to 20 by 0.5",
                      "the two grids jointly"),
  check.names = FALSE),
  caption = "The candidates. LAA-950 is not a CT-free option and is not a proposal; it is included as the ceiling a genuine structural measurement reaches when fitted the same way, which is the only way to know whether a CT-free rule is doing well or badly.")
```

**Why held-out performance is reported alongside in-sample.** A rule with a
free threshold can fit the reference labels better simply by having a free
threshold. Held-out macro-F1 refits each candidate's threshold *and* its count
threshold inside every training fold and scores it on the held-out fold, so a
candidate gains nothing from flexibility it cannot generalize. `r CV_REPEATS`
repeats of `r CV_FOLDS`-fold cross-validation, stratified on the reference
categories.

```{r cv, cache = FALSE}
set.seed(CV_SEED)
held <- setNames(numeric(length(LAB)), names(LAB))
for (r in seq_len(CV_REPEATS)) {
  fold <- integer(nrow(d))
  for (lv in O) {
    i <- which(d$ref == lv); fold[i] <- sample(rep_len(seq_len(CV_FOLDS), length(i)))
  }
  for (kf in seq_len(CV_FOLDS)) {
    tr <- d[fold != kf, ]; te <- d[fold == kf, ]
    for (cn in names(CAND)) {
      p <- fit_one(tr, cn)
      held[cn] <- held[cn] + mF1(te$ref, mk(te, score_of(te, cn, p$t), p$k))
    }
    p <- fit_both(tr)
    held["FEV1/FVC+ESI"] <- held["FEV1/FVC+ESI"] + mF1(te$ref, lab_both(te, p))
  }
}
held <- held / (CV_REPEATS * CV_FOLDS)
```

# Section 3 -- Head-to-head, and where the two objectives part company

## Section 3.1 -- Outcome models and assembly

The AFL-only hazard ratios below are adjusted for age, sex, race, current
smoking, pack-years and BMI, and are taken **against each rule's own noCOPD
group** -- not against a common reference. That is the right comparison,
because the question is whether *that rule's* AFL-only label describes a group
that is genuinely no worse off than *that rule's* healthy group.

```{r assemble}
m <- d %>%
  inner_join(vs %>% dplyr::select(pid, vital_status, days_followed), by = "pid") %>%
  left_join(cod %>% dplyr::select(pid, UCD_Resp), by = "pid") %>%
  mutate(ev_resp = ifelse(vital_status == 1 & !is.na(UCD_Resp) & UCD_Resp == 1, 1, 0),
         py = days_followed / 365.25) %>%
  filter(complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))
COV <- "age_visit + factor(gender) + factor(race) + factor(SmokCigNow) + ATS_PackYears + BMI"

hr_afl <- function(lv, ev) {
  m$g <- factor(lv[match(m$pid, d$pid)], levels = O)
  if (sum(m$g == "AFL-only") < 5) return(rep(NA_real_, 3))
  s <- summary(coxph(as.formula(sprintf("Surv(py, %s) ~ g + %s", ev, COV)),
                     data = m))$conf.int
  if (!"gAFL-only" %in% rownames(s)) return(rep(NA_real_, 3))
  unname(s["gAFL-only", c(1, 3, 4)])
}
cidx <- function(lv) {
  m$g <- factor(lv[match(m$pid, d$pid)], levels = O)
  summary(coxph(as.formula(paste("Surv(py, vital_status) ~ g +", COV)),
                data = m))$concordance[1]
}

rows <- lapply(names(LAB), function(cn) {
  lv <- LAB[[cn]]
  a <- hr_afl(lv, "vital_status"); rr <- hr_afl(lv, "ev_resp")
  p <- if (cn == "FEV1/FVC+ESI") FB else FITS[[cn]]
  data.frame(
    criterion = cn,
    rule = if (cn == "none") "no replacement"
           else if (cn == "FEV1/FVC+ESI")
             sprintf("FEV1/FVC<=%.2f & ESI>=%.2f, k=%d", p$tf, p$te, p$k)
           else sprintf("%s %s %s, k=%d", cn,
                        ifelse(CAND[[cn]]$dir == "ge", ">=", "<="),
                        format(p$t), p$k),
    n_aflonly = sum(lv == "AFL-only"), n_minor = sum(lv == "COPD-minor"),
    n_major = sum(lv == "COPD-major"),
    ct_only_recovered = sum(d$ct_only & lv == "COPD-major"),
    macroF1_insample = mF1(d$ref, lv), macroF1_heldout = held[[cn]],
    afl_all_HR = a[1], afl_all_lo = a[2], afl_all_hi = a[3],
    afl_resp_HR = rr[1], afl_resp_lo = rr[2], afl_resp_hi = rr[3],
    c_index = cidx(lv),
    # Reach: participants with preserved spirometry in whom the replacement
    # criterion fires. For the combined rule only the ESI component can fire
    # outside airflow limitation, so that is the component scored.
    reach_preserved = if (cn == "none") 0L else
      sum(!d$afl & score_of(d, if (cn == "FEV1/FVC+ESI") "ESI" else cn,
          if (cn == "FEV1/FVC+ESI") FB$te else FITS[[cn]]$t) == 1L))
})
res <- do.call(rbind, rows)

ref_row <- data.frame(
  criterion = "MD-COPD", rule = "reference (CT available)",
  n_aflonly = sum(d$ref == "AFL-only"), n_minor = sum(d$ref == "COPD-minor"),
  n_major = sum(d$ref == "COPD-major"), ct_only_recovered = sum(d$ct_only),
  macroF1_insample = 1, macroF1_heldout = NA,
  afl_all_HR  = hr_afl(d$ref, "vital_status")[1],
  afl_all_lo  = hr_afl(d$ref, "vital_status")[2],
  afl_all_hi  = hr_afl(d$ref, "vital_status")[3],
  afl_resp_HR = hr_afl(d$ref, "ev_resp")[1],
  afl_resp_lo = hr_afl(d$ref, "ev_resp")[2],
  afl_resp_hi = hr_afl(d$ref, "ev_resp")[3],
  c_index = cidx(d$ref), reach_preserved = NA)
res <- rbind(ref_row, res)
# rbind of the per-candidate frames leaves non-default row names, which kable
# then renders as an unlabelled first column in every downstream table.
rownames(res) <- NULL

# Convenience accessor used by the prose below, so no number is typed.
R <- function(cn, fld) res[[fld]][res$criterion == cn]
```

## Section 3.2 -- The table

```{r headline-table, echo = FALSE}
kable(res %>%
        transmute(Replacement = criterion, `Fitted rule` = rule,
                  `AFL-only n` = n_aflonly, `COPD-minor n` = n_minor,
                  `COPD-major n` = n_major,
                  `CT-only recovered` = ifelse(is.na(ct_only_recovered), "--",
                                               n_fmt(ct_only_recovered)),
                  `macro-F1 in-sample` = f3(macroF1_insample),
                  `macro-F1 held-out` = ifelse(is.na(macroF1_heldout), "--",
                                               f3(macroF1_heldout))),
      caption = sprintf(paste("Every candidate, fitted by the same objective over the same",
                              "count grid. n = %s participants; the CT-only column is out of %s."),
                        n_fmt(N_COHORT), n_fmt(N_CTONLY)))
```

**The result, as a sentence.** The candidate with the best held-out macro-F1 is
**`r res$criterion[which.max(res$macroF1_heldout)]`**
(`r f3(max(res$macroF1_heldout, na.rm = TRUE))`), and the candidate that
recovers the most CT-only COPD-major participants is
**`r res$criterion[res$criterion != "MD-COPD"][which.max(res$ct_only_recovered[res$criterion != "MD-COPD"])]`**
(`r n_fmt(max(res$ct_only_recovered[res$criterion != "MD-COPD"]))` of
`r n_fmt(N_CTONLY)`). **They are not the same candidate.**

## Section 3.3 -- Figure 3.1

**The message of this figure: the two yardsticks rank the candidates
differently, so choosing between candidates requires choosing between the
yardsticks first.** If the two agreed, the points would fall on a rising line
and the choice of objective would not matter.

**Reading the axes.** The x axis counts participants and is asymmetric: it
credits only one direction of one category (CT-only participants put back into
COPD-major), with a hard maximum at `r n_fmt(N_CTONLY)`, which only the
reference reaches. The y axis is a proportion, symmetric across all four
categories, and penalizes over-calling as much as under-calling. Neither axis
has a meaningful zero in the plotted range, so both are shown on their
occupied range rather than from 0.

```{r fig-3-1, fig.height = 6.0}
# Subtitles are wrapped programmatically rather than by hand: a hand-broken
# line survives until the next font or width change and then runs off the
# canvas, which it did here on the first render.
wrap <- function(txt, width = 76) paste(strwrap(txt, width), collapse = "\n")

plot_dat <- res %>%
  filter(criterion != "MD-COPD") %>%
  mutate(
    family = case_when(
      criterion == "ESI"      ~ "ESI-based",
      criterion == "LAA-950"  ~ "CT-based",
      criterion == "none"     ~ "No replacement",
      TRUE                    ~ "Other spirometric"),
    # Strip the criterion name off the front of its own rule string; repeating
    # it doubles every label's width for no information. sub() is NOT
    # vectorized over `pattern` -- it silently uses only the first element --
    # so this has to go through mapply, not straight into mutate.
    short = unname(mapply(function(cr, rl) trimws(sub(cr, "", rl, fixed = TRUE)),
                          criterion, rule)),
    short = ifelse(criterion == "none", "no replacement", short),
    label = sprintf("%s\n%s", criterion, short))

# A panel that renders empty errors nowhere, so assert the data is there.
stopifnot(nrow(plot_dat) == 6, all(is.finite(plot_dat$macroF1_heldout)),
          all(is.finite(plot_dat$ct_only_recovered)))

# Semantic colors, consistent with the manuscript figures: ESI red, CT blue,
# everything else neutral grey.
FAM_PAL <- c("ESI-based"         = unname(PALETTE["ESI-based"]),
             "CT-based"          = unname(PALETTE["CT-based"]),
             "Other spirometric" = "#555555",
             "No replacement"    = "#AAAAAA")

fig3_1 <- ggplot(plot_dat, aes(ct_only_recovered, macroF1_heldout)) +
  geom_point(aes(color = family, shape = family), size = 3.4) +
  geom_text_repel(aes(label = label, color = family), size = BODY_FS_NATIVE / .pt,
                  lineheight = 0.95, box.padding = 1.0, point.padding = 0.5,
                  min.segment.length = 0, segment.size = 0.4, force = 6,
                  max.overlaps = Inf, seed = 1L, show.legend = FALSE,
                  # "none" sits in the corner with nowhere to be repelled to,
                  # so it is given a starting offset rather than left on top
                  # of its own marker.
                  nudge_x = ifelse(plot_dat$criterion == "none", 45, 0),
                  nudge_y = ifelse(plot_dat$criterion == "none", -0.0018, 0)) +
  scale_color_manual(values = FAM_PAL, name = NULL, breaks = names(FAM_PAL)) +
  scale_shape_manual(values = c("ESI-based" = 16, "CT-based" = 17,
                                "Other spirometric" = 15, "No replacement" = 4),
                     name = NULL, breaks = names(FAM_PAL)) +
  scale_x_continuous(expand = expansion(mult = 0.22)) +
  scale_y_continuous(expand = expansion(mult = 0.16)) +
  labs(
    x = sprintf("CT-only COPD-major participants recovered (of %s)", n_fmt(N_CTONLY)),
    y = "Held-out macro-averaged F1",
    subtitle = wrap(sprintf(paste(
      "n = %s participants. Each point is one replacement criterion, fitted by",
      "macro-F1 over the same grid. Up is better on the objective the paper",
      "optimizes; right is better on the objective the paper argues."),
      n_fmt(N_COHORT)))) +
  theme_esi()

ggsave(file.path(FIGDIR, "fig3_1_objectives_disagree.png"), fig3_1,
       width = NATIVE_W, height = 6.0, dpi = 300)
fig3_1
```

**Figure 3.1.** Held-out macro-averaged F1 against recovery of the CT-only
COPD-major group, one point per candidate replacement criterion, all fitted
identically on the same `r n_fmt(N_COHORT)` participants. The reference
MD-COPD classification is not plotted: it has no free parameters, so it cannot
be scored out of sample, and it recovers all `r n_fmt(N_CTONLY)` by
construction. LAA-950 requires CT and is not a CT-free proposal; it marks what
a genuine structural measurement achieves. The FEV~1~/FVC and FEV~1~/FVC+ESI
markers very nearly coincide, which is itself a result: see Section 4.

# Section 4 -- Point 1: macro-F1 and recovery are different objectives

**The data.** Figure 3.1 and the table in Section 3.2.

**The method.** Rank the candidates on each yardstick separately and compare
the rankings.

```{r rankings, echo = FALSE}
rk <- res %>% filter(criterion != "MD-COPD") %>%
  transmute(Replacement = criterion,
            `macro-F1 held-out` = f3(macroF1_heldout),
            `rank by macro-F1` = rank(-macroF1_heldout, ties.method = "min"),
            `CT-only recovered` = ct_only_recovered,
            `rank by recovery` = rank(-ct_only_recovered, ties.method = "min")) %>%
  arrange(`rank by macro-F1`)
kable(rk, caption = "The same six candidates, ranked twice.")
```

**The overall result.** The two rankings disagree, and they disagree on the
candidates that matter.

- A second FEV1/FVC threshold, **`r R("FEV1/FVC", "rule")`**, achieves held-out
  macro-F1 **`r f3(R("FEV1/FVC", "macroF1_heldout"))`** against ESI's
  **`r f3(R("ESI", "macroF1_heldout"))`** -- it beats ESI on the objective the
  manuscript fits by. On recovery it puts back
  **`r n_fmt(R("FEV1/FVC", "ct_only_recovered"))`** of the
  `r n_fmt(N_CTONLY)` CT-only COPD-major participants, against ESI's
  **`r n_fmt(R("ESI", "ct_only_recovered"))`** --
  `r n_fmt(R("ESI", "ct_only_recovered") - R("FEV1/FVC", "ct_only_recovered"))`
  fewer people.
- **LAA-950**, quantitative CT emphysema and therefore the actual structural
  measurement the visual criteria were standing in for, scores
  **`r f3(R("LAA-950", "macroF1_heldout"))`** on held-out macro-F1 --
  `r rk$"rank by macro-F1"[rk$Replacement == "LAA-950"]`th of six, ahead only
  of no replacement at all -- while recovering
  **`r n_fmt(R("LAA-950", "ct_only_recovered"))`**, more than any other
  candidate.

- **Adding ESI on top of the FEV~1~/FVC rule changes almost nothing on either
  yardstick.** The combined rule reaches held-out macro-F1
  **`r f3(R("FEV1/FVC+ESI", "macroF1_heldout"))`** against FEV~1~/FVC alone at
  **`r f3(R("FEV1/FVC", "macroF1_heldout"))`**, and recovers
  **`r n_fmt(R("FEV1/FVC+ESI", "ct_only_recovered"))`** against
  **`r n_fmt(R("FEV1/FVC", "ct_only_recovered"))`** -- a difference of
  `r n_fmt(R("FEV1/FVC+ESI", "ct_only_recovered") - R("FEV1/FVC", "ct_only_recovered"))`
  participants. In Figure 3.1 the two markers very nearly coincide, which is
  the point: given a fitted second FEV~1~/FVC threshold, ESI adds essentially
  nothing measurable on either axis.

**A specific example, to anchor it.** Take the LAA-950 row on its own. If
macro-F1 were the right lens, the honest reading of that row is that measuring
emphysema directly on the CT scan is a poor way to replace a CT criterion,
barely better than deleting the criterion outright. That conclusion is
absurd on its face, and the reason it comes out of the analysis is that
macro-F1 is answering a different question: LAA-950 buys its recovery by
reclassifying `r n_fmt(R("LAA-950", "n_minor"))` participants into COPD-minor,
more than any other candidate, and macro-F1 charges it for every one of those
that the reference did not also call COPD-minor.

**What follows for the manuscript.** The paper fits its replacement criterion
by macro-F1 and then argues for it on the grounds that it recovers the people
CT was needed for. Those are two different objectives, and on this evidence
optimizing the first does not optimize the second. This does not make the
paper's fitted rule wrong. It means the paper cannot present the fitted rule
as though the fitting procedure had selected for recovery, because it did not.
Section 7 records the decision this raises.

# Section 5 -- Point 2: every spirometric replacement repairs the AFL-only label

## Section 5.1 -- Why this category is load-bearing

AFL-only is the category the CT-free framework is most at risk of breaking.
In the reference framework it means "airflow limitation and nothing else",
a genuinely low-risk group; the manuscript's argument for MD-COPD over the
fixed ratio rests substantially on that category being real. Remove the CT
criteria with nothing in their place and participants who had a CT finding and
no symptoms fall into AFL-only, which stops being low risk. **The test is
whether the AFL-only group's mortality, adjusted and against that rule's own
noCOPD group, is compatible with 1.**

```{r fig-5-1, fig.height = 7.0}
fp <- bind_rows(
  res %>% transmute(criterion, rule, outcome = "All-cause mortality",
                    est = afl_all_HR,  lo = afl_all_lo,  hi = afl_all_hi,
                    n = n_aflonly),
  res %>% transmute(criterion, rule, outcome = "Respiratory mortality",
                    est = afl_resp_HR, lo = afl_resp_lo, hi = afl_resp_hi,
                    n = n_aflonly)) %>%
  mutate(
    lab = sprintf("%s (n = %s)", criterion, n_fmt(n)),
    lab = factor(lab, levels = rev(unique(lab[outcome == "All-cause mortality"]))),
    outcome = factor(outcome, c("All-cause mortality", "Respiratory mortality")),
    family = case_when(criterion == "MD-COPD" ~ "Reference",
                       criterion == "none"    ~ "No replacement",
                       criterion == "ESI"     ~ "ESI-based",
                       criterion == "LAA-950" ~ "CT-based",
                       TRUE                   ~ "Other spirometric"))
stopifnot(nrow(fp) == 14, all(is.finite(fp$est)))

FP_PAL <- c("Reference" = "#000000", FAM_PAL)

fig5_1 <- ggplot(fp, aes(est, lab, color = family)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0.22) +
  geom_point(size = 2.6) +
  facet_wrap(~ outcome, scales = "free_x") +
  scale_x_log10() +
  scale_color_manual(values = FP_PAL, name = NULL, breaks = names(FP_PAL)) +
  guides(color = guide_legend(nrow = 2)) +
  labs(x = wrap("Adjusted hazard ratio for AFL-only versus that rule's own noCOPD group (log scale)", 62),
       y = NULL,
       subtitle = wrap(sprintf(paste("n = %s participants. Adjusted for age, sex, race,",
                                     "current smoking, pack-years and BMI. Dashed line at 1:",
                                     "the AFL-only label is true when the interval covers it.",
                                     "Row labels give each rule's own AFL-only group size."),
                               n_fmt(N_COHORT)), 62)) +
  theme_esi() +
  # Anchor the subtitle and axis title to the whole plot, not to the panel:
  # with row labels this wide the panel-anchored default runs off the canvas.
  theme(plot.title.position = "plot")

ggsave(file.path(FIGDIR, "fig5_1_aflonly_label.png"), fig5_1,
       width = NATIVE_W, height = 7.0, dpi = 300)
fig5_1
```

**Figure 5.1.** Whether each rule's AFL-only label is true. **The message: the
only rule whose AFL-only group is clearly not low risk is the one with no
replacement criterion at all; every replacement, spirometric or not, brings
the interval back across 1.** Note that the panels have independent x scales,
because respiratory mortality spans a far wider range than all-cause.

```{r afl-table, echo = FALSE}
kable(res %>% transmute(
        Replacement = criterion, `Fitted rule` = rule, `AFL-only n` = n_aflonly,
        `All-cause HR (95% CI)`   = ci_fmt(afl_all_HR, afl_all_lo, afl_all_hi),
        `Respiratory HR (95% CI)` = ci_fmt(afl_resp_HR, afl_resp_lo, afl_resp_hi),
        `Covers 1?` = ifelse(afl_resp_lo <= 1 & afl_resp_hi >= 1, "yes", "**no**")),
      caption = "AFL-only risk under each rule, against that rule's own noCOPD group.")
```

**The overall result.** With no replacement, AFL-only carries an adjusted
respiratory mortality hazard ratio of
**`r ci_fmt(R("none","afl_resp_HR"), R("none","afl_resp_lo"), R("none","afl_resp_hi"))`**:
the label is false, and by a wide margin. Every replacement repairs it:
FEV~1~/FVC **`r f2(R("FEV1/FVC","afl_all_HR"))`**, ESI
**`r f2(R("ESI","afl_all_HR"))`**, FEV~1~pp **`r f2(R("FEV1pp","afl_all_HR"))`**
on all-cause mortality, each with a respiratory interval covering 1, against
the reference framework's own **`r f2(R("MD-COPD","afl_all_HR"))`**.

**What follows for the manuscript. This is the point to state without
softening it.** The manuscript can claim that removing the CT criteria breaks
the AFL-only label and that a spirometric replacement repairs it; that claim
is robust, and it is supported by every candidate tested. **The manuscript
cannot claim that ESI is uniquely required to repair it.** Three different
spirometric criteria repair it about equally well, and nothing in these data
distinguishes ESI from the others on this yardstick. Any sentence in the paper
that reads as "ESI is needed to keep AFL-only low risk" should read "a
spirometric minor criterion is needed", with ESI's case made on the grounds in
Section 6 instead.

# Section 6 -- Point 3: reach into preserved spirometry, and the FEV1/FVC circularity

## Section 6.1 -- Why a second FEV1/FVC threshold cannot reach preserved spirometry

The natural objection to ESI is: why not use FEV~1~/FVC itself as the
replacement minor criterion, rather than importing a new measure? There is a
structural answer before there is an empirical one. FEV~1~/FVC < 0.70 is
already the **major** criterion, so a *minor* criterion built on a threshold at
or above 0.70 fires for every participant who has airflow limitation. Everyone
in AFL-only then has at least one minor criterion, and AFL-only collapses to
exactly zero -- destroying the very category the framework exists to get right.

**The method.** Sweep the FEV~1~/FVC minor threshold across both regimes at the
fitted count threshold and count the resulting AFL-only category.

```{r fig-6-1, fig.height = 5.0}
sweep <- do.call(rbind, lapply(seq(0.50, 0.85, by = 0.01), function(t) {
  s <- as.integer(d$FEV1_FVC_post <= t)
  est <- mk(d, s, 2L)
  data.frame(threshold = t, aflonly = sum(est == "AFL-only"),
             macroF1 = mF1(d$ref, est),
             reach = sum(!d$afl & d$FEV1_FVC_post <= t))
}))
stopifnot(nrow(sweep) == 36, any(sweep$aflonly > 0))

# FEV1/FVC is recorded to two decimals, so the collapse is already complete one
# grid step below the major criterion. Both facts are computed, not asserted.
FF_GAP        <- sum(d$FEV1_FVC_post > 0.69 & d$FEV1_FVC_post < 0.70)
FF_MAX_USABLE <- max(sweep$threshold[sweep$aflonly > 0])
FF_REACH_MAX  <- sweep$reach[which.min(abs(sweep$threshold - FF_MAX_USABLE))]

fig6_1 <- ggplot(sweep, aes(threshold, aflonly)) +
  geom_vline(xintercept = 0.70, linetype = "dashed", color = "grey50") +
  geom_line(color = "#555555", linewidth = 0.9) +
  geom_point(color = "#555555", size = 1.8) +
  geom_hline(yintercept = sum(d$ref == "AFL-only"), linetype = "dotted",
             color = unname(PALETTE["CT-based"])) +
  annotate("text", x = 0.705, y = max(sweep$aflonly) * 0.92, hjust = 0,
           size = BODY_FS_NATIVE / .pt, color = "grey30",
           label = "at and above 0.70 the minor criterion\nfires for everyone with airflow\nlimitation: AFL-only collapses to 0") +
  annotate("text", x = 0.50, y = sum(d$ref == "AFL-only"), vjust = 1.6, hjust = 0,
           size = BODY_FS_NATIVE / .pt, color = unname(PALETTE["CT-based"]),
           label = sprintf("reference MD-COPD AFL-only, n = %s",
                           n_fmt(sum(d$ref == "AFL-only")))) +
  labs(x = "FEV1/FVC threshold for the replacement minor criterion",
       y = "Participants classified AFL-only",
       subtitle = wrap(sprintf(paste("n = %s participants, count threshold k = 2.",
                                     "Dashed line: the major criterion, FEV1/FVC = 0.70."),
                               n_fmt(N_COHORT)))) +
  theme_esi()

ggsave(file.path(FIGDIR, "fig6_1_fevfvc_circularity.png"), fig6_1,
       width = NATIVE_W, height = 5.0, dpi = 300)
fig6_1
```

**Figure 6.1.** The circularity. **The message: a second FEV~1~/FVC threshold
is confined to below 0.70, because at or above 0.70 it annihilates the
AFL-only category.** Below 0.70 it acts only within the airflow-limitation
group, and reaches nobody with preserved spirometry -- necessarily, because
the replacement criterion and the major criterion are thresholds on the same
variable.

Two details the figure shows that are worth stating. The curve reaches zero at
**`r f2(0.69)`**, one grid step before the major criterion, because
FEV~1~/FVC is recorded to two decimals in COPDGene and
**`r n_fmt(FF_GAP)`** participants fall strictly between 0.69 and 0.70. The
highest threshold that leaves any AFL-only group at all is therefore
**`r f2(FF_MAX_USABLE)`**, and its reach into preserved spirometry is
**`r n_fmt(FF_REACH_MAX)`**. And the curve crosses the reference AFL-only size
of `r n_fmt(sum(d$ref == "AFL-only"))` well below the fitted threshold: matching
the reference *count* and maximizing label agreement are not the same target
either, which is Section 4's point in miniature.

## Section 6.2 -- Reach, and the honest version of ESI's advantage

**Reach** is the number of participants with preserved spirometry
(FEV~1~/FVC >= 0.70, n = `r n_fmt(N_PRESERVED)`) in whom the replacement
criterion fires at all. It is a capability, not a reclassification count: with
*k* = 2 a participant with preserved spirometry needs a second minor criterion
as well to become COPD-minor. Reach of zero means the criterion can never act
outside airflow limitation, whatever the rest of the framework does.

```{r reach-table, echo = FALSE}
kable(res %>% filter(criterion != "MD-COPD") %>%
        transmute(Replacement = criterion, `Fitted rule` = rule,
                  `Reach in preserved spirometry` = n_fmt(reach_preserved),
                  `% of preserved spirometry` = sprintf("%.1f", 100 * reach_preserved / N_PRESERVED),
                  `CT-free?` = ifelse(criterion == "LAA-950", "**no**", "yes")),
      caption = sprintf("Reach, out of %s participants with preserved spirometry.",
                        n_fmt(N_PRESERVED)))
```

**The overall result, stated carefully.** A second FEV~1~/FVC threshold reaches
**`r n_fmt(R("FEV1/FVC","reach_preserved"))`** participants with preserved
spirometry -- necessarily, by Section 6.1. ESI reaches
**`r n_fmt(R("ESI","reach_preserved"))`**. Adding ESI on top of the FEV~1~/FVC
rule adds **`r n_fmt(R("FEV1/FVC+ESI","reach_preserved"))`**, because the
combined fit pushes ESI's threshold up to
`r f2(FB$te)` where few participants meet it.

**But the README framing of this point needs correcting, and the correction
matters.** ESI is not the only CT-free criterion with reach.
**FEV~1~ percent predicted reaches `r n_fmt(R("FEV1pp","reach_preserved"))`
participants with preserved spirometry, more than ESI's
`r n_fmt(R("ESI","reach_preserved"))`.** LAA-950 reaches
`r n_fmt(R("LAA-950","reach_preserved"))`, but requires CT and so is not
available. The correct statement is therefore narrower than "reach is where
ESI is distinctive":

> A second threshold on FEV~1~/FVC cannot serve as the replacement minor
> criterion, because it is structurally confined to the airflow-limitation
> group. ESI is not so confined. Neither is FEV~1~pp, which reaches a larger
> group still; ESI's case against FEV~1~pp has to be made on what it reaches
> (`r n_fmt(R("ESI","ct_only_recovered"))` CT-only COPD-major participants
> recovered against FEV~1~pp's `r n_fmt(R("FEV1pp","ct_only_recovered"))`),
> not on whether it reaches.

That last clause is where ESI's real advantage sits, and it is a recovery
argument, not a macro-F1 argument -- which is Section 4's point again from the
other direction.

# Section 7 -- What this licenses, what it does not, and the open question

```{r verdict, echo = FALSE}
kable(data.frame(
  Claim = c(
    "Removing the two CT criteria breaks the AFL-only label",
    "A spirometric minor criterion repairs the AFL-only label",
    "ESI is uniquely required to repair the AFL-only label",
    "ESI reproduces the reference labels better than any CT-free alternative",
    "A second FEV1/FVC threshold is not a viable replacement",
    "ESI is the only CT-free criterion that reaches preserved spirometry",
    "ESI recovers more CT-only COPD-major participants than any CT-free alternative"),
  Verdict = c("**supported**", "**supported**", "**not supported**",
              "**not supported**", "**supported**", "**not supported**",
              "**supported**"),
  Evidence = c(
    "AFL-only respiratory HR with no replacement, Section 5",
    "every candidate's AFL-only interval covers 1, Section 5",
    "three spirometric criteria repair it about equally, Section 5",
    "a second FEV1/FVC threshold has higher held-out macro-F1, Section 4",
    "the circularity at 0.70, Section 6.1",
    "FEV1pp reaches more, Section 6.2",
    "recovery column, Section 3.2, among CT-free candidates"),
  check.names = FALSE),
  caption = "What the comparison does and does not license the manuscript to claim.")
```

**The open question, which this analysis raises and does not settle.** Should
the replacement criterion be fitted by recovery of the CT-identified group
rather than by macro-averaged F1? Section 4 shows the two objectives select
differently, and that the objective the paper fits by is not the objective the
paper argues from. Changing it is a paper-level decision with consequences
beyond this comparison -- it would change the fitted threshold, the
cross-validation, and the framing of the whole Results section -- and it was
deliberately not made here. Three options, for the co-authors:

1. **Keep macro-F1 and align the prose to it.** Argue the fitted rule on label
   agreement, and present recovery as a descriptive consequence rather than as
   the justification.
2. **Refit by recovery** and re-run the comparison. This would likely move
   ESI's threshold and would need its own cross-validation, since recovery has
   no penalty for over-calling and will push thresholds toward firing often.
3. **Keep macro-F1 as the fitting objective and report both yardsticks
   explicitly**, including this comparison, so a reader can see that the two
   disagree and why the fitted rule was chosen anyway.

# Section 8 -- Self-check

Every value below is recomputed by this report and compared against the value
recorded in the committed `replacement_criteria_comparison.csv`, which the
original side-analysis script wrote. A FAIL means this report and that script
have diverged, and one of them must be corrected deliberately rather than
silently.

```{r self-check, echo = FALSE}
csv_path <- file.path(HERE, "replacement_criteria_comparison.csv")
selfcheck <- function(name, computed, expected, tol = 0.002) {
  # Both NA is agreement, not failure: the reference classification has no
  # free parameters, so it has no held-out score and no reach, and NA is the
  # correct recorded value. Without this branch abs(NA - NA) < tol is NA and
  # isTRUE() turns a correct row into a spurious FAIL.
  ok <- if (is.na(computed) && is.na(expected)) TRUE
        else if (is.na(computed) || is.na(expected)) FALSE
        else if (is.numeric(computed) && is.numeric(expected))
          isTRUE(abs(computed - expected) < tol) else identical(computed, expected)
  data.frame(check = name,
             computed = if (is.numeric(computed)) sprintf("%.4f", computed)
                        else as.character(computed),
             expected = if (is.numeric(expected)) sprintf("%.4f", expected)
                        else as.character(expected),
             status = ifelse(ok, "PASS", "FAIL"))
}

checks <- list()
if (file.exists(csv_path)) {
  ref_csv <- read.csv(csv_path, stringsAsFactors = FALSE, check.names = FALSE)
  E <- function(cn, fld) ref_csv[[fld]][ref_csv$criterion == cn]
  for (cn in res$criterion) {
    checks <- c(checks, list(
      selfcheck(sprintf("%s -- fitted rule string", cn),
                R(cn, "rule"), E(cn, "rule")),
      selfcheck(sprintf("%s -- CT-only recovered", cn),
                as.numeric(R(cn, "ct_only_recovered")),
                as.numeric(E(cn, "ct_only_recovered")), tol = 0.5),
      selfcheck(sprintf("%s -- AFL-only n", cn),
                as.numeric(R(cn, "n_aflonly")), as.numeric(E(cn, "n_aflonly")), tol = 0.5),
      selfcheck(sprintf("%s -- held-out macro-F1", cn),
                as.numeric(R(cn, "macroF1_heldout")),
                as.numeric(E(cn, "macroF1_heldout"))),
      selfcheck(sprintf("%s -- reach in preserved spirometry", cn),
                as.numeric(R(cn, "reach_preserved")),
                as.numeric(E(cn, "reach_preserved")), tol = 0.5)))
  }
} else {
  checks <- list(selfcheck("replacement_criteria_comparison.csv present",
                           "missing", "present"))
}

# Structural invariants that must hold whatever the numbers are.
checks <- c(checks, list(
  selfcheck("FEV1/FVC reach in preserved spirometry is exactly 0 (the circularity)",
            as.numeric(R("FEV1/FVC", "reach_preserved")), 0, tol = 0.5),
  selfcheck("reference recovers all CT-only COPD-major by construction",
            as.numeric(R("MD-COPD", "ct_only_recovered")), as.numeric(N_CTONLY), tol = 0.5),
  selfcheck("no replacement recovers none by construction",
            as.numeric(R("none", "ct_only_recovered")), 0, tol = 0.5),
  selfcheck("the two objectives select different candidates",
            identical(res$criterion[which.max(res$macroF1_heldout)],
                      res$criterion[res$criterion != "MD-COPD"][
                        which.max(res$ct_only_recovered[res$criterion != "MD-COPD"])]),
            FALSE)))

sc <- do.call(rbind, checks)
kable(sc, caption = "Self-check: this report recomputed, against the committed CSV and against structural invariants.")
cat(sprintf("\nSelf-check summary: %d PASS, %d FAIL (of %d).\n",
            sum(sc$status == "PASS"), sum(sc$status == "FAIL"), nrow(sc)))
```

`r if (exists("sc") && any(sc$status == "FAIL")) "**One or more self-checks FAILED. Do not circulate this report until the discrepancy is resolved.**" else "All self-checks pass."`

# Section 9 -- Provenance

```{r provenance, echo = FALSE}
kable(data.frame(
  Item = c("This report", "Recomputes", "Compared against", "Figures written to",
           "Cohort", "Data inputs", "Manuscript state at writing", "Status"),
  Value = c(
    file.path("ctfree/side_analyses", SELF),
    "side_analyses/replacement_criteria_comparison.R, side_analyses/fevfvc_as_criterion.R",
    "side_analyses/replacement_criteria_comparison.csv",
    "side_analyses/figures/ (fig3_1, fig5_1, fig6_1)",
    sprintf("%s COPDGene participants, visit 1", n_fmt(N_COHORT)),
    "PHE_PATH, MDCOPD_PATH, ESI_PATH, VS_PATH, COD_PATH (config_paths.R)",
    "manuscript draft v1",
    "SIDE ANALYSIS -- not part of the manuscript pipeline; no number here appears in the paper"),
  check.names = FALSE),
  caption = "Provenance.")
```

```{r session, echo = FALSE}
sessionInfo()
```
