---
title: "ESI in a CT-free multidimensional COPD framework: analysis report accompanying the manuscript"
author: "Peter J. Castaldi"
date: "2026-09-03"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

<!--
Single reproducible analysis report that produces every result, table, and
figure cited in the CT-free MD-COPD manuscript. Bhatt et al.'s JAMA 2025
multidimensional COPD (MD-COPD) framework requires chest CT; this manuscript
asks whether ESI, a spirometry-only measure, can substitute for the CT
criteria while preserving MD-COPD's advantage over the fixed ratio.

Four classifications are compared on the same COPDGene participants:
  1. Fixed ratio    - post-bronchodilator FEV1/FVC < 0.70 alone
  2. MD-COPD        - the published CT-based framework (the reference)
  3. NoCT-MD-COPD   - MD-COPD with the two CT criteria removed
  4. ESI-MD-COPD    - MD-COPD with the two CT criteria replaced by ESI

MD-COPD categories are taken from Bhatt et al.'s own per-subject
classification file (MDCOPD_PATH) rather than reconstructed heuristically.
An earlier reconstruction attempt agreed with them on classification LOGIC
for every one of 9,406 participants classified in both, but differed on
cohort membership by 23 people in identifiable ways (13 with missing symptom
data we kept and they excluded; 10 without a quantitative CT variable that
was never part of their published criteria). Using their file directly
reproduces all four published category counts exactly (n = 9,416: noCOPD
4,439, AFL-only 282, COPD-minor 811, COPD-major 3,884). NoCT-MD-COPD
and ESI-MD-COPD are ours, built from the individual criteria directly.
-->

# Overview

This report reproduces the numerical claims of the manuscript from six
COPDGene source files: ESI values, phenotype/CT, vital status, cause of
death, long-term follow-up exacerbations, and Bhatt et al.'s per-subject
MD-COPD classification. Each section writes its output (tables as CSV,
figures as PNG) into `OUT_DIR`. A closing self-check block echoes every
canonical value cited in the manuscript prose against the expected number
so drift can be detected on any re-run.

# PATHS -- supplied at run time

Copy `config_paths.R.example` to `config_paths.R`, fill in the six input
locations for the machine you are running on, and knit.

```{r paths, fig.keep = "none"}
# ---- Headless raster device --------------------------------------------
# Sets the device for the chunks that follow. It cannot suppress the one X11
# warning seen on a headless node: knitr probes the png device during knit()
# startup, in set_html_dev() -> dev_available("png"), before any chunk runs,
# so no code in this document is early enough. render_ctfree.R sets the
# option before calling knit() and does prevent it; this block keeps the
# chunk devices correct for anyone who knits the file directly.
local({
  .bt <- getOption("bitmapType")
  if (capabilities("cairo") &&
      (is.null(.bt) || !nzchar(.bt) || identical(.bt, "Xlib"))) {
    options(bitmapType = "cairo")
    knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo"))
  }
})

# Locate this .Rmd's own directory so the report runs from any working
# directory. knitr sets the working directory to the .Rmd's location during
# rendering, so getwd() is the primary signal; the --file= fallback covers a
# plain `Rscript -e knitr::knit(...)` invocation, and the last branch covers
# an interactive knit from the repository root.
.b <- grep("^--file=", commandArgs(trailingOnly = FALSE), value = TRUE)
HERE <- {
  if (file.exists("ESI_ctfree_analysis_2026.9.3.Rmd")) {
    normalizePath(".")
  } else if (length(.b) &&
             file.exists(file.path(dirname(normalizePath(sub("^--file=", "", .b[1]))),
                                   "ESI_ctfree_analysis_2026.9.3.Rmd"))) {
    dirname(normalizePath(sub("^--file=", "", .b[1])))
  } else if (file.exists("ctfree/ESI_ctfree_analysis_2026.9.3.Rmd")) {
    normalizePath("ctfree")
  } else {
    stop("cannot locate this report's directory; knit from the ctfree/ ",
         "directory or from the repository root")
  }
}
ROOT    <- dirname(HERE)
CONFIG  <- file.path(ROOT, "config_paths.R")
if (!file.exists(CONFIG))
  stop("cannot find ", CONFIG, "; copy config_paths.R.example to ",
       "config_paths.R and fill in the local file paths")
source(CONFIG)

# This paper writes to its own assets directory, ctfree/assets/, NOT to the
# repository-wide manuscript_assets/ that belongs to the v15 (parent-directory)
# ESI manuscript. Any OUT_DIR set by config_paths.R applies to that older
# analysis and is deliberately ignored here.
OUT_DIR <- file.path(HERE, "assets")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ---- Analytic constants --------------------------------------------------
CV_REPEATS      <- 5L
CV_FOLDS        <- 5L
CV_SEED         <- 20260901L
CRUDE_B         <- 1000L

# Minimum events in a cell for an interval to be reported. Below this the
# interval is driven by a handful of deaths and conveys precision the data do
# not carry. The point estimate is still reported, without an interval and
# flagged, alongside the raw event count. Applied to crude and adjusted alike.
MIN_EVENTS <- 10L
BHATT_MDCOPD_N  <- c(noCOPD = 4439L, `AFL-only` = 282L,
                     `COPD-minor` = 811L, `COPD-major` = 3884L)
```

# Setup

```{r setup, include = FALSE, fig.keep = "none"}
# Print warnings as they happen rather than buffering them to the end of the
# knit. R's default (warn = 0) defers every warning until the top-level call
# returns, so a device warning raised in chunk 12 appears after chunk 68 with
# nothing to attribute it to. With warn = 1 the warning lands immediately
# after the chunk marker that produced it.
options(warn = 1)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 150, fig.width = 7, fig.height = 5)

suppressPackageStartupMessages({
  library(dplyr); library(survival); library(MASS); library(knitr); library(digest)
})

read_any <- function(p, ...) {
  h <- readLines(p, n = 1, warn = FALSE)
  read.delim(p, sep = if (grepl("\t", h)) "\t" else ",",
             stringsAsFactors = FALSE, ...)
}
bind_pid <- function(df, label) {
  cand <- c(ID_COL, paste0(ID_COL, ".x"), paste0(ID_COL, ".y"))
  hit <- cand[cand %in% names(df)]
  if (!length(hit)) stop("no identifier column in ", label)
  df$pid <- as.character(df[[hit[1]]]); df
}
```

# Section 1 -- Cohort construction

Bhatt et al.'s published exclusion chain, with the counts they report:

* 10,305 enrolled after removing ILD and bronchiectasis
* less 107 never-smokers                      -> 10,198
* less 66 unacceptable spirometry             -> 10,132
* less 716 excluded from the diagnostic algorithm -> 9,416

The first three steps reproduce exactly by construction (ILD/bronchiectasis
flag, cohort != "Never smoked", `FEV1_FVC_post` non-missing). Their fourth
step (CT quality control + minor-criteria data completeness) and their
MD-COPD classification are NOT reconstructed here: two independent attempts
at a heuristic reconstruction each landed within a handful of participants
of their counts, and a subject-level merge against their own classification
file showed exactly where the residual came from. See the header comment on
`MDCOPD_PATH` in `config_paths.R.example` for the full story. Using their
file directly gives an exact match on all four category counts.

Only after MD-COPD is validated against the published counts do we impose
this analysis's own additional requirement: baseline ESI available.

```{r ingest-and-cohort}
esi_raw <- bind_pid(read_any(ESI_PATH), "ESI_PATH")
phe_raw <- bind_pid(read_any(PHE_PATH, na.strings = c("", "NA")), "PHE_PATH")
vs      <- bind_pid(read_any(VS_PATH),  "VS_PATH")
cod     <- bind_pid(read_any(COD_PATH), "COD_PATH")
ex_raw  <- bind_pid(read_any(EX_PATH),  "EX_PATH")
mdcopd  <- bind_pid(read_any(MDCOPD_PATH), "MDCOPD_PATH")

# ---- Cause of death: TORCH Basic underlying cause ------------------------
ucd <- suppressWarnings(as.integer(cod$Torch_Group_Basic))
cod$UCD_Resp <- as.integer(!is.na(ucd) & ucd == 1L)

# ---- Cohort exclusion chain ----------------------------------------------
v1 <- phe_raw[phe_raw$visitnum == 1, ]
flow <- list(data.frame(step = "visit 1 in phenotype file",
                        n = nrow(v1), bhatt_n = NA_integer_,
                        stringsAsFactors = FALSE))
keep <- function(label, sel, target = NA_integer_) {
  v1 <<- v1[sel, , drop = FALSE]
  flow[[length(flow) + 1]] <<- data.frame(
    step = label, n = nrow(v1), bhatt_n = target, stringsAsFactors = FALSE)
  invisible(NULL)
}
keep("less interstitial lung disease or bronchiectasis",
     is.na(v1$ExclusionaryDisease), 10305L)
keep("less never-smokers",
     trimws(v1$cohort) != "Never smoked", 10198L)
keep("less unacceptable spirometry",
     !is.na(v1$FEV1_FVC_post), 10132L)

# Cohort membership past this point comes from Bhatt et al.'s per-subject
# classification file, not from a reconstruction of their CT-quality and
# minor-criteria completeness rules.
mdcopd_valid <- mdcopd[!is.na(mdcopd$MultiDim_COPD_score),
                       c("pid", "MultiDim_COPD_score")]
v1 <- merge(v1, mdcopd_valid, by = "pid")
flow[[length(flow) + 1]] <- data.frame(
  step = "less no valid MD-COPD classification (Bhatt ground truth)",
  n = nrow(v1), bhatt_n = 9416L, stringsAsFactors = FALSE)

# ---- Individual criteria (used by NoCT-MD-COPD and ESI-MD-COPD only) -----
# MD-COPD major/minor status comes from the file above; these criteria are
# needed to build the two CT-free classifications this paper is comparing
# against MD-COPD.
d_b <- within(v1, {
  gender <- factor(gender)
  race   <- factor(race)
  SmokCigNow <- factor(SmokCigNow)
  major_criterion <- FEV1_FVC_post < 0.70
  # The Fleischner scale runs none, trace, mild, moderate, confluent,
  # advanced destructive, so "mild or greater" is a score of 2, not 1.
  emph_yn <- CT_Visual_Emph_Severity >= 2
  wall_yn <- CT_Visual_Wall_Thickening == 2
  dysp_yn <- MMRCDyspneaScor >= 2
  qol_yn  <- SGRQ_scoreTotal >= 25
  cb_yn   <- Chronic_Bronchitis == 1
})
for (v in c("emph_yn", "wall_yn", "dysp_yn", "qol_yn", "cb_yn"))
  d_b[[v]][is.na(d_b[[v]])] <- FALSE

# ---- MD-COPD classification, from Bhatt's file directly ------------------
# score 1 = COPD-major, score 2 = COPD-minor, score 0 = neither; the last
# is split into AFL-only vs. noCOPD by the major criterion.
ORD <- c("noCOPD", "AFL-only", "COPD-minor", "COPD-major")
d_b$bhatt_cls <- factor(
  ifelse(d_b$MultiDim_COPD_score == 1, "COPD-major",
  ifelse(d_b$MultiDim_COPD_score == 2, "COPD-minor",
  ifelse(d_b$major_criterion, "AFL-only", "noCOPD"))),
  levels = ORD)

# Validate against Bhatt et al. Table 1. Only the score == 0 split is
# computed here (from FEV1_FVC_post); a mismatch would be a real finding,
# not a reconstruction gap, and should fail loudly.
got <- as.integer(table(d_b$bhatt_cls)[names(BHATT_MDCOPD_N)])
val <- data.frame(category = names(BHATT_MDCOPD_N), ours = got,
                  bhatt = as.integer(BHATT_MDCOPD_N),
                  diff = got - as.integer(BHATT_MDCOPD_N),
                  stringsAsFactors = FALSE)
stopifnot(all(val$diff == 0L))
write.csv(val, file.path(OUT_DIR, "mdcopd_validation.csv"), row.names = FALSE)

# ---- Baseline ESI (ours, not theirs) -------------------------------------
# Post-bronchodilator ESI at visit 1, one value per participant.
esi_v1 <- esi_raw[esi_raw$visitnum == 1 & esi_raw$PrePost == 1, ]
esi_v1 <- aggregate(list(ESI_v1post = esi_v1$ESI), list(pid = esi_v1$pid),
                    function(x) mean(x, na.rm = TRUE))
d_b <- merge(d_b, esi_v1, by = "pid")
d_b <- d_b[!is.na(d_b$ESI_v1post), , drop = FALSE]
flow[[length(flow) + 1]] <- data.frame(
  step = "less no baseline ESI", n = nrow(d_b), bhatt_n = NA_integer_,
  stringsAsFactors = FALSE)

cohort_flow <- do.call(rbind, flow)
# Every step for which Bhatt et al. published a count must match exactly,
# or the chain has diverged from theirs and everything downstream is on a
# different sample than intended.
chk <- cohort_flow[!is.na(cohort_flow$bhatt_n), ]
stopifnot(all(chk$n == chk$bhatt_n))
write.csv(cohort_flow, file.path(OUT_DIR, "cohort_flow.csv"), row.names = FALSE)

N_COHORT <- nrow(d_b)
cat(sprintf("Analytic cohort (with baseline ESI): n = %d\n", N_COHORT))
kable(cohort_flow, caption = "Cohort construction. `bhatt_n` is the published count for that step.")
kable(val, caption = "MD-COPD validated exactly against Bhatt et al. Table 1.")
```

## Section 1.1 -- Baseline characteristics by GOLD stratum (Supplemental Table S1)

Described on this cohort rather than carried over from the earlier analysis.
The exclusion chain removes never-smokers, so there is no never-smoker
stratum here.

```{r baseline-by-stratum}
# GOLD stratum from the COPDGene staging variable: -1 is PRISm, 0 is GOLD 0,
# 1-4 are GOLD 1-4. No "Never" level: never-smokers are excluded upstream.
stopifnot(!any(trimws(d_b$cohort) == "Never smoked"))
d_b$stratum <- factor(
  ifelse(d_b$finalgold_visit == -1, "PRISm",
         paste0("GOLD", d_b$finalgold_visit)),
  levels = c("PRISm", paste0("GOLD", 0:4)))

msd <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
pct <- function(x) sprintf("%.0f%%", 100 * mean(x, na.rm = TRUE))

base_row <- function(dd, label) data.frame(
  stratum      = label,
  n            = nrow(dd),
  age          = msd(dd$age_visit),
  pct_female   = pct(dd$gender == 2),
  pct_current  = pct(dd$SmokCigNow == 1),
  pack_years   = msd(dd$ATS_PackYears),
  BMI          = msd(dd$BMI),
  FEV1_pp      = msd(dd$FEV1pp_utah),
  FEV1_FVC     = sprintf("%.2f (%.2f)", mean(dd$FEV1_FVC_post, na.rm = TRUE),
                                        sd(dd$FEV1_FVC_post, na.rm = TRUE)),
  ESI          = sprintf("%.2f (%.2f)", mean(dd$ESI_v1post, na.rm = TRUE),
                                        sd(dd$ESI_v1post, na.rm = TRUE)),
  LAA950       = msd(dd$Insp_LAA950_total_Thirona),
  pct_mMRC2p   = pct(dd$dysp_yn),
  pct_SGRQ25p  = pct(dd$qol_yn),
  pct_CB       = pct(dd$cb_yn),
  stringsAsFactors = FALSE)

baseline <- do.call(rbind, c(
  lapply(levels(d_b$stratum), function(lv) base_row(d_b[d_b$stratum == lv, ], lv)),
  list(base_row(d_b, "Overall"))))
# The strata must partition the cohort, or a row is being double counted.
stopifnot(sum(baseline$n[baseline$stratum != "Overall"]) == nrow(d_b))
write.csv(baseline, file.path(OUT_DIR, "supp_baseline.csv"), row.names = FALSE)
kable(baseline, caption = "Baseline characteristics by GOLD stratum.")
```

## Section 1.2 -- ESI against quantitative CT (Supplemental Table S2)

Pearson correlations of ESI with quantitative CT emphysema, overall and within
stratum. The earlier analysis's supplemental table was a correlation matrix
*among the CT measures*, which does not support the statement it was cited for;
this computes what the text actually claims.

```{r esi-ct-correlations}
rpair <- function(dd, v) {
  ok <- stats::complete.cases(dd$ESI_v1post, dd[[v]])
  if (sum(ok) < 3) return(c(n = sum(ok), r = NA_real_))
  c(n = sum(ok), r = cor(dd$ESI_v1post[ok], dd[[v]][ok]))
}
corr_row <- function(dd, label) {
  a <- rpair(dd, "Insp_LAA950_total_Thirona")
  b <- rpair(dd, "PRM_pct_emphysema_Thirona")
  data.frame(stratum = label,
             n_LAA = as.integer(a["n"]), r_LAA = round(unname(a["r"]), 3),
             n_PRM = as.integer(b["n"]), r_PRM = round(unname(b["r"]), 3),
             stringsAsFactors = FALSE)
}
esi_ct <- do.call(rbind, c(
  lapply(levels(d_b$stratum), function(lv) corr_row(d_b[d_b$stratum == lv, ], lv)),
  list(corr_row(d_b, "All strata"))))
write.csv(esi_ct, file.path(OUT_DIR, "supp_esi_ct.csv"), row.names = FALSE)
kable(esi_ct, caption = "ESI against quantitative CT emphysema, by stratum.")
```

# Section 2 -- The four classifications

Fixed ratio, MD-COPD, NoCT-MD-COPD, and ESI-MD-COPD, applied to the same
`r nrow(d_b)` participants. All four use the same major criterion
(post-bronchodilator FEV1/FVC < 0.70), so no participant moves between the
airflow-limitation pair (AFL-only and COPD-major) and the
preserved-spirometry pair (noCOPD and COPD-minor).

```{r classifications}
# Handy shorthand columns used throughout the rest of the report.
d_b$afl <- d_b$major_criterion
d_b$ESI <- d_b$ESI_v1post
d_b$om  <- d_b$dysp_yn + d_b$qol_yn + d_b$cb_yn
O <- c("noCOPD", "AFL-only", "COPD-minor", "COPD-major")

# Generic classifier: given membership rules for the four categories,
# produce the factor.
label <- function(df, minor, aflo) factor(
  ifelse(df$afl & !aflo,   "COPD-major",
  ifelse(df$afl &  aflo,   "AFL-only",
  ifelse(!df$afl & minor,  "COPD-minor", "noCOPD"))), levels = O)

# S1  Fixed ratio.
d_b$S1 <- factor(ifelse(d_b$afl, "COPD", "noCOPD"),
                 levels = c("noCOPD", "COPD"))

# S2  MD-COPD (from Bhatt's file, level names normalised).
d_b$S2 <- factor(as.character(d_b$bhatt_cls), levels = O)

# S3  NoCT-MD-COPD. Symptom criteria only; count threshold is fitted below.
s3 <- function(df, k) label(df, df$om >= k, df$om == 0)

# S4  ESI-MD-COPD. ESI is a single criterion, met at or above one fitted
# threshold, so it counts like any of the three symptom criteria. An earlier
# version scored ESI 0/1/2 using a second, upper threshold that let ESI alone
# satisfy the count. That threshold was fitted at 7.00 and reached by 2 of the
# 5,156 participants with preserved spirometry, so it bought essentially
# nothing for an extra fitted parameter and is dropped.
s4 <- function(df, k, tl) label(df, (as.integer(df$ESI >= tl) + df$om) >= k,
                                   (as.integer(df$ESI >= tl) + df$om) == 0)

# Retained only to score the v15 draft rule, which WAS a two-threshold rule,
# so that historical comparison stays faithful to what was actually drafted.
s4_two <- function(df, k, tl, th) {
  s <- ifelse(df$ESI >= th, 2, ifelse(df$ESI >= tl, 1, 0))
  label(df, (s + df$om) >= k, (s + df$om) == 0)
}
```

# Section 3 -- Fitting NoCT-MD-COPD and ESI-MD-COPD

Both need a count threshold (three-of-five doesn't apply once the criterion
count changes), and ESI-MD-COPD also needs the two ESI thresholds. Fitting
is outcome-blind: only the MD-COPD labels enter, never mortality or
exacerbations, so the risk profiles below are an independent check rather
than a restatement of the fit.

The objective is macro-averaged F1 across the four MD-COPD categories.
Cohen's kappa and mean per-category sensitivity were considered and rejected
because they fail in OPPOSITE directions on this problem: kappa lets a
schema over-diagnose COPD by assigning almost no one to AFL-only,
mean sensitivity encourages the reverse by ignoring false positives.
Macro-F1 penalises both.

```{r fitting}
catF1 <- function(ref, est, g) {
  tp <- sum(ref == g & est == g); if (tp == 0) return(0)
  2 * tp / (sum(est == g) + sum(ref == g))
}
macroF1 <- function(ref, est) mean(vapply(O, function(g) {
  tp <- sum(ref == g & est == g); if (tp == 0) return(0)
  2 * tp / (sum(est == g) + sum(ref == g))
}, numeric(1)))
mean_recall <- function(ref, est)
  mean(vapply(O, function(g) mean(est[ref == g] == g), numeric(1)))
kappa_full <- function(ref, est) {
  t <- table(ref, est); n <- sum(t)
  po <- sum(diag(t)) / n
  pe <- sum(rowSums(t) * colSums(t)) / n^2
  (po - pe) / (1 - pe)
}

# Search grids.
KS  <- 1:3
TL  <- seq(0.50, 4.00, by = 0.25)
G4  <- do.call(rbind, lapply(2:4, function(k)
  data.frame(k = k, t_low = TL)))

fit3 <- function(df) {
  r <- df$S2
  KS[which.max(vapply(KS, function(k) macroF1(r, s3(df, k)), numeric(1)))]
}
fit4 <- function(df) {
  r <- df$S2
  scores <- vapply(seq_len(nrow(G4)), function(j)
    macroF1(r, s4(df, G4$k[j], G4$t_low[j])), numeric(1))
  G4[which.max(scores), ]
}

k3   <- fit3(d_b)
p4   <- fit4(d_b)
d_b$S3 <- s3(d_b, k3)
d_b$S4 <- s4(d_b, p4$k, p4$t_low)

cat(sprintf("NoCT-MD-COPD fitted: COPD-minor if >= %d of 3 symptom criteria\n", k3))
cat(sprintf("ESI-MD-COPD  fitted: COPD-minor if >= %d of 4; ESI criterion met at >= %.2f\n",
            p4$k, p4$t_low))
```

## Section 3.1 -- What each candidate objective selects

Reported as a Results finding rather than a Methods choice, since kappa
and mean sensitivity fail in different directions and describing them as
failing the same way would be wrong.

```{r objective-selection}
obj_rows <- list()
for (nm in c("kappa", "mean_sensitivity", "macro_F1")) {
  scorer <- switch(nm,
    kappa            = function(e) kappa_full(d_b$S2, e),
    mean_sensitivity = function(e) mean_recall(d_b$S2, e),
    macro_F1         = function(e) macroF1(d_b$S2, e))
  v <- vapply(seq_len(nrow(G4)), function(j)
    scorer(s4(d_b, G4$k[j], G4$t_low[j])), numeric(1))
  b <- G4[which.max(v), ]
  est <- s4(d_b, b$k, b$t_low)
  obj_rows[[length(obj_rows) + 1]] <- data.frame(
    objective = nm, k = b$k, t_low = b$t_low, score = max(v),
    n_aflonly = sum(est == "AFL-only"),
    n_major   = sum(est == "COPD-major"),
    stringsAsFactors = FALSE)
}
objsel <- do.call(rbind, obj_rows)
objsel$ref_aflonly <- sum(d_b$S2 == "AFL-only")
objsel$ref_major   <- sum(d_b$S2 == "COPD-major")
# The two rejected objectives must err in opposite directions on
# AFL-only or the Results sentence describing them is wrong.
stopifnot(objsel$n_aflonly[objsel$objective == "kappa"]
            < objsel$ref_aflonly[1],
          objsel$n_aflonly[objsel$objective == "mean_sensitivity"]
            > objsel$ref_aflonly[1])
write.csv(objsel, file.path(OUT_DIR, "objective_selection.csv"), row.names = FALSE)
kable(objsel, caption = "What each candidate objective selects. Kappa and mean sensitivity fail in opposite directions.")
```

## Section 3.1b -- Parameter sweep and where ESI's advantage comes from

The fitting above reports only the selected rule. These two blocks record what
the alternatives would have given, which is what justifies the selection, and
decompose the difference between the two CT-free classifications by category.

```{r metric-sweep}
bal_acc <- function(ref, est) mean_recall(ref, est)   # balanced accuracy is
                                                     # macro-averaged recall
sweep_row <- function(label, schema, param, est) data.frame(
  schema = schema, parameter = param, label = label,
  n_noCOPD = sum(est == "noCOPD"), n_aflonly = sum(est == "AFL-only"),
  n_minor  = sum(est == "COPD-minor"), n_major = sum(est == "COPD-major"),
  macroF1 = macroF1(d_b$S2, est), bal_acc = bal_acc(d_b$S2, est),
  kappa = kappa_full(d_b$S2, est), stringsAsFactors = FALSE)

rows <- list(sweep_row("MD-COPD (reference)", "reference", NA_character_, d_b$S2))
for (k in 1:3)
  rows[[length(rows)+1]] <- sweep_row(sprintf(">= %d of 3 symptom criteria", k),
                                      "S3", as.character(k), s3(d_b, k))
for (t in c(0.75, 1.00, 1.25, 1.50, 1.75, 2.00, 2.50, 3.00, 4.00))
  rows[[length(rows)+1]] <- sweep_row(sprintf("ESI >= %.2f", t), "S4",
                                      sprintf("%.2f", t), s4(d_b, 2, t))
metric_sweep <- do.call(rbind, rows)
metric_sweep$selected <- with(metric_sweep,
  (schema == "S3" & parameter == as.character(k3)) |
  (schema == "S4" & parameter == sprintf("%.2f", p4$t_low)))
# Each metric's optimum must fall somewhere in the swept range, or the range is
# too narrow to justify the selection.
for (m in c("macroF1", "bal_acc", "kappa")) {
  sub <- metric_sweep[metric_sweep$schema == "S4", ]
  stopifnot(which.max(sub[[m]]) %in% seq_len(nrow(sub)))
}
write.csv(metric_sweep, file.path(OUT_DIR, "metric_sweep.csv"), row.names = FALSE)
kable(metric_sweep, digits = 3, caption = "Parameter sweep. Category sizes and each candidate metric.")
```

```{r f1-decomposition}
# Macro-F1 is the mean of four per-category F1 values, so the difference
# between the two CT-free classifications decomposes exactly by category.
f1_cat <- function(ref, est, g) {
  tp <- sum(ref == g & est == g)
  if (tp == 0) return(0)
  2 * tp / (sum(est == g) + sum(ref == g))
}
f1_by_cat <- do.call(rbind, lapply(O, function(g) data.frame(
  category = g,
  n_ref  = sum(d_b$S2 == g),
  n_noct = sum(d_b$S3 == g), n_esi = sum(d_b$S4 == g),
  f1_noct = f1_cat(d_b$S2, d_b$S3, g), f1_esi = f1_cat(d_b$S2, d_b$S4, g),
  stringsAsFactors = FALSE)))
f1_by_cat$gain <- f1_by_cat$f1_esi - f1_by_cat$f1_noct
# The per-category F1 values must average to the macro-F1 reported above, or
# the decomposition is not of the quantity that was optimized.
stopifnot(abs(mean(f1_by_cat$f1_esi)  - macroF1(d_b$S2, d_b$S4)) < 1e-12,
          abs(mean(f1_by_cat$f1_noct) - macroF1(d_b$S2, d_b$S3)) < 1e-12)
write.csv(f1_by_cat, file.path(OUT_DIR, "f1_by_category.csv"), row.names = FALSE)
kable(f1_by_cat, digits = 3, caption = "Per-category F1. The macro-averaged value is the mean of the four.")
```

## Section 3.2 -- Repeated stratified cross-validation

Five repeats of five folds, stratified on the MD-COPD categories, with the
thresholds refitted inside every training fold.

```{r cv-fit}
set.seed(CV_SEED)
cv_rows <- list()
for (rep_i in seq_len(CV_REPEATS)) {
  fold <- integer(nrow(d_b))
  for (lv in O) {
    ix <- which(d_b$S2 == lv)
    fold[ix] <- sample(rep_len(seq_len(CV_FOLDS), length(ix)))
  }
  for (kf in seq_len(CV_FOLDS)) {
    tr <- d_b[fold != kf, ]
    te <- d_b[fold == kf, ]
    rt <- te$S2
    a <- fit3(tr); b <- fit4(tr)
    cv_rows[[length(cv_rows) + 1]] <- data.frame(
      rep = rep_i, fold = kf,
      k3 = a,  f3 = macroF1(rt, s3(te, a)),
      k4 = b$k, tl = b$t_low,
      f4 = macroF1(rt, s4(te, b$k, b$t_low)),
      f_draft = macroF1(rt, s4_two(te, 3, 1.00, 2.50)),
      stringsAsFactors = FALSE)
    .e3 <- s3(te, a); .e4 <- s4(te, b$k, b$t_low)
    for (g in O) {
      cv_rows[[length(cv_rows)]][[paste0("f3_", g)]] <- catF1(rt, .e3, g)
      cv_rows[[length(cv_rows)]][[paste0("f4_", g)]] <- catF1(rt, .e4, g)
    }
  }
}
cv <- do.call(rbind, cv_rows)
write.csv(cv, file.path(OUT_DIR, "cv_schema_fits.csv"), row.names = FALSE)

dd <- cv$f4 - cv$f3
cat(sprintf("Held-out macro-F1: NoCT-MD-COPD %.4f | ESI-MD-COPD %.4f | v15 draft %.4f\n",
            mean(cv$f3), mean(cv$f4), mean(cv$f_draft)))
cat(sprintf("ESI-MD-COPD minus NoCT-MD-COPD: %+.4f (%+.4f to %+.4f across folds)\n",
            mean(dd), quantile(dd, .025), quantile(dd, .975)))

write.csv(data.frame(
  schema = c("S3", "S4", "S4_v15draft"),
  k      = c(k3, p4$k, 3),
  t_low  = c(NA, p4$t_low, 1.00),
  t_high = c(NA, NA, 2.50),   # S4 is single-threshold; only the v15 draft had an upper one
  macroF1_insample = c(macroF1(d_b$S2, s3(d_b, k3)),
                       macroF1(d_b$S2, s4(d_b, p4$k, p4$t_low)),
                       macroF1(d_b$S2, s4_two(d_b, 3, 1.00, 2.50))),
  macroF1_heldout  = c(mean(cv$f3), mean(cv$f4), mean(cv$f_draft)),
  stringsAsFactors = FALSE),
  file.path(OUT_DIR, "schema_fit.csv"), row.names = FALSE)
write.csv(data.frame(diff_mean = mean(dd),
                     diff_lo = unname(quantile(dd, .025)),
                     diff_hi = unname(quantile(dd, .975)),
                     n_folds = nrow(cv)),
          file.path(OUT_DIR, "schema_fit_cv_diff.csv"), row.names = FALSE)

# Is the difference larger than the fitting noise? The folds are not
# independent: within a repeat they partition the same participants, and
# across repeats the training sets overlap. A paired t-test on them treats
# that shared data as new information and is anticonservative, so the
# variance is inflated by the Nadeau-Bengio correction for the resampled
# t-test, 1/n + n_test/n_train, with n_test/n_train = 1/(k - 1) for k-fold.
.n <- length(dd); .k <- length(unique(cv$fold))
.corr <- 1 / .n + 1 / (.k - 1)
.se <- sqrt(.corr * var(dd))
.t  <- mean(dd) / .se
cv_test <- data.frame(
  n_folds = .n, k = .k, n_repeats = length(unique(cv$rep)),
  diff_mean = mean(dd),
  ci_lo = mean(dd) - qt(.975, .n - 1) * .se,
  ci_hi = mean(dd) + qt(.975, .n - 1) * .se,
  t_stat = .t, df = .n - 1, p_value = 2 * pt(-abs(.t), .n - 1),
  folds_favoring_S4 = sum(dd > 0))
# The correction must widen the interval, never narrow it; if it does not,
# the factor has been applied the wrong way round.
stopifnot(.corr > 1 / .n, .se > sqrt(var(dd) / .n), cv_test$p_value < 1)
write.csv(cv_test, file.path(OUT_DIR, "schema_fit_cv_test.csv"), row.names = FALSE)

# The same corrected resampled t-test, per diagnostic group. The Results quote
# per-group F1 gains, so each needs its own test rather than borrowing the
# macro-averaged one, which is their mean.
cat_rows <- list()
for (g in O) {
  dg <- cv[[paste0("f4_", g)]] - cv[[paste0("f3_", g)]]
  se <- sqrt(.corr * var(dg)); tg <- mean(dg) / se
  cat_rows[[length(cat_rows) + 1]] <- data.frame(
    category = g,
    f1_noct = mean(cv[[paste0("f3_", g)]]),
    f1_esi  = mean(cv[[paste0("f4_", g)]]),
    diff_mean = mean(dg),
    ci_lo = mean(dg) - qt(.975, .n - 1) * se,
    ci_hi = mean(dg) + qt(.975, .n - 1) * se,
    t_stat = tg, df = .n - 1, p_value = 2 * pt(-abs(tg), .n - 1),
    folds_favoring_S4 = sum(dg > 0), stringsAsFactors = FALSE)
}
cv_cat_test <- do.call(rbind, cat_rows)
stopifnot(nrow(cv_cat_test) == length(O), all(is.finite(cv_cat_test$p_value)))
write.csv(cv_cat_test, file.path(OUT_DIR, "cv_category_test.csv"), row.names = FALSE)
print(cv_cat_test[, c("category","f1_noct","f1_esi","diff_mean","p_value")])
cat(sprintf("Corrected resampled t-test: diff %+.4f (%.4f to %.4f), t = %.2f, P = %.2e; ESI higher in %d of %d folds\n",
            cv_test$diff_mean, cv_test$ci_lo, cv_test$ci_hi, cv_test$t_stat,
            cv_test$p_value, cv_test$folds_favoring_S4, .n))
writeLines(c(sprintf("n_cohort=%d", nrow(d_b)),
             sprintf("n_afl=%d",    sum(d_b$afl)),
             sprintf("n_noafl=%d",  sum(!d_b$afl))),
           file.path(OUT_DIR, "cohort.txt"))
```

# Section 4 -- Category counts and reclassification

Save per-subject schema labels so figures and tables downstream don't
re-derive them, and write the flow of reclassification against MD-COPD as
its own artifact.

```{r labels-and-flow}
saveRDS(d_b[, c("pid", "S1", "S2", "S3", "S4")],
        file.path(OUT_DIR, "schema_labels.rds"))

lab_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) {
  tb <- table(d_b[[s]])
  for (g in names(tb)) lab_rows[[length(lab_rows) + 1]] <- data.frame(
    schema = s, category = g, n = as.integer(tb[[g]]),
    stringsAsFactors = FALSE)
}
schema_labels <- do.call(rbind, lab_rows)
write.csv(schema_labels, file.path(OUT_DIR, "schema_labels.csv"), row.names = FALSE)
kable(schema_labels, caption = "Category counts under each classification.")

# Reclassification counts quoted directly in the Results, plus the share of
# COPD-major that qualifies only through CT findings (the reason a CT-free
# schema can lose that category at all).
reclass <- do.call(rbind, lapply(c("S3", "S4"), function(s) data.frame(
  schema           = s,
  major_to_aflonly = sum(d_b$S2 == "COPD-major" & d_b[[s]] == "AFL-only"),
  minor_to_nocopd  = sum(d_b$S2 == "COPD-minor" & d_b[[s]] == "noCOPD"),
  nocopd_to_minor  = sum(d_b$S2 == "noCOPD"     & d_b[[s]] == "COPD-minor"),
  aflonly_kept     = sum(d_b$S2 == "AFL-only"   & d_b[[s]] == "AFL-only"),
  concordant       = sum(d_b$S2 == d_b[[s]]),
  stringsAsFactors = FALSE)))
n_major   <- sum(d_b$S2 == "COPD-major")
ct_only   <- sum(d_b$S2 == "COPD-major" & d_b$om == 0)
reclass$n_major       <- n_major
reclass$ct_only_major <- ct_only
reclass$ct_only_pct   <- 100 * ct_only / n_major
write.csv(reclass, file.path(OUT_DIR, "reclassification.csv"), row.names = FALSE)

# Full cross-classification of each CT-free schema against MD-COPD, long
# form so the supplement table can be built without recomputing anything.
xtab <- do.call(rbind, lapply(c("S3", "S4"), function(s) {
  tb <- table(factor(d_b[[s]], levels = O), factor(d_b$S2, levels = O))
  do.call(rbind, lapply(O, function(rw) data.frame(
    schema = s, row_cat = rw, col_cat = O,
    n = as.integer(tb[rw, O]), stringsAsFactors = FALSE)))
}))
stopifnot(sum(xtab$n) == 2 * nrow(d_b))
write.csv(xtab, file.path(OUT_DIR, "crossclass.csv"), row.names = FALSE)

# Is one classification correct more often than the other? The two are applied
# to the same participants and scored against the same reference, so the
# comparison is paired and McNemar's test on the discordant pairs is the test
# for it. An unpaired comparison of two proportions would ignore that 83% of
# participants are classified identically by both and would be anticonservative.
.c3 <- d_b$S3 == d_b$S2; .c4 <- d_b$S4 == d_b$S2
.b <- sum(.c3 & !.c4); .c <- sum(!.c3 & .c4)
.mc <- mcnemar.test(matrix(c(sum(!.c3 & !.c4), .b, .c, sum(.c3 & .c4)), 2),
                    correct = FALSE)
acc_test <- data.frame(
  n = nrow(d_b),
  correct_noct = sum(.c3), correct_esi = sum(.c4),
  acc_noct = mean(.c3), acc_esi = mean(.c4),
  only_esi_correct = .c, only_noct_correct = .b,
  chisq = unname(.mc$statistic), df = unname(.mc$parameter),
  p_value = .mc$p.value,
  p_exact = binom.test(.c, .b + .c)$p.value)
# The discordant pairs are the whole of the test; if they are zero there is
# nothing to test and the p-value would be meaningless.
stopifnot(.b + .c > 0, acc_test$correct_esi > acc_test$correct_noct)
write.csv(acc_test, file.path(OUT_DIR, "accuracy_test.csv"), row.names = FALSE)
cat(sprintf("Accuracy: NoCT %.1f%% vs ESI %.1f%%; discordant %d vs %d; McNemar P = %.2e\n",
            100 * acc_test$acc_noct, 100 * acc_test$acc_esi,
            acc_test$only_esi_correct, acc_test$only_noct_correct, acc_test$p_value))
```

# Section 5 -- Outcome data

Attach mortality, cause of death, and long-term follow-up exacerbations,
requiring complete covariate data for the adjusted models.

```{r outcome-data}
COV <- "age_visit + gender + race + SmokCigNow + ATS_PackYears + BMI"

mort <- d_b %>%
  inner_join(vs  %>% dplyr::select(pid, vital_status, days_followed), by = "pid") %>%
  left_join (cod %>% dplyr::select(pid, UCD_Resp), by = "pid") %>%
  mutate(ev_resp = ifelse(vital_status == 1 & !is.na(UCD_Resp) & UCD_Resp == 1, 1, 0),
         py      = days_followed / 365.25) %>%
  filter(complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))

# Prior-year exacerbation frequency, per the source MD-COPD paper.
prior <- phe_raw %>%
  filter(visitnum == 1) %>%
  transmute(pid, prior_exac = suppressWarnings(as.numeric(Exacerbation_Frequency))) %>%
  distinct(pid, .keep_all = TRUE)

exa <- d_b %>%
  inner_join(ex_raw %>% dplyr::select(pid, Total_Exacerbations, Years_Followed), by = "pid") %>%
  left_join (prior, by = "pid") %>%
  filter(!is.na(Total_Exacerbations), Years_Followed > 0, !is.na(prior_exac),
         complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))

cat(sprintf("Mortality n = %d (%d deaths, %d respiratory)\n",
            nrow(mort), sum(mort$vital_status), sum(mort$ev_resp)))
cat(sprintf("Exacerbation n = %d\n", nrow(exa)))
```

# Section 6 -- MD-COPD vs. the fixed ratio (nested LR test)

Fixed-ratio COPD is exactly the AFL-only and COPD-major categories
of MD-COPD, so the fixed-ratio model is nested in the four-category model
and the two are compared by likelihood ratio test on 2 degrees of freedom
rather than by comparing summary measures.

```{r gate-fixedratio}
d_b$bhatt_grp <- d_b$bhatt_cls
mort$fixed_ratio <- factor(ifelse(mort$afl, "COPD", "noCOPD"),
                           levels = c("noCOPD", "COPD"))
mort$bhatt_grp   <- d_b$bhatt_cls[match(mort$pid, d_b$pid)]
exa$fixed_ratio  <- factor(ifelse(exa$afl,  "COPD", "noCOPD"),
                           levels = c("noCOPD", "COPD"))
exa$bhatt_grp    <- d_b$bhatt_cls[match(exa$pid,  d_b$pid)]
stopifnot(all((mort$fixed_ratio == "COPD") ==
              (mort$bhatt_grp %in% c("AFL-only", "COPD-major"))))

cox_pair <- function(ev, label) {
  f0 <- as.formula(sprintf("Surv(days_followed/365.25, %s) ~ fixed_ratio + %s", ev, COV))
  f1 <- as.formula(sprintf("Surv(days_followed/365.25, %s) ~ bhatt_grp   + %s", ev, COV))
  m0 <- coxph(f0, data = mort); m1 <- coxph(f1, data = mort)
  lr <- 2 * (m1$loglik[2] - m0$loglik[2])
  df <- length(coef(m1)) - length(coef(m0))
  c0 <- summary(m0)$concordance; c1 <- summary(m1)$concordance
  cat(sprintf("--- %s ---\n", label))
  cat(sprintf("  fixed ratio : C = %.4f (SE %.4f)\n", c0[1], c0[2]))
  cat(sprintf("  MD-COPD     : C = %.4f (SE %.4f)   gain %+.4f\n",
              c1[1], c1[2], c1[1] - c0[1]))
  cat(sprintf("  LR chi-square = %.1f on %d df, p = %s\n\n",
              lr, df, format.pval(pchisq(lr, df, lower.tail = FALSE),
                                  digits = 3, eps = 1e-16)))
  data.frame(outcome = label, c_fixedratio = c0[1], c_mdcopd = c1[1],
             lrt_chisq = lr, lrt_df = df,
             lrt_p = pchisq(lr, df, lower.tail = FALSE),
             stringsAsFactors = FALSE)
}
gate_rows <- list()
gate_rows[[1]] <- cox_pair("vital_status", "ALL-CAUSE MORTALITY")
gate_rows[[2]] <- cox_pair("ev_resp",      "RESPIRATORY MORTALITY")

# Exacerbations: negative binomial with the same nesting.
n0 <- glm.nb(as.formula(paste("Total_Exacerbations ~ fixed_ratio +", COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa)
n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ bhatt_grp +",  COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa)
lr <- 2 * (as.numeric(logLik(n1)) - as.numeric(logLik(n0)))
df <- length(coef(n1)) - length(coef(n0))
cat("--- EXACERBATIONS ---\n")
cat(sprintf("  fixed ratio : AIC %.1f\n  MD-COPD     : AIC %.1f   improvement %.1f\n",
            AIC(n0), AIC(n1), AIC(n0) - AIC(n1)))
cat(sprintf("  LR chi-square = %.1f on %d df, p = %s\n",
            lr, df, format.pval(pchisq(lr, df, lower.tail = FALSE),
                                digits = 3, eps = 1e-16)))
gate_rows[[3]] <- data.frame(outcome = "EXACERBATIONS",
  c_fixedratio = NA_real_, c_mdcopd = NA_real_,
  lrt_chisq = lr, lrt_df = df,
  lrt_p = pchisq(lr, df, lower.tail = FALSE),
  stringsAsFactors = FALSE)

write.csv(do.call(rbind, gate_rows),
          file.path(OUT_DIR, "gate_fixedratio.csv"), row.names = FALSE)
write.csv(data.frame(aic_fixedratio = AIC(n0), aic_mdcopd = AIC(n1)),
          file.path(OUT_DIR, "gate_exac_aic.csv"), row.names = FALSE)
```

# Section 7 -- Adjusted risk within each classification's own categories

Cox proportional hazards models for mortality and negative binomial models
with a follow-up-year offset for exacerbations. Each estimate is expressed
against its OWN classification's noCOPD category, so a CT-free
classification's AFL-only group is compared to its own noCOPD group
rather than to MD-COPD's.

```{r adjusted-risk}
NAMES <- c(S1 = "Fixed ratio",
           S2 = "MD-COPD (reference)",
           S3 = "NoCT-MD-COPD",
           S4 = "ESI-MD-COPD")

risk_rows <- list(); disc_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) {
  m1 <- coxph(as.formula(sprintf("Surv(py, vital_status) ~ %s + %s", s, COV)), data = mort)
  m2 <- coxph(as.formula(sprintf("Surv(py, ev_resp) ~ %s + %s",      s, COV)), data = mort)
  n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~", s, "+", COV,
         "+ prior_exac + offset(log(Years_Followed))")), data = exa)
  cm <- summary(m1)$conf.int; cr <- summary(m2)$conf.int; ce <- summary(n1)$coef
  for (g in levels(d_b[[s]])) {
    r <- paste0(s, g); ix <- mort[[s]] == g
    gv <- function(M, j) if (r %in% rownames(M)) M[r, j] else NA_real_
    risk_rows[[length(risk_rows) + 1]] <- data.frame(
      schema = s, category = g,
      n_mort = sum(ix), n_exac = sum(exa[[s]] == g),
      n_cohort = sum(d_b[[s]] == g), deaths = sum(mort$vital_status[ix]),
      resp_deaths = sum(mort$ev_resp[ix]),
      all_few_events  = sum(mort$vital_status[ix]) < MIN_EVENTS,
      resp_few_events = sum(mort$ev_resp[ix]) < MIN_EVENTS,
      all_HR  = gv(cm, 1), all_LCI  = gv(cm, 3), all_UCI  = gv(cm, 4),
      resp_HR = gv(cr, 1), resp_LCI = gv(cr, 3), resp_UCI = gv(cr, 4),
      exac_IRR = if (r %in% rownames(ce)) exp(ce[r, 1]) else NA_real_,
      exac_LCI = if (r %in% rownames(ce)) exp(ce[r, 1] - 1.96 * ce[r, 2]) else NA_real_,
      exac_UCI = if (r %in% rownames(ce)) exp(ce[r, 1] + 1.96 * ce[r, 2]) else NA_real_,
      stringsAsFactors = FALSE)
  }
  disc_rows[[length(disc_rows) + 1]] <- data.frame(
    schema = s,
    c_allcause = summary(m1)$concordance[1],
    c_resp     = summary(m2)$concordance[1],
    exac_AIC   = AIC(n1), stringsAsFactors = FALSE)
}
risk <- do.call(rbind, risk_rows)
# Mortality and exacerbation models are fitted on different participant
# sets. Reporting one n per category for all three outcomes would overstate
# exacerbation denominators.
stopifnot(sum(risk$n_exac[risk$schema == "S2"]) == nrow(exa),
          sum(risk$n_mort[risk$schema == "S2"]) == nrow(mort),
          sum(risk$n_cohort[risk$schema == "S2"]) == nrow(d_b))
write.csv(risk, file.path(OUT_DIR, "schema_risk.csv"), row.names = FALSE)
write.csv(do.call(rbind, disc_rows),
          file.path(OUT_DIR, "schema_discrimination.csv"), row.names = FALSE)
kable(risk[risk$schema == "S2", c("category","n_mort","all_HR","all_LCI","all_UCI",
                                   "resp_HR","resp_LCI","resp_UCI",
                                   "exac_IRR","exac_LCI","exac_UCI")],
      caption = "Adjusted risk under MD-COPD.", digits = 3)
```

# Section 8 -- Crude rate ratios beside the adjusted estimates

An adjusted estimate is what the classification contributes beyond the
covariates; the crude ratio is what participants in the category actually
experienced. Both belong in the results.

Interval convention: a resample in which a category contributes no events
is retained as a legitimate draw with a rate ratio of zero (dropping such
draws truncates the interval from below and can push a lower bound above
1 on the strength of a single event). Only resamples in which the
REFERENCE contributes no events leave the ratio undefined and are cut.

```{r crude-rr}
set.seed(CV_SEED)

# Interval method by outcome. Deaths are counts over person-time, so the
# exact conditional (binomial) interval for a ratio of Poisson rates
# applies and returns a strictly positive lower bound whenever the group
# has at least one event. The subject-resample percentile bootstrap used
# previously returned exactly 0 whenever any resample of a small group
# contained no deaths, which is an artifact of the resampling rather than a
# statement about the rate. Exacerbation counts are overdispersed and
# clustered within participant, so the Poisson assumption does not hold and
# those intervals stay on the bootstrap.
exact_rate_ratio <- function(e1, t1, e0, t0) {
  if (e1 == 0) return(c(NA_real_, NA_real_))   # ratio is 0; reported as such
  ci <- poisson.test(c(e1, e0), c(t1, t0))$conf.int
  c(ci[1], ci[2])
}

crude_ratio <- function(sch, ev, py, dat) {
  g   <- dat[[sch]]
  num <- rowsum(dat[[ev]], g, reorder = FALSE)
  den <- rowsum(dat[[py]], g, reorder = FALSE)
  r   <- setNames(as.vector(num) / as.vector(den), rownames(num))
  ref <- r[["noCOPD"]]
  if (!is.finite(ref) || ref <= 0)
    return(setNames(rep(NA_real_, length(r)), names(r)))
  r / ref
}

crude_rows <- list()
for (spec in list(list("all",  "vital_status",        "py",             mort),
                  list("resp", "ev_resp",             "py",             mort),
                  list("exac", "Total_Exacerbations", "Years_Followed", exa))) {
  key <- spec[[1]]; ev <- spec[[2]]; py <- spec[[3]]; dat <- spec[[4]]
  boot <- vector("list", CRUDE_B)
  for (b in seq_len(CRUDE_B)) {
    idx <- sample.int(nrow(dat), nrow(dat), replace = TRUE)
    dd  <- dat[idx, , drop = FALSE]
    boot[[b]] <- lapply(c("S1","S2","S3","S4"),
                        function(s) crude_ratio(s, ev, py, dd))
  }
  for (si in seq_along(c("S1","S2","S3","S4"))) {
    s   <- c("S1","S2","S3","S4")[si]
    est <- crude_ratio(s, ev, py, dat)
    for (g in names(est)) {
      v <- vapply(boot, function(bb) {
        z <- bb[[si]]; if (g %in% names(z)) z[[g]] else NA_real_
      }, numeric(1))
      v <- v[is.finite(v)]
      ix <- dat[[s]] == g; rx <- dat[[s]] == "noCOPD"
      if (key == "exac") {
        lo <- if (length(v) > 1) unname(quantile(v, .025)) else NA_real_
        hi <- if (length(v) > 1) unname(quantile(v, .975)) else NA_real_
      } else {
        b <- exact_rate_ratio(sum(dat[[ev]][ix]), sum(dat[[py]][ix]),
                              sum(dat[[ev]][rx]), sum(dat[[py]][rx]))
        lo <- b[1]; hi <- b[2]
      }
      crude_rows[[length(crude_rows) + 1]] <- data.frame(
        schema = s, category = g, outcome = key,
        events = sum(dat[[ev]][ix]),
        rr = unname(est[[g]]), lo = lo, hi = hi,
        method = if (key == "exac") "bootstrap" else "exact Poisson",
        few_events = sum(dat[[ev]][ix]) < MIN_EVENTS,
        B_eff = sum(v > 0), stringsAsFactors = FALSE)
    }
  }
}
crude <- do.call(rbind, crude_rows)
# The reference category's crude ratio is 1 by construction. Drift means
# the numerator and denominator were computed off different rows.
stopifnot(all(abs(crude$rr[crude$category == "noCOPD"] - 1) < 1e-12))
# A zero lower bound now means the group genuinely had no events, not that a
# resample happened to miss them.
stopifnot(all(is.na(crude$lo[crude$method == "exact Poisson"]) |
              crude$lo[crude$method == "exact Poisson"] > 0 |
              crude$events[crude$method == "exact Poisson"] == 0))
write.csv(crude, file.path(OUT_DIR, "schema_crude.csv"), row.names = FALSE)
```

## COPD versus no COPD under each classification

Each classification collapsed to two groups (Pete, 2026-09-11): noCOPD and
AFL-only are no COPD, COPD-minor and COPD-major are COPD. The fixed ratio is
already binary. Every model then carries the same two-level exposure and the
same covariates, so the C-index and AIC compare the classifications on equal
terms, which the four-group models do not. Each classification's COPD group is
compared with its own no-COPD group; crude ratios first, then adjusted.

```{r binary-discrimination}
to_bin <- function(v) factor(ifelse(as.character(v) %in% c("COPD", "COPD-minor", "COPD-major"),
                                    "COPD", "noCOPD"), levels = c("noCOPD", "COPD"))
set.seed(CV_SEED)
bin_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) {
  mb <- mort; mb$B <- to_bin(mort[[s]])
  eb <- exa;  eb$B <- to_bin(exa[[s]])
  stopifnot(!anyNA(mb$B), !anyNA(eb$B), nlevels(droplevels(mb$B)) == 2)
  m1 <- coxph(as.formula(paste("Surv(py, vital_status) ~ B +", COV)), data = mb)
  m2 <- coxph(as.formula(paste("Surv(py, ev_resp) ~ B +",      COV)), data = mb)
  n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ B +", COV,
         "+ prior_exac + offset(log(Years_Followed))")), data = eb)
  sm <- function(dat, col, g) sum(dat[[col]][dat$B == g])
  crude <- function(dat, ev, pt) (sm(dat, ev, "COPD") / sm(dat, pt, "COPD")) /
                                 (sm(dat, ev, "noCOPD") / sm(dat, pt, "noCOPD"))
  ci_a <- exact_rate_ratio(sm(mb, "vital_status", "COPD"), sm(mb, "py", "COPD"),
                           sm(mb, "vital_status", "noCOPD"), sm(mb, "py", "noCOPD"))
  ci_r <- exact_rate_ratio(sm(mb, "ev_resp", "COPD"), sm(mb, "py", "COPD"),
                           sm(mb, "ev_resp", "noCOPD"), sm(mb, "py", "noCOPD"))
  be <- replicate(CRUDE_B, crude(eb[sample.int(nrow(eb), nrow(eb), replace = TRUE), ],
                                 "Total_Exacerbations", "Years_Followed"))
  be <- be[is.finite(be)]
  c1 <- summary(m1)$concordance; c2 <- summary(m2)$concordance
  hr <- function(m) summary(m)$conf.int["BCOPD", c(1, 3, 4)]
  ce <- summary(n1)$coef["BCOPD", ]
  bin_rows[[length(bin_rows) + 1]] <- data.frame(
    schema = s, n_copd = sum(to_bin(d_b[[s]]) == "COPD"), n_nocopd = sum(to_bin(d_b[[s]]) == "noCOPD"),
    deaths_copd = sm(mb, "vital_status", "COPD"), deaths_nocopd = sm(mb, "vital_status", "noCOPD"),
    resp_copd = sm(mb, "ev_resp", "COPD"), resp_nocopd = sm(mb, "ev_resp", "noCOPD"),
    all_rr = crude(mb, "vital_status", "py"), all_rr_lo = ci_a[1], all_rr_hi = ci_a[2],
    resp_rr = crude(mb, "ev_resp", "py"), resp_rr_lo = ci_r[1], resp_rr_hi = ci_r[2],
    exac_rr = crude(eb, "Total_Exacerbations", "Years_Followed"),
    exac_rr_lo = unname(quantile(be, .025)), exac_rr_hi = unname(quantile(be, .975)),
    all_HR = hr(m1)[1], all_LCI = hr(m1)[2], all_UCI = hr(m1)[3],
    resp_HR = hr(m2)[1], resp_LCI = hr(m2)[2], resp_UCI = hr(m2)[3],
    exac_IRR = exp(ce[1]), exac_LCI = exp(ce[1] - 1.96 * ce[2]), exac_UCI = exp(ce[1] + 1.96 * ce[2]),
    c_all = c1[1], c_all_lo = c1[1] - 1.96 * c1[2], c_all_hi = c1[1] + 1.96 * c1[2],
    c_resp = c2[1], c_resp_lo = c2[1] - 1.96 * c2[2], c_resp_hi = c2[1] + 1.96 * c2[2],
    exac_AIC = AIC(n1), B_eff = length(be), stringsAsFactors = FALSE, row.names = NULL)
}
schema_binary <- do.call(rbind, bin_rows)
stopifnot(nrow(schema_binary) == 4,
          all(schema_binary$n_copd + schema_binary$n_nocopd == nrow(d_b)),
          all(is.finite(as.matrix(schema_binary[, -1]))),
          all(schema_binary$resp_nocopd >= MIN_EVENTS), all(schema_binary$resp_copd >= MIN_EVENTS))
write.csv(schema_binary, file.path(OUT_DIR, "schema_binary.csv"), row.names = FALSE)
print(schema_binary[, c("schema", "n_copd", "all_rr", "all_HR", "resp_HR", "exac_IRR",
                        "c_all", "c_resp", "exac_AIC")], digits = 4)
```

Unadjusted discrimination and fit: each classification is the only predictor
in the model, so these show how much each classification on its own tells
about each outcome. Two codings: COPD versus no COPD, and the full diagnostic
groups (the fixed ratio has two groups either way). For a given outcome every
model uses the same participants, so AIC is comparable across classifications.
Exacerbations are a count, so they get AIC only.

```{r raw-discrimination}
raw_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) for (coding in c("binary", "groups")) {
  mb <- mort; eb <- exa
  mb$X <- if (coding == "binary") to_bin(mort[[s]]) else droplevels(factor(mort[[s]]))
  eb$X <- if (coding == "binary") to_bin(exa[[s]])  else droplevels(factor(exa[[s]]))
  stopifnot(!anyNA(mb$X), !anyNA(eb$X), nlevels(mb$X) >= 2)
  m1 <- coxph(Surv(py, vital_status) ~ X, data = mb)
  m2 <- coxph(Surv(py, ev_resp) ~ X, data = mb)
  n1 <- glm.nb(Total_Exacerbations ~ X + offset(log(Years_Followed)), data = eb)
  c1 <- summary(m1)$concordance; c2 <- summary(m2)$concordance
  raw_rows[[length(raw_rows) + 1]] <- data.frame(
    schema = s, coding = coding, n_groups = nlevels(mb$X), n_mort = nrow(mb), n_exac = nrow(eb),
    c_all = unname(c1[1]), c_all_lo = unname(c1[1] - 1.96 * c1[2]), c_all_hi = unname(c1[1] + 1.96 * c1[2]),
    aic_all = AIC(m1),
    c_resp = unname(c2[1]), c_resp_lo = unname(c2[1] - 1.96 * c2[2]), c_resp_hi = unname(c2[1] + 1.96 * c2[2]),
    aic_resp = AIC(m2), aic_exac = AIC(n1), stringsAsFactors = FALSE)
}
raw_discrimination <- do.call(rbind, raw_rows)
stopifnot(nrow(raw_discrimination) == 8, length(unique(raw_discrimination$n_mort)) == 1,
          length(unique(raw_discrimination$n_exac)) == 1,
          all(raw_discrimination$c_all > 0.5), all(raw_discrimination$c_resp > 0.5))
write.csv(raw_discrimination, file.path(OUT_DIR, "raw_discrimination.csv"), row.names = FALSE)
print(raw_discrimination, digits = 4)
```

Does each classification's crude COPD versus no-COPD ratio differ from
MD-COPD's? The classifications label the same participants, so the comparison
is paired: resample participants, recompute every classification's crude ratio
on the same draw, and take the ratio of each classification's ratio to
MD-COPD's. Two-sided bootstrap P from the share of draws on the other side of 1.

```{r binary-vs-mdcopd}
SCH <- c("S1", "S2", "S3", "S4")
crude_bin <- function(dat, s, ev, pt) {
  b <- to_bin(dat[[s]])
  e <- rowsum(dat[[ev]], b)[, 1]; t <- rowsum(dat[[pt]], b)[, 1]
  unname((e["COPD"] / t["COPD"]) / (e["noCOPD"] / t["noCOPD"]))
}
OUTC <- list(all  = list(mort, "vital_status", "py", "all_rr"),
             resp = list(mort, "ev_resp", "py", "resp_rr"),
             exac = list(exa, "Total_Exacerbations", "Years_Followed", "exac_rr"))
set.seed(CV_SEED)
vs_rows <- list()
for (o in names(OUTC)) {
  ev <- OUTC[[o]][[2]]; pt <- OUTC[[o]][[3]]
  dat <- OUTC[[o]][[1]][, c(SCH, ev, pt)]
  point <- vapply(SCH, function(s) crude_bin(dat, s, ev, pt), numeric(1))
  # the point estimates must reproduce the COPD versus no COPD table
  stopifnot(isTRUE(all.equal(unname(point), schema_binary[[OUTC[[o]][[4]]]][match(SCH, schema_binary$schema)])))
  bs <- replicate(CRUDE_B, {
    dd <- dat[sample.int(nrow(dat), nrow(dat), replace = TRUE), ]
    vapply(SCH, function(s) crude_bin(dd, s, ev, pt), numeric(1)) })
  for (s in setdiff(SCH, "S2")) {
    d <- log(bs[s, ]) - log(bs["S2", ]); d <- d[is.finite(d)]
    vs_rows[[length(vs_rows) + 1]] <- data.frame(
      schema = s, outcome = o, rr = point[[s]], rr_mdcopd = point[["S2"]],
      ratio_of_ratios = point[[s]] / point[["S2"]],
      lo = exp(unname(quantile(d, .025))), hi = exp(unname(quantile(d, .975))),
      p_boot = min(1, 2 * min(mean(d <= 0), mean(d >= 0))), B_eff = length(d),
      stringsAsFactors = FALSE)
  }
}
binary_vs_mdcopd <- do.call(rbind, vs_rows)
stopifnot(nrow(binary_vs_mdcopd) == 9, all(binary_vs_mdcopd$B_eff >= 0.99 * CRUDE_B))
write.csv(binary_vs_mdcopd, file.path(OUT_DIR, "binary_vs_mdcopd.csv"), row.names = FALSE)
print(binary_vs_mdcopd, digits = 3)
```

# Section 8a -- A common reference group across the three classifications

Section 7 estimates every category against its own classification's noCOPD
group. Those references differ in composition (4,357, 3,853 and 3,815
participants), so an estimate under one classification is not on the same
scale as an estimate under another, and part of any difference between them
is the moving denominator.

This section fixes one reference for all three: the participants all three
multidimensional classifications agree are noCOPD. Because every member of
that group is noCOPD under each classification, it is disjoint from the
AFL-only, COPD-minor and COPD-major categories of all three, so each of those
categories can be estimated against it without overlap. Participants the three
classifications disagree about are not in the reference; they still appear in
whichever category their own classification assigns.

AFL-only is a reported row here rather than the baseline. That is what the
argument needs: the claim is that AFL-only should sit at the noCOPD level, and
making it the reference would assume what is to be shown.

```{r consensus-reference}
cons <- with(d_b, S2 == "noCOPD" & S3 == "noCOPD" & S4 == "noCOPD")

# The construction rests on the reference being noCOPD under every
# classification. If it is not, the contrasts below compare overlapping
# groups and the estimates are meaningless.
for (s in c("S2", "S3", "S4"))
  stopifnot(all(d_b[[s]][cons] == "noCOPD"))
stopifnot(sum(cons) > 0)

# The fixed ratio is not carried here. Its advantage over the multidimensional
# framework is established in the source report and is not what this study asks.
GRPS <- c("AFL-only", "COPD-minor", "COPD-major")
OTHER <- list(S2 = GRPS, S3 = GRPS, S4 = GRPS)
SCHEMAS <- names(OTHER)
d_b$cons <- cons
mort_c <- merge(mort, d_b[, c("pid", "cons")], by = "pid")
exa_c  <- merge(exa,  d_b[, c("pid", "cons")], by = "pid")

cons_rows <- list()
for (s in SCHEMAS) {
  oth <- OTHER[[s]]
  keep_m <- mort_c$cons | mort_c[[s]] %in% oth
  keep_e <- exa_c$cons  | exa_c[[s]]  %in% oth
  mm <- mort_c[keep_m, ]; ee <- exa_c[keep_e, ]
  mm$g <- factor(ifelse(mm$cons, "REF", as.character(mm[[s]])), c("REF", oth))
  ee$g <- factor(ifelse(ee$cons, "REF", as.character(ee[[s]])), c("REF", oth))
  # Reference and comparison groups must not share participants.
  stopifnot(!any(mm$cons & mm[[s]] %in% oth), !anyNA(mm$g), !anyNA(ee$g))

  m1 <- coxph(as.formula(paste("Surv(py, vital_status) ~ g +", COV)), data = mm)
  m2 <- coxph(as.formula(paste("Surv(py, ev_resp) ~ g +",      COV)), data = mm)
  n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ g +", COV,
         "+ prior_exac + offset(log(Years_Followed))")), data = ee)
  cm <- summary(m1)$conf.int; cr <- summary(m2)$conf.int; ce <- summary(n1)$coef

  for (g in oth) {
    r <- paste0("g", g)                       # exact name; "^g" also matches gender
    stopifnot(r %in% rownames(cm), r %in% rownames(ce))
    cons_rows[[length(cons_rows) + 1]] <- data.frame(
      schema = s, category = g,
      n_mort = sum(mm$g == g), n_exac = sum(ee$g == g),
      deaths = sum(mm$vital_status[mm$g == g]),
      resp_deaths = sum(mm$ev_resp[mm$g == g]),
      all_few_events  = sum(mm$vital_status[mm$g == g]) < MIN_EVENTS,
      resp_few_events = sum(mm$ev_resp[mm$g == g]) < MIN_EVENTS,
      all_HR = cm[r, 1], all_LCI = cm[r, 3], all_UCI = cm[r, 4],
      resp_HR = cr[r, 1], resp_LCI = cr[r, 3], resp_UCI = cr[r, 4],
      exac_IRR = exp(ce[r, 1]),
      exac_LCI = exp(ce[r, 1] - 1.96 * ce[r, 2]),
      exac_UCI = exp(ce[r, 1] + 1.96 * ce[r, 2]),
      stringsAsFactors = FALSE)
  }
}
cons_risk <- do.call(rbind, cons_rows)

# Crude rates and rate ratios against the same common reference, by the same
# subject-resample bootstrap used in Section 8. The reference rate is
# recomputed inside every resample, as there.
set.seed(CV_SEED)
cons_rate <- function(sch, ev, py, dat) {
  g <- ifelse(dat$cons, "REF", as.character(dat[[sch]]))
  k <- g == "REF" | g %in% OTHER[[sch]]
  num <- tapply(dat[[ev]][k], g[k], sum); den <- tapply(dat[[py]][k], g[k], sum)
  r <- num / den
  if (!("REF" %in% names(r)) || !is.finite(r[["REF"]]) || r[["REF"]] <= 0)
    return(setNames(rep(NA_real_, length(r)), names(r)))
  r / r[["REF"]]
}

crude_c_rows <- list()
for (spec in list(list("all",  "vital_status",        "py",             mort_c),
                  list("resp", "ev_resp",             "py",             mort_c),
                  list("exac", "Total_Exacerbations", "Years_Followed", exa_c))) {
  key <- spec[[1]]; ev <- spec[[2]]; py <- spec[[3]]; dat <- spec[[4]]
  boot <- vector("list", CRUDE_B)
  for (b in seq_len(CRUDE_B)) {
    dd <- dat[sample.int(nrow(dat), nrow(dat), replace = TRUE), , drop = FALSE]
    boot[[b]] <- lapply(SCHEMAS, function(s) cons_rate(s, ev, py, dd))
  }
  for (si in seq_along(SCHEMAS)) {
    s <- SCHEMAS[si]
    est <- cons_rate(s, ev, py, dat)
    for (g in OTHER[[s]]) {
      v <- vapply(boot, function(bb) {
        z <- bb[[si]]; if (g %in% names(z)) z[[g]] else NA_real_ }, numeric(1))
      v <- v[is.finite(v)]
      ix <- dat[[s]] == g & !dat$cons; rx <- dat$cons
      if (key == "exac") {
        lo <- if (length(v) > 1) unname(quantile(v, .025)) else NA_real_
        hi <- if (length(v) > 1) unname(quantile(v, .975)) else NA_real_
      } else {
        b <- exact_rate_ratio(sum(dat[[ev]][ix]), sum(dat[[py]][ix]),
                              sum(dat[[ev]][rx]), sum(dat[[py]][rx]))
        lo <- b[1]; hi <- b[2]
      }
      crude_c_rows[[length(crude_c_rows) + 1]] <- data.frame(
        schema = s, category = g, outcome = key,
        events = sum(dat[[ev]][ix]),
        rate_100py = 100 * sum(dat[[ev]][ix]) / sum(dat[[py]][ix]),
        rr = unname(est[[g]]), lo = lo, hi = hi,
        method = if (key == "exac") "bootstrap" else "exact Poisson",
        few_events = sum(dat[[ev]][ix]) < MIN_EVENTS,
        B_eff = sum(v > 0), stringsAsFactors = FALSE)
    }
  }
}
cons_crude <- do.call(rbind, crude_c_rows)
stopifnot(nrow(cons_crude) == 27, all(is.finite(cons_crude$rr)))
write.csv(cons_crude, file.path(OUT_DIR, "consensus_ref_crude.csv"), row.names = FALSE)

# Reference composition is identical for all three classifications by
# construction; record it once so the prose cannot quote a drifting n.
cons_ref <- data.frame(
  n_cohort    = sum(cons),
  n_mort      = sum(mort_c$cons),
  n_exac      = sum(exa_c$cons),
  deaths      = sum(mort_c$vital_status[mort_c$cons]),
  resp_deaths = sum(mort_c$ev_resp[mort_c$cons]),
  exac_events = sum(exa_c$Total_Exacerbations[exa_c$cons]),
  person_yrs  = sum(mort_c$py[mort_c$cons]),
  fu_yrs      = sum(exa_c$Years_Followed[exa_c$cons]))

stopifnot(nrow(cons_risk) == 9, all(is.finite(cons_risk$all_HR)))
write.csv(cons_risk, file.path(OUT_DIR, "consensus_ref_risk.csv"), row.names = FALSE)
write.csv(cons_ref,  file.path(OUT_DIR, "consensus_ref_group.csv"), row.names = FALSE)

cat(sprintf(paste("Common reference: %d participants, %d deaths (%d respiratory)",
                  "over %.0f person-years, %d exacerbations over %.0f years.\n"),
            cons_ref$n_cohort, cons_ref$deaths, cons_ref$resp_deaths,
            cons_ref$person_yrs, cons_ref$exac_events, cons_ref$fu_yrs))
if (cons_ref$resp_deaths < 10)
  cat(sprintf(paste("NOTE: %d respiratory deaths in the reference. Respiratory",
                    "estimates in this section are directional only.\n"),
              cons_ref$resp_deaths))
# The participants NoCT-MD-COPD moves out of COPD-major, estimated as their own
# group against the same reference. The Results assert that their risk sits
# between the MD-COPD AFL-only and COPD-major levels; this is what tests it.
# They are COPD-major under MD-COPD, so they are disjoint from the reference.
lost <- d_b$S2 == "COPD-major" & d_b$S3 == "AFL-only"
stopifnot(sum(lost) > 0, !any(lost & cons))
mort_c$lost <- lost[match(mort_c$pid, d_b$pid)]
exa_c$lost  <- lost[match(exa_c$pid,  d_b$pid)]

mm <- mort_c[mort_c$cons | mort_c$lost, ]
ee <- exa_c[exa_c$cons | exa_c$lost, ]
mm$g <- factor(ifelse(mm$cons, "REF", "reclassified"), c("REF", "reclassified"))
ee$g <- factor(ifelse(ee$cons, "REF", "reclassified"), c("REF", "reclassified"))
stopifnot(!any(mm$cons & mm$lost), !anyNA(mm$g), !anyNA(ee$g))

.m1 <- coxph(as.formula(paste("Surv(py, vital_status) ~ g +", COV)), data = mm)
.m2 <- coxph(as.formula(paste("Surv(py, ev_resp) ~ g +",      COV)), data = mm)
.n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ g +", COV,
        "+ prior_exac + offset(log(Years_Followed))")), data = ee)
.cm <- summary(.m1)$conf.int; .cr <- summary(.m2)$conf.int; .ce <- summary(.n1)$coef
.r <- "greclassified"
.crude <- function(ev, py, dat) {
  a <- sum(dat[[ev]][dat$lost]) / sum(dat[[py]][dat$lost])
  b <- sum(dat[[ev]][dat$cons]) / sum(dat[[py]][dat$cons]); a / b
}
.ex <- function(e1, t1, e0, t0)
  if (e1 == 0) c(NA_real_, NA_real_) else poisson.test(c(e1, e0), c(t1, t0))$conf.int[1:2]
.ci_all  <- .ex(sum(mm$vital_status[mm$lost]), sum(mm$py[mm$lost]),
                sum(mm$vital_status[mm$cons]), sum(mm$py[mm$cons]))
.ci_resp <- .ex(sum(mm$ev_resp[mm$lost]), sum(mm$py[mm$lost]),
                sum(mm$ev_resp[mm$cons]), sum(mm$py[mm$cons]))

cons_lost <- data.frame(
  n = sum(lost), n_mort = sum(mm$lost), n_exac = sum(ee$lost),
  deaths = sum(mm$vital_status[mm$lost]), resp_deaths = sum(mm$ev_resp[mm$lost]),
  crude_all = .crude("vital_status", "py", mm),
  crude_all_lo = .ci_all[1], crude_all_hi = .ci_all[2],
  crude_resp = .crude("ev_resp", "py", mm),
  crude_resp_lo = .ci_resp[1], crude_resp_hi = .ci_resp[2],
  crude_exac = .crude("Total_Exacerbations", "Years_Followed", ee),
  all_HR = .cm[.r, 1], all_LCI = .cm[.r, 3], all_UCI = .cm[.r, 4],
  resp_HR = .cr[.r, 1], resp_LCI = .cr[.r, 3], resp_UCI = .cr[.r, 4],
  exac_IRR = exp(.ce[.r, 1]),
  exac_LCI = exp(.ce[.r, 1] - 1.96 * .ce[.r, 2]),
  exac_UCI = exp(.ce[.r, 1] + 1.96 * .ce[.r, 2]),
  all_few_events  = sum(mm$vital_status[mm$lost]) < MIN_EVENTS,
  resp_few_events = sum(mm$ev_resp[mm$lost]) < MIN_EVENTS)
stopifnot(nrow(cons_lost) == 1, is.finite(cons_lost$all_HR))
write.csv(cons_lost, file.path(OUT_DIR, "consensus_ref_lost.csv"), row.names = FALSE)
cat(sprintf("Reclassified group (n=%d): all-cause crude %.2f, adjusted %.2f; respiratory %d deaths, adjusted %.2f\n",
            cons_lost$n, cons_lost$crude_all, cons_lost$all_HR,
            cons_lost$resp_deaths, cons_lost$resp_HR))

sup <- cons_risk[cons_risk$resp_few_events, c("schema","category","resp_deaths")]
if (nrow(sup)) {
  cat(sprintf("Cells below the %d-event floor; ratio reported, interval not:\n",
              MIN_EVENTS))
  print(sup, row.names = FALSE)
} else cat("No cell falls below the", MIN_EVENTS, "event floor.\n")

kable(merge(cons_crude[cons_crude$outcome == "all",
                       c("schema","category","rate_100py","rr","lo","hi")],
            cons_risk[, c("schema","category","deaths","all_HR")],
            by = c("schema","category")),
      caption = "All-cause risk against the common noCOPD reference.", digits = 2)
```

Do the three classifications differ in the risk they assign each diagnostic
group? They label the same participants, so the comparison is paired: resample
participants, recompute every classification's crude and adjusted estimate for
each group against the common reference on the same draw, and take each CT-free
classification's estimate over MD-COPD's. The resampling order is the one the
COPD-minor-only version used, so its results reproduce. A comparison involves
two groups, so it gets no interval or p-value when either has fewer than
MIN_EVENTS events.

```{r group-paired-bootstrap}
B_GRP <- 1000L   # matches every other bootstrap in the paper (Pete, 2026-09-13)
CACHE <- file.path(OUT_DIR, "_cache"); dir.create(CACHE, showWarnings = FALSE)
GPATH <- file.path(CACHE, "group_paired_bootstrap.rds")
MEAS <- c("crude_all", "crude_resp", "crude_exac", "adj_all", "adj_resp", "adj_exac")
KEYS <- as.vector(t(outer(GRPS, MEAS, paste, sep = "|")))

fit_groups <- function(mm0, ee0, s) {
  oth <- OTHER[[s]]
  mm <- mm0[mm0$cons | mm0[[s]] %in% oth, ]; ee <- ee0[ee0$cons | ee0[[s]] %in% oth, ]
  mm$g <- factor(ifelse(mm$cons, "REF", as.character(mm[[s]])), c("REF", oth))
  ee$g <- factor(ifelse(ee$cons, "REF", as.character(ee[[s]])), c("REF", oth))
  rate <- function(d, g, ev, py) {
    a <- d$g == g; b <- d$g == "REF"
    (sum(d[[ev]][a]) / sum(d[[py]][a])) / (sum(d[[ev]][b]) / sum(d[[py]][b]))
  }
  m1 <- coxph(as.formula(paste("Surv(py, vital_status) ~ g +", COV)), data = mm)
  m2 <- coxph(as.formula(paste("Surv(py, ev_resp) ~ g +",      COV)), data = mm)
  n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ g +", COV,
         "+ prior_exac + offset(log(Years_Followed))")), data = ee)
  out <- setNames(rep(NA_real_, length(KEYS)), KEYS)
  for (g in GRPS) {
    r <- paste0("g", g)
    out[paste(g, MEAS, sep = "|")] <- c(
      log(rate(mm, g, "vital_status", "py")), log(rate(mm, g, "ev_resp", "py")),
      log(rate(ee, g, "Total_Exacerbations", "Years_Followed")),
      unname(coef(m1)[r]), unname(coef(m2)[r]), unname(coef(n1)[r]))
  }
  out
}
pt <- vapply(SCHEMAS, function(s) fit_groups(mort_c, exa_c, s), numeric(length(KEYS)))
rownames(pt) <- KEYS
# the adjusted point estimates must reproduce the common-reference table
for (s in SCHEMAS) for (g in GRPS) {
  k <- cons_risk$schema == s & cons_risk$category == g
  stopifnot(isTRUE(all.equal(exp(pt[paste(g, "adj_all", sep = "|"), s]),  cons_risk$all_HR[k])),
            isTRUE(all.equal(exp(pt[paste(g, "adj_exac", sep = "|"), s]), cons_risk$exac_IRR[k])))
}
.want <- list(n_m = nrow(mort_c), n_e = nrow(exa_c), B = B_GRP, version = 1L,
              labels = substr(digest::digest(paste(d_b$S2, d_b$S3, d_b$S4, d_b$cons)), 1, 12))
.cached <- if (file.exists(GPATH)) readRDS(GPATH) else NULL
if (!is.null(.cached) && identical(.cached$fingerprint, .want)) {
  BA <- .cached$BA
} else {
  set.seed(CV_SEED)
  BA <- array(NA_real_, c(B_GRP, length(SCHEMAS), length(KEYS)),
              dimnames = list(NULL, SCHEMAS, KEYS))
  for (b in seq_len(B_GRP)) {
    mm0 <- mort_c[sample.int(nrow(mort_c), nrow(mort_c), replace = TRUE), ]
    ee0 <- exa_c [sample.int(nrow(exa_c),  nrow(exa_c),  replace = TRUE), ]
    for (s in SCHEMAS)
      BA[b, s, ] <- tryCatch(suppressWarnings(fit_groups(mm0, ee0, s)),
                             error = function(e) rep(NA_real_, length(KEYS)))
  }
  saveRDS(list(BA = BA, fingerprint = .want), GPATH)
}
n_ev <- function(s, g, o)
  as.numeric(cons_crude$events[cons_crude$schema == s & cons_crude$category == g &
                               cons_crude$outcome == o])
rows <- list()
for (s in c("S3", "S4")) for (g in GRPS) for (ty in c("crude", "adj")) for (o in c("all", "resp", "exac")) {
  key <- paste(g, paste0(ty, "_", o), sep = "|")
  e1 <- n_ev(s, g, o); e0 <- n_ev("S2", g, o)
  stopifnot(length(e1) == 1, length(e0) == 1)
  few <- min(e1, e0) < MIN_EVENTS
  v <- BA[, s, key] - BA[, "S2", key]; v <- v[is.finite(v)]
  rows[[length(rows) + 1]] <- data.frame(
    schema = s, category = g, type = if (ty == "crude") "crude" else "adjusted", outcome = o,
    events = e1, events_mdcopd = e0,
    est = exp(pt[key, s]), est_mdcopd = exp(pt[key, "S2"]),
    ratio_of_ratios = exp(pt[key, s] - pt[key, "S2"]),
    lo = if (few) NA_real_ else exp(unname(quantile(v, .025))),
    hi = if (few) NA_real_ else exp(unname(quantile(v, .975))),
    p_boot = if (few) NA_real_ else min(1, 2 * min(mean(v <= 0), mean(v >= 0))),
    B_eff = length(v), few_events = few, stringsAsFactors = FALSE)
}
group_vs_mdcopd <- do.call(rbind, rows)
stopifnot(nrow(group_vs_mdcopd) == 36,
          all(group_vs_mdcopd$B_eff[!group_vs_mdcopd$few_events] >= 0.95 * B_GRP))
write.csv(group_vs_mdcopd, file.path(OUT_DIR, "group_vs_mdcopd.csv"), row.names = FALSE)
print(group_vs_mdcopd[, c("schema", "category", "type", "outcome", "ratio_of_ratios",
                          "lo", "hi", "p_boot", "few_events")], digits = 3)
```

# Section 8b -- Longitudinal FEV~1~ decline by category

The fourth outcome the MD-COPD framework was validated on, and the one this
analysis had been omitting. Linear mixed models over visits 1 to 3 with a
random intercept per participant; the decline attributable to a category is
the interaction between follow-up time and that category, in mL per year
against the classification's own noCOPD group.

```{r fev1-decline}
suppressPackageStartupMessages({library(lme4); library(lmerTest)})

fev1_long <- phe_raw %>%
  filter(visitnum %in% c(1, 2, 3), !is.na(FEV1_post), !is.na(years_from_baseline)) %>%
  transmute(pid, visitnum, years_from_baseline,
            FEV1_post_mL  = FEV1_post * 1000,
            age_visit     = as.numeric(age_visit),
            ATS_PackYears = as.numeric(ATS_PackYears),
            SmokCigNow    = factor(SmokCigNow))

base_cols <- d_b %>% transmute(
  pid, S1, S2, S3, S4, cons,
  Height_CM       = as.numeric(Height_CM),
  gender_baseline = factor(gender),
  race_baseline   = factor(race))

# Baseline post-bronchodilator FEV1, for the adjusted model below. Taken from
# visit 1 of the longitudinal file rather than d_b so it is on the same scale
# and from the same source as the outcome.
base_fev1 <- fev1_long %>%
  filter(visitnum == 1) %>%
  transmute(pid, FEV1_base_mL = FEV1_post_mL) %>%
  distinct(pid, .keep_all = TRUE)

decline <- fev1_long %>%
  inner_join(base_cols, by = "pid") %>%
  inner_join(base_fev1, by = "pid") %>%
  filter(!is.na(Height_CM), !is.na(age_visit), !is.na(SmokCigNow))

cat(sprintf("FEV1 decline: %d observations on %d participants\n",
            nrow(decline), length(unique(decline$pid))))

lmm_row <- function(fit, term) {
  co <- summary(fit)$coefficients
  if (!term %in% rownames(co)) return(c(est = NA_real_, se = NA_real_, p = NA_real_))
  c(est = unname(co[term, "Estimate"]), se = unname(co[term, "Std. Error"]),
    p = unname(co[term, ncol(co)]))
}

# The source MD-COPD report adjusted its FEV1 model for baseline
# post-bronchodilator FEV1, and the groups differ sharply in baseline lung
# function, so the estimate asks whether a group declines faster than others
# starting from the same FEV1. Baseline FEV1 enters through its interaction
# with time only. A main effect would have visit 1 predicting itself, since the
# outcome at time zero IS the baseline value.
# Against the same common noCOPD reference as every other outcome. Each
# classification's own noCOPD group differs in composition, so estimates made
# against them are not comparable between classifications, which is the
# comparison this outcome is here to support.
dec_rows <- list()
for (sname in SCHEMAS) {
  dd <- decline[decline$cons | decline[[sname]] %in% OTHER[[sname]], ]
  dd$g <- factor(ifelse(dd$cons, "REF", as.character(dd[[sname]])),
                 c("REF", OTHER[[sname]]))
  stopifnot(!any(dd$cons & dd[[sname]] %in% OTHER[[sname]]), !anyNA(dd$g))
  f <- as.formula(paste0(
    "FEV1_post_mL ~ years_from_baseline * g",
    " + Height_CM + gender_baseline + race_baseline + age_visit + SmokCigNow",
    " + ATS_PackYears + years_from_baseline:FEV1_base_mL + (1 | pid)"))
  fit <- lmer(f, data = dd)
  for (g in OTHER[[sname]]) {
    r <- lmm_row(fit, paste0("years_from_baseline:g", g))
    dec_rows[[length(dec_rows) + 1]] <- data.frame(
      schema = sname, category = g,
      est_mL_yr = unname(r["est"]), se = unname(r["se"]), p = unname(r["p"]),
      lo = unname(r["est"] - 1.96 * r["se"]),
      hi = unname(r["est"] + 1.96 * r["se"]),
      n_subj = length(unique(dd$pid[dd$g == g])),
      stringsAsFactors = FALSE)
  }
}
fev1_decline <- do.call(rbind, dec_rows)
# Both models must cover every group, or a comparison between them is silently
# comparing different sets of rows.
stopifnot(nrow(fev1_decline) == 9, all(is.finite(fev1_decline$est_mL_yr)))
write.csv(fev1_decline, file.path(OUT_DIR, "fev1_decline.csv"), row.names = FALSE)
kable(fev1_decline, digits = 2,
      caption = "FEV1 decline (mL/yr) against the common noCOPD reference.")
```

Median follow-up for each outcome, over that outcome's analysis data:
person-years to death or censoring for mortality, years followed in the
longitudinal follow-up program for exacerbations, and years from visit 1 to the
last spirometry visit for FEV1 decline. FEV1 is given twice, over everyone in
the decline data and over those with at least one follow-up visit, because
participants seen only at visit 1 contribute zero follow-up.

```{r followup-medians}
fu_fev <- decline %>% group_by(pid) %>%
  summarise(last = max(years_from_baseline), n_vis = n_distinct(visitnum), .groups = "drop")
fu_row <- function(label, v) data.frame(
  outcome = label, n = length(v), median_yr = median(v),
  q1_yr = unname(quantile(v, .25)), q3_yr = unname(quantile(v, .75)),
  stringsAsFactors = FALSE)
followup_medians <- rbind(
  fu_row("mortality (all-cause and respiratory)", mort$py),
  fu_row("exacerbations", exa$Years_Followed),
  fu_row("FEV1 decline, everyone in the decline data", fu_fev$last),
  fu_row("FEV1 decline, participants with a follow-up visit", fu_fev$last[fu_fev$n_vis > 1]))
stopifnot(nrow(followup_medians) == 4, all(followup_medians$median_yr >= 0),
          followup_medians$n[1] == nrow(mort), followup_medians$n[2] == nrow(exa))
write.csv(followup_medians, file.path(OUT_DIR, "followup_medians.csv"), row.names = FALSE)
print(followup_medians, digits = 3)
```

Does FEV1 decline differ between the two CT-free classifications' groups? The
groups overlap (every ESI AFL-only participant is also NoCT AFL-only), so the
two estimates are not independent. Participants are resampled, both
classifications' models are refitted on the same draw, and the difference in
each group's slope is taken, NoCT minus ESI. Point estimates are the full-data
ones above; the first assertion checks that this chunk fits the same model.

```{r fev1-paired-bootstrap}
B_FEV <- 1000L
CACHE <- file.path(OUT_DIR, "_cache"); dir.create(CACHE, showWarnings = FALSE)
FPATH <- file.path(CACHE, "fev1_paired_bootstrap.rds")
fit_slopes <- function(dat, sname) {
  dd <- dat[dat$cons | dat[[sname]] %in% OTHER[[sname]], ]
  dd$g <- factor(ifelse(dd$cons, "REF", as.character(dd[[sname]])), c("REF", OTHER[[sname]]))
  f <- as.formula(paste0(
    "FEV1_post_mL ~ years_from_baseline * g",
    " + Height_CM + gender_baseline + race_baseline + age_visit + SmokCigNow",
    " + ATS_PackYears + years_from_baseline:FEV1_base_mL + (1 | pid)"))
  co <- summary(lmer(f, data = dd))$coefficients
  vapply(OTHER[[sname]], function(g) {
    t <- paste0("years_from_baseline:g", g)
    if (t %in% rownames(co)) co[t, "Estimate"] else NA_real_ }, numeric(1))
}
stopifnot(identical(OTHER[["S3"]], OTHER[["S4"]]))
pt <- list(S3 = fit_slopes(decline, "S3"), S4 = fit_slopes(decline, "S4"))
for (sn in c("S3", "S4")) {
  ref <- fev1_decline$est_mL_yr[fev1_decline$schema == sn][match(OTHER[[sn]],
          fev1_decline$category[fev1_decline$schema == sn])]
  stopifnot(isTRUE(all.equal(unname(pt[[sn]]), ref)))
}
.want <- list(n = nrow(decline), B = B_FEV, version = 1L,
              labels = substr(digest::digest(paste(decline$pid, decline$S3, decline$S4,
                                                   decline$cons)), 1, 12))
.cached <- if (file.exists(FPATH)) readRDS(FPATH) else NULL
if (!is.null(.cached) && identical(.cached$fingerprint, .want)) {
  D <- .cached$D
} else {
  set.seed(CV_SEED)
  rows_by <- split(seq_len(nrow(decline)), decline$pid); ids <- names(rows_by)
  D <- matrix(NA_real_, B_FEV, length(OTHER[["S3"]]), dimnames = list(NULL, OTHER[["S3"]]))
  for (b in seq_len(B_FEV)) {
    draw <- sample(ids, length(ids), replace = TRUE); idx <- rows_by[draw]
    dd <- decline[unlist(idx, use.names = FALSE), ]
    dd$pid <- rep(seq_along(draw), lengths(idx))   # a participant drawn twice is two participants
    D[b, ] <- tryCatch(suppressMessages(fit_slopes(dd, "S3") - fit_slopes(dd, "S4")),
                       error = function(e) NA_real_)
  }
  saveRDS(list(D = D, fingerprint = .want), FPATH)
}
fev1_noct_vs_esi <- do.call(rbind, lapply(OTHER[["S3"]], function(g) {
  v <- D[, g]; v <- v[is.finite(v)]
  data.frame(category = g, est_noct = pt$S3[[g]], est_esi = pt$S4[[g]],
             diff_noct_minus_esi = pt$S3[[g]] - pt$S4[[g]],
             lo = unname(quantile(v, .025)), hi = unname(quantile(v, .975)),
             p_boot = min(1, 2 * min(mean(v <= 0), mean(v >= 0))), B_eff = length(v),
             stringsAsFactors = FALSE) }))
stopifnot(nrow(fev1_noct_vs_esi) == 3, all(fev1_noct_vs_esi$B_eff >= 0.95 * B_FEV))
write.csv(fev1_noct_vs_esi, file.path(OUT_DIR, "fev1_noct_vs_esi.csv"), row.names = FALSE)
print(fev1_noct_vs_esi, digits = 3)
```

# Section 8c -- Continuous ESI

ESI as a continuous measure rather than a thresholded criterion. Four
questions, each of which the manuscript's Discussion refers to.

## Section 8c.1 -- Does ESI add to FEV~1~/FVC?

```{r continuous-esi-prognosis}
d_cont <- mort %>%
  filter(!is.na(stratum), !is.na(FEV1_FVC_post))
d_cont$stratum <- relevel(factor(d_cont$stratum), ref = "GOLD0")
# FEV1/FVC on a per-0.1 scale so its coefficient is not a 100-fold number
# beside ESI's per-unit one.
d_cont$FF_per_0_1 <- d_cont$FEV1_FVC_post / 0.1

cont_rows <- list()
for (o in c("all-cause", "respiratory")) {
  ev <- if (o == "all-cause") "vital_status" else "ev_resp"
  RHS <- "age_visit + gender + race + SmokCigNow + ATS_PackYears + stratum"
  m_e <- coxph(as.formula(sprintf("Surv(py, %s) ~ ESI_v1post + %s", ev, RHS)), data = d_cont)
  m_f <- coxph(as.formula(sprintf("Surv(py, %s) ~ FF_per_0_1 + %s", ev, RHS)), data = d_cont)
  m_b <- coxph(as.formula(sprintf("Surv(py, %s) ~ ESI_v1post + FF_per_0_1 + %s", ev, RHS)), data = d_cont)
  lr  <- anova(m_f, m_b)
  ce  <- summary(m_b)$conf.int; co <- summary(m_b)$coef
  cont_rows[[length(cont_rows) + 1]] <- data.frame(
    outcome = o,
    ESI_HR   = ce["ESI_v1post", 1], ESI_LCI = ce["ESI_v1post", 3],
    ESI_UCI  = ce["ESI_v1post", 4], ESI_p   = co["ESI_v1post", 5],
    FF_HR    = ce["FF_per_0_1", 1], FF_LCI  = ce["FF_per_0_1", 3],
    FF_UCI   = ce["FF_per_0_1", 4], FF_p    = co["FF_per_0_1", 5],
    lr_chisq = lr$Chisq[2], lr_p = lr$`Pr(>|Chi|)`[2],
    n_subj = nrow(d_cont), stringsAsFactors = FALSE)
}
cont_esi <- do.call(rbind, cont_rows)
write.csv(cont_esi, file.path(OUT_DIR, "continuous_esi_mortality.csv"), row.names = FALSE)
kable(cont_esi, digits = 3,
      caption = "Continuous ESI beside FEV1/FVC in the same model. HR per 1 ESI unit and per 0.1 FEV1/FVC; the likelihood ratio tests adding ESI to a FEV1/FVC model.")
```

## Section 8c.1b -- ESI against the two visual CT criteria

The ESI criterion stands in for the two visual CT criteria, so how well it
tracks them, and where, is what explains the pattern in the classification
results. Reported with FEV1/FVC alongside, because the question is not whether
ESI relates to CT abnormality but whether it does so beyond the spirometry
already in the framework, and stratified by airflow limitation, because a
pooled figure over a cohort this heterogeneous mostly reflects severity.

```{r esi-visual-ct}
auc <- function(score, lab) {
  lab <- as.logical(lab)
  n1 <- sum(lab); n0 <- sum(!lab)
  if (n1 == 0 || n0 == 0) return(NA_real_)
  r <- rank(score)
  (sum(r[lab]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
d_b$emph_pos <- d_b$emph_yn
d_b$wall_pos <- d_b$wall_yn

# Mean ESI at each level of each visual scale.
lvl_rows <- list()
EMPH_LAB <- c("none","trace","mild","moderate","confluent","advanced destructive")
for (k in 0:5) {
  i <- which(d_b$CT_Visual_Emph_Severity == k)
  if (length(i)) lvl_rows[[length(lvl_rows)+1]] <- data.frame(
    criterion = "Visual emphysema", level = k, label = EMPH_LAB[k+1],
    n = length(i), mean_ESI = mean(d_b$ESI[i]), stringsAsFactors = FALSE)
}
WALL_LAB <- c("absent","borderline","definite")
for (k in 0:2) {
  i <- which(d_b$CT_Visual_Wall_Thickening == k)
  if (length(i)) lvl_rows[[length(lvl_rows)+1]] <- data.frame(
    criterion = "Airway wall thickening", level = k, label = WALL_LAB[k+1],
    n = length(i), mean_ESI = mean(d_b$ESI[i]), stringsAsFactors = FALSE)
}
esi_ct_levels <- do.call(rbind, lvl_rows)
# ESI must rise monotonically across both scales for the dose-response
# statement in the Results to hold.
for (cr in unique(esi_ct_levels$criterion)) {
  v <- esi_ct_levels$mean_ESI[esi_ct_levels$criterion == cr]
  stopifnot(all(diff(v) > 0))
}
write.csv(esi_ct_levels, file.path(OUT_DIR, "esi_ct_levels.csv"), row.names = FALSE)

# Discrimination of each criterion, pooled and by airflow limitation.
strata <- list(list("All participants", rep(TRUE, nrow(d_b))),
               list("Airflow limitation", d_b$afl),
               list("Preserved spirometry", !d_b$afl))
crit <- list(list("Visual emphysema", d_b$emph_pos),
             list("Airway wall thickening", d_b$wall_pos))
auc_rows <- list()
for (st in strata) for (cr in crit) {
  i <- which(st[[2]]); y <- cr[[2]][i]
  auc_rows[[length(auc_rows)+1]] <- data.frame(
    stratum = st[[1]], criterion = cr[[1]], n = length(i),
    prevalence = 100 * mean(y),
    auc_ESI = auc(d_b$ESI[i], y),
    auc_FEV1FVC = auc(-d_b$FEV1_FVC_post[i], y), stringsAsFactors = FALSE)
}
esi_ct_auc <- do.call(rbind, auc_rows)
write.csv(esi_ct_auc, file.path(OUT_DIR, "esi_ct_auc.csv"), row.names = FALSE)
kable(esi_ct_levels, digits = 2, caption = "Mean ESI at each level of the two visual CT criteria.")
kable(esi_ct_auc, digits = 3, caption = "Discrimination of each visual CT criterion, by ESI and by FEV1/FVC.")
```

## Section 8c.2 -- Baseline ESI and subsequent decline in GOLD 0

```{r continuous-esi-gold0}
g0 <- fev1_long %>%
  inner_join(d_b %>% transmute(pid,
               ESI_baseline      = ESI_v1post,
               FEV1_FVC_baseline = FEV1_FVC_post,
               stratum_baseline  = stratum,
               Height_CM         = as.numeric(Height_CM),
               gender_baseline   = factor(gender),
               race_baseline     = factor(race)), by = "pid") %>%
  filter(stratum_baseline == "GOLD0", !is.na(Height_CM), !is.na(age_visit),
         !is.na(SmokCigNow), !is.na(ESI_baseline), !is.na(FEV1_FVC_baseline))

lmm_g0 <- lmer(FEV1_post_mL ~ years_from_baseline * (ESI_baseline + FEV1_FVC_baseline) +
                 Height_CM + age_visit + ATS_PackYears + SmokCigNow +
                 gender_baseline + race_baseline + (1 | pid), data = g0)
esi_int <- lmm_row(lmm_g0, "years_from_baseline:ESI_baseline")
g0_out <- data.frame(
  subgroup = "GOLD 0", est_mL_yr = unname(esi_int["est"]), se = unname(esi_int["se"]),
  p = unname(esi_int["p"]), n_subj = length(unique(g0$pid)),
  stringsAsFactors = FALSE)
write.csv(g0_out, file.path(OUT_DIR, "continuous_esi_gold0_decline.csv"), row.names = FALSE)
cat(sprintf("GOLD 0: each 1-unit higher baseline ESI predicts %.2f mL/yr of additional FEV1 decline (SE %.2f, p = %.3g), n = %d\n",
            esi_int["est"], esi_int["se"], esi_int["p"], nrow(g0_out) && length(unique(g0$pid))))
```

## Section 8c.3 -- ESI trajectory, PRISm against GOLD 0

```{r esi-trajectory}
esi_long <- esi_raw %>%
  filter(visitnum %in% c(1, 2, 3), PrePost == 1) %>%
  group_by(pid, visitnum) %>%
  summarise(ESI = mean(ESI, na.rm = TRUE), .groups = "drop")
esi_v1 <- esi_long %>% filter(visitnum == 1) %>% dplyr::select(pid, ESI_v1 = ESI)
traj <- esi_long %>%
  filter(visitnum > 1) %>%
  inner_join(esi_v1, by = "pid") %>%
  mutate(dESI = ESI - ESI_v1) %>%
  inner_join(d_b %>% dplyr::select(pid, stratum), by = "pid") %>%
  filter(stratum %in% c("GOLD0", "PRISm")) %>%
  group_by(stratum, visitnum) %>%
  summarise(n = dplyr::n(), mean_dESI = mean(dESI, na.rm = TRUE),
            sd_dESI = sd(dESI, na.rm = TRUE), .groups = "drop") %>%
  as.data.frame()
write.csv(traj, file.path(OUT_DIR, "esi_trajectory.csv"), row.names = FALSE)
kable(traj, digits = 3, caption = "Change in ESI from baseline, by baseline stratum.")
```

## Section 8c.4 -- Bronchodilator response

Computed on this cohort. The earlier analysis reported this on all paired
measurements in COPDGene, which is a larger and differently constituted group
than the one this paper analyzes.

```{r bronchodilator-delta}
esi_pair <- esi_raw %>%
  filter(visitnum == 1, PrePost %in% c(0, 1)) %>%
  group_by(pid, PrePost) %>%
  summarise(ESI = mean(ESI, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = PrePost, values_from = ESI, names_prefix = "ESI_") %>%
  filter(!is.na(ESI_0), !is.na(ESI_1)) %>%
  filter(pid %in% d_b$pid)          # this cohort only

bd <- data.frame(n_paired = nrow(esi_pair),
                 mean_delta = mean(esi_pair$ESI_1 - esi_pair$ESI_0),
                 sd_delta   = sd(esi_pair$ESI_1 - esi_pair$ESI_0),
                 stringsAsFactors = FALSE)
write.csv(bd, file.path(OUT_DIR, "bronchodilator_delta_esi.csv"), row.names = FALSE)
cat(sprintf("Bronchodilator change in ESI: mean %.3f (SD %.2f) over %d paired measurements in this cohort\n",
            bd$mean_delta, bd$sd_delta, bd$n_paired))
```

# Section 9 -- Concordant and discordant classification (MD-COPD vs ESI-MD-COPD)

Pairwise, as in the earlier analysis. NoCT-MD-COPD is not carried into this
comparison: it underdiagnoses a group with raised mortality, which the
preceding sections establish, so the question of what happens to individually
discordant participants is asked of ESI-MD-COPD alone.

Binary COPD status under each classification is cross-classified into four
groups. Restricted to preserved spirometry, where the two classifications can
actually disagree about a diagnosis rather than about which COPD category
applies.

```{r discordance-groups}
copd_of <- function(v) v %in% c("COPD-minor", "COPD-major")
mk_discord <- function(dd) factor(
  ifelse( copd_of(dd$S2) &  copd_of(dd$S4), "Both-COPD",
  ifelse( copd_of(dd$S2) & !copd_of(dd$S4), "CT-only-COPD",
  ifelse(!copd_of(dd$S2) &  copd_of(dd$S4), "ESI-only-COPD", "Both-noCOPD"))),
  levels = c("Both-noCOPD", "ESI-only-COPD", "CT-only-COPD", "Both-COPD"))

d_b$discord  <- mk_discord(d_b)
mort$discord <- mk_discord(mort)
exa$discord  <- mk_discord(exa)

# Preserved spirometry only.
d_ps    <- d_b[!d_b$afl, ]
mort_ps <- mort[!mort$afl, ]
exa_ps  <- exa[!exa$afl, ]
cat(sprintf("preserved-spirometry cohort: %d (mortality %d, exacerbation %d)\n\n",
            nrow(d_ps), nrow(mort_ps), nrow(exa_ps)))
print(table(d_ps$discord))

# CT-only-COPD is what ESI misses; ESI-only-COPD is what it adds.
disc_n <- as.data.frame(table(d_ps$discord), stringsAsFactors = FALSE)
names(disc_n) <- c("group", "n")
write.csv(disc_n, file.path(OUT_DIR, "discord_counts.csv"), row.names = FALSE)

# Structural and spirometric profile of the four groups, restored from the v15
# draft. The point is that the two discordant groups are picked out by
# different measurements, so their CT and ESI profiles should differ in
# opposite directions.
.pct <- function(v) 100 * mean(v, na.rm = TRUE)
.msd <- function(v) c(m = mean(v, na.rm = TRUE), s = sd(v, na.rm = TRUE),
                      n = sum(!is.na(v)))
disc_rows_p <- list()
for (g in levels(d_ps$discord)) {
  z <- d_ps[d_ps$discord == g, ]
  e <- .msd(z$ESI); l <- .msd(z$Insp_LAA950_total_Thirona)
  q <- .msd(z$PRM_pct_emphysema_Thirona)
  disc_rows_p[[length(disc_rows_p) + 1]] <- data.frame(
    group = g, n = nrow(z),
    pct_visual_emph = .pct(z$emph_yn),
    pct_wall_thick  = .pct(z$wall_yn),
    mean_ESI = unname(e["m"]), sd_ESI = unname(e["s"]),
    mean_LAA950 = unname(l["m"]), sd_LAA950 = unname(l["s"]),
    n_LAA950 = unname(l["n"]),
    mean_PRM = unname(q["m"]), sd_PRM = unname(q["s"]),
    stringsAsFactors = FALSE)
}
disc_profile <- do.call(rbind, disc_rows_p)
stopifnot(nrow(disc_profile) == 4, all(disc_profile$n > 0))
write.csv(disc_profile, file.path(OUT_DIR, "discord_profile.csv"), row.names = FALSE)

# The same cross-classification within the airflow-limitation stratum, where
# MD-COPD and ESI-MD-COPD assign COPD-major or AFL-only and therefore disagree
# about whether a diagnosis is given rather than about which group applies.
# The reference is the participants both call AFL-only.
d_al <- d_b[d_b$afl, ]; mort_al <- mort[mort$afl, ]; exa_al <- exa[exa$afl, ]
mk_al <- function(dd) factor(
  ifelse(dd$S2 == "COPD-major" & dd$S4 == "COPD-major", "Both-COPD",
  ifelse(dd$S2 == "COPD-major", "CT-only-COPD",
  ifelse(dd$S4 == "COPD-major", "ESI-only-COPD", "Both-AFL-only"))),
  levels = c("Both-AFL-only", "ESI-only-COPD", "CT-only-COPD", "Both-COPD"))
d_al$discord <- mk_al(d_al); mort_al$discord <- mk_al(mort_al)
exa_al$discord <- mk_al(exa_al)
stopifnot(nrow(d_al) == sum(d_b$afl), !anyNA(d_al$discord))

rate_rows <- list(); prof_rows <- list()
for (g in levels(d_al$discord)) {
  i <- mort_al$discord == g; j <- exa_al$discord == g; z <- d_al[d_al$discord == g, ]
  rate_rows[[length(rate_rows) + 1]] <- data.frame(
    group = g, n = nrow(z), n_mort = sum(i), deaths = sum(mort_al$vital_status[i]),
    resp_deaths = sum(mort_al$ev_resp[i]),
    rate_all_100py  = 100 * sum(mort_al$vital_status[i]) / sum(mort_al$py[i]),
    rate_resp_100py = 100 * sum(mort_al$ev_resp[i]) / sum(mort_al$py[i]),
    n_exac = sum(j),
    rate_exac_100py = 100 * sum(exa_al$Total_Exacerbations[j]) / sum(exa_al$Years_Followed[j]),
    prior_exac_mean = mean(exa_al$prior_exac[j]), stringsAsFactors = FALSE)
  prof_rows[[length(prof_rows) + 1]] <- data.frame(
    group = g, n = nrow(z),
    pct_visual_emph = 100 * mean(z$emph_yn), pct_wall_thick = 100 * mean(z$wall_yn),
    pct_meets_esi   = 100 * mean(z$ESI >= p4$t_low),
    mean_ESI = mean(z$ESI),
    mean_LAA950 = mean(z$Insp_LAA950_total_Thirona, na.rm = TRUE),
    mean_PRM = mean(z$PRM_pct_emphysema_Thirona, na.rm = TRUE),
    pct_dyspnea = 100 * mean(z$dysp_yn), pct_sgrq = 100 * mean(z$qol_yn),
    pct_cb = 100 * mean(z$cb_yn), stringsAsFactors = FALSE)
}
disc_rates_al <- do.call(rbind, rate_rows)
disc_profile_al <- do.call(rbind, prof_rows)

m1 <- coxph(as.formula(paste("Surv(py, vital_status) ~ discord +", COV)), data = mort_al)
m2 <- coxph(as.formula(paste("Surv(py, ev_resp) ~ discord +", COV)), data = mort_al)
n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ discord +", COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa_al)
cm <- summary(m1)$conf.int; cr <- summary(m2)$conf.int; ce <- summary(n1)$coef
adj_rows <- list()
for (g in levels(d_al$discord)[-1]) {
  r <- paste0("discord", g); k <- disc_rates_al$group == g
  gv <- function(M, j) if (r %in% rownames(M)) M[r, j] else NA_real_
  adj_rows[[length(adj_rows) + 1]] <- data.frame(
    group = g, deaths_all = disc_rates_al$deaths[k],
    deaths_resp = disc_rates_al$resp_deaths[k],
    all_HR = gv(cm, 1), all_LCI = gv(cm, 3), all_UCI = gv(cm, 4),
    resp_HR = gv(cr, 1), resp_LCI = gv(cr, 3), resp_UCI = gv(cr, 4),
    exac_IRR = if (r %in% rownames(ce)) exp(ce[r, 1]) else NA_real_,
    exac_LCI = if (r %in% rownames(ce)) exp(ce[r, 1] - 1.96 * ce[r, 2]) else NA_real_,
    exac_UCI = if (r %in% rownames(ce)) exp(ce[r, 1] + 1.96 * ce[r, 2]) else NA_real_,
    all_few_events  = disc_rates_al$deaths[k] < MIN_EVENTS,
    resp_few_events = disc_rates_al$resp_deaths[k] < MIN_EVENTS,
    stringsAsFactors = FALSE)
}
disc_adj_al <- do.call(rbind, adj_rows)
stopifnot(nrow(disc_rates_al) == 4, nrow(disc_adj_al) == 3,
          sum(disc_rates_al$n) == sum(d_b$afl))
write.csv(disc_rates_al,   file.path(OUT_DIR, "discord_rates_afl.csv"),   row.names = FALSE)
write.csv(disc_adj_al,     file.path(OUT_DIR, "discord_adjusted_afl.csv"), row.names = FALSE)
write.csv(disc_profile_al, file.path(OUT_DIR, "discord_profile_afl.csv"),  row.names = FALSE)
print(disc_rates_al[, c("group","n","deaths","resp_deaths","rate_all_100py","rate_exac_100py")],
      digits = 3)
print(disc_profile[, c("group", "n", "pct_visual_emph", "pct_wall_thick",
                       "mean_ESI", "mean_LAA950", "mean_PRM")], digits = 3)
```

```{r discordance-rates}
# Observed rates per 100 person-years, and prior exacerbation burden, which
# the earlier analysis found explains much of the gradient in future
# exacerbations.
rate100 <- function(ev, py) 100 * sum(ev) / sum(py)
rows <- lapply(levels(d_ps$discord), function(g) {
  m <- mort_ps[mort_ps$discord == g, ]
  e <- exa_ps [exa_ps$discord  == g, ]
  data.frame(
    group          = g,
    n              = sum(d_ps$discord == g),
    n_mort         = nrow(m),
    deaths         = sum(m$vital_status),
    resp_deaths    = sum(m$ev_resp),
    rate_all_100py = rate100(m$vital_status, m$py),
    rate_resp_100py= rate100(m$ev_resp,      m$py),
    n_exac         = nrow(e),
    rate_exac_100py= rate100(e$Total_Exacerbations, e$Years_Followed),
    prior_exac_mean= mean(e$prior_exac, na.rm = TRUE),
    stringsAsFactors = FALSE)
})
disc_rates <- do.call(rbind, rows)
# Rates must be ordered by increasing severity of the group definition for the
# narrative to hold; assert nothing, but surface it.
write.csv(disc_rates, file.path(OUT_DIR, "discord_rates.csv"), row.names = FALSE)
kable(disc_rates, digits = 2,
      caption = "Observed rates per 100 person-years by cross-classification group.")
```

```{r discordance-adjusted}
# Adjusted estimates against Both-noCOPD.
cm <- coxph(as.formula(paste("Surv(py, vital_status) ~ discord +", COV)), data = mort_ps)
cr <- coxph(as.formula(paste("Surv(py, ev_resp) ~ discord +",      COV)), data = mort_ps)
ne <- glm.nb(as.formula(paste("Total_Exacerbations ~ discord +", COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa_ps)
sm <- summary(cm)$conf.int; sr <- summary(cr)$conf.int; se <- summary(ne)$coef

# A group with no events of a given type has no estimable hazard ratio. The
# Cox fit still returns a number: for CT-only-COPD, which has zero respiratory
# deaths, it returns 3e-07 with an interval of 0 to Inf. That is a monotone
# likelihood, not an effect size, and printing it as one would be worse than
# printing nothing. Such cells are set NA and the event count is carried
# alongside so the reason is visible in the artifact rather than inferred.
n_ev <- function(g, col) sum(mort_ps[[col]][mort_ps$discord == g])
adj <- do.call(rbind, lapply(levels(d_ps$discord)[-1], function(g) {
  r <- paste0("discord", g)
  gv <- function(M, j) if (r %in% rownames(M)) M[r, j] else NA_real_
  d_all  <- n_ev(g, "vital_status")
  d_resp <- n_ev(g, "ev_resp")
  est <- function(v, ok) if (ok) v else NA_real_
  data.frame(group = g,
    deaths_all = d_all, deaths_resp = d_resp,
    all_HR  = est(gv(sm,1), d_all  > 0), all_LCI = est(gv(sm,3), d_all  > 0),
    all_UCI = est(gv(sm,4), d_all  > 0),
    resp_HR = est(gv(sr,1), d_resp > 0), resp_LCI = est(gv(sr,3), d_resp > 0),
    resp_UCI= est(gv(sr,4), d_resp > 0),
    exac_IRR = if (r %in% rownames(se)) exp(se[r,1]) else NA_real_,
    exac_LCI = if (r %in% rownames(se)) exp(se[r,1]-1.96*se[r,2]) else NA_real_,
    exac_UCI = if (r %in% rownames(se)) exp(se[r,1]+1.96*se[r,2]) else NA_real_,
    stringsAsFactors = FALSE)
}))
# Any surviving estimate must be finite and positive; a degenerate fit that
# slipped through the event-count guard would still be caught here.
.fin <- function(v) all(is.na(v) | (is.finite(v) & v > 0))
stopifnot(.fin(adj$all_HR), .fin(adj$resp_HR), .fin(adj$exac_IRR),
          .fin(adj$resp_UCI))
write.csv(adj, file.path(OUT_DIR, "discord_adjusted.csv"), row.names = FALSE)
kable(adj, digits = 2, caption = "Adjusted estimates against Both-noCOPD.")
```

# Section 9b -- Both strata together, for Figures 5 and 6

Profile, crude ratios and adjusted estimates for the concordant and discordant
groups in each stratum, laid out identically so the figures can set the strata
side by side. Crude intervals follow the rest of the paper: exact Poisson for
deaths, subject bootstrap for exacerbations, and no interval below the event
floor. Also the per-pathway agreement the Results quote, with a paired
bootstrap P for the difference between the two classifications.

```{r strata-combined}
set.seed(CV_SEED)
se_of <- function(v) { v <- v[!is.na(v)]; sd(v) / sqrt(length(v)) }
STRATA <- list(
  list(stratum = "Preserved spirometry", d = d_ps, m = mort_ps, e = exa_ps,
       ref = "Both-noCOPD"),
  list(stratum = "Airflow limitation", d = d_al, m = mort_al, e = exa_al,
       ref = "Both-AFL-only"))

prof_rows <- list(); crude_rows_s <- list()
for (st in STRATA) {
  lv <- levels(st$d$discord); ref <- st$ref
  stopifnot(ref %in% lv, ref %in% st$m$discord, ref %in% st$e$discord)
  for (g in lv) {
    z <- st$d[st$d$discord == g, ]
    prof_rows[[length(prof_rows) + 1]] <- data.frame(
      stratum = st$stratum, group = g, n = nrow(z),
      mean_ESI = mean(z$ESI), se_ESI = se_of(z$ESI),
      mean_LAA950 = mean(z$Insp_LAA950_total_Thirona, na.rm = TRUE),
      se_LAA950 = se_of(z$Insp_LAA950_total_Thirona),
      pct_visual_emph = 100 * mean(z$emph_yn), pct_wall_thick = 100 * mean(z$wall_yn),
      pct_dyspnea = 100 * mean(z$dysp_yn), pct_sgrq = 100 * mean(z$qol_yn),
      pct_cb = 100 * mean(z$cb_yn), pct_meets_esi = 100 * mean(z$ESI >= p4$t_low),
      stringsAsFactors = FALSE)
  }
  m <- st$m; e <- st$e
  ratio <- function(ev, py, dat, g)
    (sum(dat[[ev]][dat$discord == g]) / sum(dat[[py]][dat$discord == g])) /
    (sum(dat[[ev]][dat$discord == ref]) / sum(dat[[py]][dat$discord == ref]))
  others <- setdiff(lv, ref)
  boot_e <- replicate(CRUDE_B, {
    dd <- e[sample.int(nrow(e), nrow(e), replace = TRUE), ]
    vapply(others, function(g) ratio("Total_Exacerbations", "Years_Followed", dd, g),
           numeric(1)) })
  for (g in others) {
    for (o in c("all", "resp", "exac")) {
      if (o == "exac") {
        ev <- sum(e$Total_Exacerbations[e$discord == g])
        rr <- ratio("Total_Exacerbations", "Years_Followed", e, g)
        v <- boot_e[g, ]; v <- v[is.finite(v)]
        lo <- unname(quantile(v, .025)); hi <- unname(quantile(v, .975))
      } else {
        col <- if (o == "all") "vital_status" else "ev_resp"
        ev <- sum(m[[col]][m$discord == g])
        rr <- ratio(col, "py", m, g)
        ci <- exact_rate_ratio(ev, sum(m$py[m$discord == g]),
                               sum(m[[col]][m$discord == ref]), sum(m$py[m$discord == ref]))
        lo <- ci[1]; hi <- ci[2]
      }
      crude_rows_s[[length(crude_rows_s) + 1]] <- data.frame(
        stratum = st$stratum, group = g, outcome = o, events = ev,
        rr = rr, lo = lo, hi = hi, few_events = ev < MIN_EVENTS,
        stringsAsFactors = FALSE)
    }
  }
}
disc_profile_strata <- do.call(rbind, prof_rows)
disc_crude_strata   <- do.call(rbind, crude_rows_s)
stopifnot(nrow(disc_profile_strata) == 8, nrow(disc_crude_strata) == 18,
          sum(disc_profile_strata$n) == nrow(d_b))

cols <- c("group", "deaths_all", "deaths_resp", "all_HR", "all_LCI", "all_UCI",
          "resp_HR", "resp_LCI", "resp_UCI", "exac_IRR", "exac_LCI", "exac_UCI")
stopifnot(all(cols %in% names(adj)), all(cols %in% names(disc_adj_al)))
disc_adj_strata <- rbind(
  cbind(stratum = "Preserved spirometry", adj[, cols], stringsAsFactors = FALSE),
  cbind(stratum = "Airflow limitation",  disc_adj_al[, cols], stringsAsFactors = FALSE))
stopifnot(nrow(disc_adj_strata) == 6)

# Per-pathway agreement, both directions, with a paired bootstrap for the
# difference in the share of each classification's assignments MD-COPD agrees
# with. Paired because both classifications label the same participants.
L3 <- d_b[, c("S2", "S3", "S4")]
share <- function(dd, s, g) { a <- dd[[s]] == g; if (!any(a)) NA_real_ else mean(dd$S2[a] == g) }
bs_ag <- replicate(CRUDE_B, {
  dd <- L3[sample.int(nrow(L3), nrow(L3), replace = TRUE), ]
  vapply(O, function(g) share(dd, "S4", g) - share(dd, "S3", g), numeric(1)) })
agree_rows <- lapply(O, function(g) {
  v <- bs_ag[g, ]; v <- v[is.finite(v)]
  data.frame(category = g, n_md = sum(L3$S2 == g),
             n_noct = sum(L3$S3 == g), agree_noct = sum(L3$S3 == g & L3$S2 == g),
             n_esi  = sum(L3$S4 == g), agree_esi  = sum(L3$S4 == g & L3$S2 == g),
             recall_noct = 100 * mean(L3$S3[L3$S2 == g] == g),
             recall_esi  = 100 * mean(L3$S4[L3$S2 == g] == g),
             p_boot = 2 * min(mean(v <= 0), mean(v >= 0)), B_eff = length(v),
             stringsAsFactors = FALSE) })
agreement_by_group <- do.call(rbind, agree_rows)
agreement_by_group$pct_noct <- 100 * agreement_by_group$agree_noct / agreement_by_group$n_noct
agreement_by_group$pct_esi  <- 100 * agreement_by_group$agree_esi  / agreement_by_group$n_esi
stopifnot(sum(agreement_by_group$n_md) == nrow(d_b),
          sum(agreement_by_group$agree_esi)  == sum(d_b$S4 == d_b$S2),
          sum(agreement_by_group$agree_noct) == sum(d_b$S3 == d_b$S2))

write.csv(disc_profile_strata, file.path(OUT_DIR, "discord_profile_strata.csv"),  row.names = FALSE)
write.csv(disc_crude_strata,   file.path(OUT_DIR, "discord_crude_strata.csv"),    row.names = FALSE)
write.csv(disc_adj_strata,     file.path(OUT_DIR, "discord_adjusted_strata.csv"), row.names = FALSE)
write.csv(agreement_by_group,  file.path(OUT_DIR, "agreement_by_group.csv"),      row.names = FALSE)
print(agreement_by_group[, c("category", "pct_noct", "pct_esi", "recall_noct", "recall_esi", "p_boot")],
      digits = 3)
```

```{r strata-profile-tests}
# Pairwise tests behind the Figure 5 brackets, within each stratum: Wilcoxon
# rank-sum for the two continuous measures, Fisher's exact test for the five
# criteria. Unadjusted. Where neither group has a criterion, or both have it
# throughout, there is nothing to test and P is 1.
PT_MEAS  <- c("ESI", "Insp_LAA950_total_Thirona", "emph_yn", "wall_yn",
              "dysp_yn", "qol_yn", "cb_yn")
PT_PAIRS <- list(c("CT-only-COPD", "ESI-only-COPD"), c("CT-only-COPD", "Both-COPD"),
                 c("ESI-only-COPD", "Both-COPD"))
pt_rows <- list()
for (st in STRATA) for (pr in PT_PAIRS) for (v in PT_MEAS) {
  a <- st$d[[v]][st$d$discord == pr[1]]; b <- st$d[[v]][st$d$discord == pr[2]]
  a <- a[!is.na(a)]; b <- b[!is.na(b)]
  cont <- v %in% PT_MEAS[1:2]
  p <- if (cont) wilcox.test(a, b, exact = FALSE)$p.value else {
    k <- sum(a) + sum(b)
    if (k == 0 || k == length(a) + length(b)) 1 else
      fisher.test(matrix(c(sum(a), length(a) - sum(a), sum(b), length(b) - sum(b)), 2))$p.value }
  pt_rows[[length(pt_rows) + 1]] <- data.frame(
    stratum = st$stratum, measure = v, group1 = pr[1], group2 = pr[2],
    n1 = length(a), n2 = length(b), test = if (cont) "Wilcoxon" else "Fisher",
    p = p, stringsAsFactors = FALSE)
}
disc_profile_tests <- do.call(rbind, pt_rows)
stopifnot(nrow(disc_profile_tests) == 42, all(is.finite(disc_profile_tests$p)),
          all(disc_profile_tests$n1 > 0), all(disc_profile_tests$n2 > 0))
write.csv(disc_profile_tests, file.path(OUT_DIR, "discord_profile_tests.csv"), row.names = FALSE)
print(disc_profile_tests[disc_profile_tests$group2 == "ESI-only-COPD", ], digits = 3)
```

# Section 10 -- Do the classifications differ? Paired bootstrap

The preceding sections report each classification's estimates side by side.
Reading a difference off two intervals is not a test of one, so this resamples
participants and refits BOTH classifications within every resample, taking the
difference of the log estimates for the same category. Pairing is what makes
the comparison valid: both estimates come from the same people in each draw,
so the between-classification correlation is preserved rather than assumed
away.

Each outcome is fitted in its own `tryCatch`. A warning from a sparse
respiratory fit must not discard the valid all-cause fits for that resample,
because dropped resamples are not missing at random: they are precisely the
ones with few respiratory events. Effective B is therefore tracked per outcome.

```{r paired-bootstrap}
B_DIFF <- 1000L
CACHE  <- file.path(OUT_DIR, "_cache"); dir.create(CACHE, showWarnings = FALSE)
CPATH  <- file.path(CACHE, "paired_bootstrap.rds")
CATS   <- c("AFL-only", "COPD-minor", "COPD-major")

# The fingerprint carries the model formulas, not just row counts, so any
# change to an outcome, covariate or offset invalidates the cache whether or
# not anyone remembers to bump a version.
.fml <- c(paste("Surv(py, vital_status) ~ S +", COV),
          paste("Surv(py, ev_resp) ~ S +",      COV),
          paste("Total_Exacerbations ~ S +", COV,
                "+ prior_exac + offset(log(Years_Followed))"))
# The formulas and row counts are identical whether or not the S4 rule
# changed, so a fingerprint built from those alone would silently reuse a
# bootstrap describing the previous classification. Carry the label vectors
# themselves.
.labels <- paste(paste(as.integer(d_b$S2), collapse = ""),
                 paste(as.integer(d_b$S4), collapse = ""), sep = "|")
.want <- list(n_mort = nrow(mort), n_exac = nrow(exa), B = B_DIFF,
              deaths = sum(mort$vital_status), resp = sum(mort$ev_resp),
              cats = paste(CATS, collapse = ","), fml = paste(.fml, collapse = "|"),
              labels = substr(digest::digest(.labels), 1, 12),
              version = 2L)
.cached <- if (file.exists(CPATH)) readRDS(CPATH) else NULL
if (!is.null(.cached) && !identical(.cached$fingerprint, .want)) {
  message("paired-bootstrap cache stale; recomputing")
  .cached <- NULL
}

if (!is.null(.cached)) {
  bootdiff <- .cached
  cat(sprintf("loaded cached paired bootstrap (B effective: all %d, resp %d, exac %d)\n",
              bootdiff$B_eff[["all"]], bootdiff$B_eff[["resp"]], bootdiff$B_eff[["exac"]]))
} else {
  set.seed(CV_SEED)
  mk <- function() matrix(NA_real_, nrow = B_DIFF, ncol = length(CATS),
                          dimnames = list(NULL, CATS))
  store <- list(all_S2 = mk(), all_S4 = mk(), resp_S2 = mk(), resp_S4 = mk(),
                exac_S2 = mk(), exac_S4 = mk())
  lg <- function(fit, term, nb = FALSE) {
    co <- if (nb) summary(fit)$coef else summary(fit)$coef
    if (!term %in% rownames(co)) return(NA_real_)
    unname(co[term, 1])
  }
  fit_cox <- function(dat, ev) tryCatch({
    lhs <- sprintf("Surv(py, %s)", ev)
    list(S2 = coxph(as.formula(paste(lhs, "~ S2 +", COV)), data = dat),
         S4 = coxph(as.formula(paste(lhs, "~ S4 +", COV)), data = dat))
  }, error = function(e) NULL, warning = function(w) NULL)
  fit_nb <- function(dat) tryCatch({
    rhs <- paste(COV, "+ prior_exac + offset(log(Years_Followed))")
    list(S2 = glm.nb(as.formula(paste("Total_Exacerbations ~ S2 +", rhs)), data = dat),
         S4 = glm.nb(as.formula(paste("Total_Exacerbations ~ S4 +", rhs)), data = dat))
  }, error = function(e) NULL, warning = function(w) NULL)

  B_eff <- c(all = 0L, resp = 0L, exac = 0L)
  for (b in seq_len(B_DIFF)) {
    im <- sample.int(nrow(mort), nrow(mort), replace = TRUE); dm <- mort[im, ]
    ie <- sample.int(nrow(exa),  nrow(exa),  replace = TRUE); de <- exa [ie, ]
    fa <- fit_cox(dm, "vital_status"); fr <- fit_cox(dm, "ev_resp"); fe <- fit_nb(de)
    if (!is.null(fa)) { for (g in CATS) {
        store$all_S2[b, g] <- lg(fa$S2, paste0("S2", g))
        store$all_S4[b, g] <- lg(fa$S4, paste0("S4", g)) }
      B_eff["all"] <- B_eff["all"] + 1L }
    if (!is.null(fr)) { for (g in CATS) {
        store$resp_S2[b, g] <- lg(fr$S2, paste0("S2", g))
        store$resp_S4[b, g] <- lg(fr$S4, paste0("S4", g)) }
      B_eff["resp"] <- B_eff["resp"] + 1L }
    if (!is.null(fe)) { for (g in CATS) {
        store$exac_S2[b, g] <- lg(fe$S2, paste0("S2", g), nb = TRUE)
        store$exac_S4[b, g] <- lg(fe$S4, paste0("S4", g), nb = TRUE) }
      B_eff["exac"] <- B_eff["exac"] + 1L }
  }
  bootdiff <- list(store = store, B_eff = B_eff, fingerprint = .want)
  saveRDS(bootdiff, CPATH)
  cat(sprintf("paired bootstrap done (B effective: all %d, resp %d, exac %d)\n",
              B_eff[["all"]], B_eff[["resp"]], B_eff[["exac"]]))
}

# Percentile interval on the difference in log estimates, and a two-sided
# bootstrap p from the proportion of draws on the other side of zero.
diff_row <- function(key, label, g) {
  d <- bootdiff$store[[paste0(key, "_S4")]][, g] -
       bootdiff$store[[paste0(key, "_S2")]][, g]
  d <- d[is.finite(d)]
  if (!length(d)) return(NULL)
  p <- 2 * min(mean(d <= 0), mean(d >= 0)); p <- min(1, p)
  data.frame(outcome = label, category = g,
             ratio_S4_over_S2 = exp(mean(d)),
             lo = exp(unname(quantile(d, .025))),
             hi = exp(unname(quantile(d, .975))),
             p_two_sided = p, B_eff = length(d), stringsAsFactors = FALSE)
}
diffs <- do.call(rbind, unlist(recursive = FALSE, lapply(
  list(c("all","all-cause mortality"), c("resp","respiratory mortality"),
       c("exac","exacerbations")),
  function(o) lapply(CATS, function(g) diff_row(o[1], o[2], g)))))
write.csv(diffs, file.path(OUT_DIR, "schema_diff_bootstrap.csv"), row.names = FALSE)
kable(diffs, digits = 3,
      caption = "ESI-MD-COPD versus MD-COPD, same category, paired resamples. A ratio of 1 means the two classifications assign the same effect size.")
```

# Section 9 -- Self-check (canonical values echoed for drift detection)

Each row prints the freshly-computed value alongside the value cited in
the manuscript (`expected`). Rerunning this report on the cluster should
reproduce the `expected` column exactly (within rounding).

```{r self-check}
selfcheck <- function(name, computed, expected, tol = 0.02) {
  ok <- if (is.numeric(computed) && is.numeric(expected)) {
          isTRUE(abs(computed - expected) < tol)
        } else identical(computed, expected)
  data.frame(check = name,
             computed = if (is.numeric(computed)) sprintf("%.3f", computed)
                        else as.character(computed),
             expected = if (is.numeric(expected)) sprintf("%.3f", expected)
                        else as.character(expected),
             status = ifelse(ok, "PASS", "FAIL"),
             stringsAsFactors = FALSE)
}

R2 <- function(s, cat, fld) risk[[fld]][risk$schema == s & risk$category == cat]
CR <- function(s, cat, out, fld) crude[[fld]][crude$schema == s & crude$category == cat & crude$outcome == out]
GT <- function(o, fld) {
  g <- do.call(rbind, gate_rows); g[[fld]][g$outcome == o]
}
DS <- function(s, fld) {
  d <- do.call(rbind, disc_rows); d[[fld]][d$schema == s]
}

# Canonical values are those cited in the manuscript prose. The point of
# this table is to detect drift on re-run: if any row goes FAIL, an
# expected number must be updated deliberately (with a corresponding
# manuscript edit) rather than silently changed. Every value is loaded
# from the already-written artifact so this block reflects what downstream
# reads, not what the in-memory objects happened to hold.

# MD-COPD counts BEFORE the baseline-ESI restriction are what should
# reproduce Bhatt Table 1 exactly. mdcopd_validation.csv captures that
# state; d_b at self-check time has already been ESI-restricted.
mdval <- read.csv(file.path(OUT_DIR, "mdcopd_validation.csv"),
                  stringsAsFactors = FALSE)
MDV <- function(cat) mdval$ours[mdval$category == cat]

checks <- list(
  # ---- Bhatt Table 1 reproduction (before ESI restriction) --------------
  selfcheck("MD-COPD noCOPD n vs Bhatt 4,439",
            MDV("noCOPD"),          4439L),
  selfcheck("MD-COPD AFL-only n vs Bhatt 282",
            MDV("AFL-only"), 282L),
  selfcheck("MD-COPD COPD-minor n vs Bhatt 811",
            MDV("COPD-minor"),      811L),
  selfcheck("MD-COPD COPD-major n vs Bhatt 3,884",
            MDV("COPD-major"),      3884L),

  # ---- Analytic cohort after ESI restriction ---------------------------
  selfcheck("analytic cohort n (with baseline ESI)",
            as.integer(nrow(d_b)),  9240L, tol = 0.5),

  # ---- Nested LR gate: MD-COPD vs. fixed ratio (2 df) ------------------
  selfcheck("all-cause LR chi-square, MD-COPD over fixed ratio",
            GT("ALL-CAUSE MORTALITY", "lrt_chisq"),   122.1, tol = 0.1),
  selfcheck("respiratory LR chi-square, MD-COPD over fixed ratio",
            GT("RESPIRATORY MORTALITY", "lrt_chisq"),  76.1, tol = 0.1),
  selfcheck("exacerbation LR chi-square, MD-COPD over fixed ratio",
            GT("EXACERBATIONS", "lrt_chisq"),         158.9, tol = 0.1),

  # ---- Fitted CT-free rules --------------------------------------------
  selfcheck("NoCT-MD-COPD count threshold", k3,       2L),
  selfcheck("ESI-MD-COPD count threshold",  p4$k,     2L),
  selfcheck("ESI-MD-COPD T_low",            p4$t_low, 1.50, tol = 0.001),

  # ---- AFL-only risk, the load-bearing category ------------------
  selfcheck("MD-COPD AFL-only all-cause HR",
            R2("S2", "AFL-only", "all_HR"),  0.87, tol = 0.02),
  selfcheck("NoCT-MD-COPD AFL-only respiratory HR (should be highly elevated)",
            R2("S3", "AFL-only", "resp_HR"), 6.37, tol = 0.05),
  selfcheck("ESI-MD-COPD AFL-only all-cause HR (should straddle 1)",
            R2("S4", "AFL-only", "all_HR"),  0.96, tol = 0.02),

  # ---- Discrimination --------------------------------------------------
  selfcheck("MD-COPD all-cause C-index",
            DS("S2", "c_allcause"), 0.703, tol = 0.005)
)
selfcheck_tbl <- do.call(rbind, checks)
write.csv(selfcheck_tbl, file.path(OUT_DIR, "SELFCHECK.csv"), row.names = FALSE)
kable(selfcheck_tbl, caption = "Self-check -- computed vs canonical values.")
cat(sprintf("\nSelf-check summary: %d PASS, %d FAIL (of %d total)\n",
            sum(selfcheck_tbl$status == "PASS"),
            sum(selfcheck_tbl$status == "FAIL"),
            nrow(selfcheck_tbl)))
```

# Section 10 -- Completeness check

Enumerates every CSV downstream tables and figures depend on and aborts if
any is missing, so a partial run is caught here rather than surfacing as a
missing-file error deep in `draft_tables.R` or `build_manuscript.py`.

```{r completeness}
required_csvs <- c(
  "raw_discrimination.csv",
  "followup_medians.csv",
  "group_vs_mdcopd.csv",
  "cohort_flow.csv", "cohort.txt", "mdcopd_validation.csv",
  "schema_labels.csv", "schema_labels.rds", "schema_risk.csv",
  "schema_crude.csv", "schema_discrimination.csv", "schema_binary.csv", "binary_vs_mdcopd.csv",
  "schema_fit.csv", "schema_fit_cv_diff.csv", "schema_fit_cv_test.csv",
  "cv_schema_fits.csv",
  "objective_selection.csv", "reclassification.csv", "crossclass.csv",
  "supp_baseline.csv", "supp_esi_ct.csv", "metric_sweep.csv",
  "f1_by_category.csv", "esi_ct_levels.csv", "esi_ct_auc.csv",
  "discord_counts.csv", "discord_rates.csv", "discord_adjusted.csv",
  "schema_diff_bootstrap.csv", "fev1_decline.csv", "fev1_noct_vs_esi.csv",
  "continuous_esi_mortality.csv", "continuous_esi_gold0_decline.csv",
  "esi_trajectory.csv", "bronchodilator_delta_esi.csv",
  "gate_fixedratio.csv", "gate_exac_aic.csv", "SELFCHECK.csv",
  "consensus_ref_lost.csv", "cv_category_test.csv", "accuracy_test.csv",
  "discord_profile.csv", "discord_rates_afl.csv",
  "discord_adjusted_afl.csv", "discord_profile_afl.csv",
  "discord_profile_strata.csv", "discord_crude_strata.csv",
  "discord_adjusted_strata.csv", "agreement_by_group.csv", "discord_profile_tests.csv"
)
missing <- required_csvs[!file.exists(file.path(OUT_DIR, required_csvs))]
if (length(missing))
  stop("missing artifacts in OUT_DIR: ", paste(missing, collapse = ", "))
cat(sprintf("All %d required artifacts present in %s\n",
            length(required_csvs), normalizePath(OUT_DIR)))
```

# Section 11 -- Provenance

```{r provenance}
provenance <- c(
  sprintf("rendered=%s", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
  sprintf("R=%s", R.version.string),
  sprintf("cohort_n=%d", nrow(d_b)),
  sprintf("cv_seed=%d", CV_SEED),
  sprintf("cv_repeats=%d", CV_REPEATS),
  sprintf("cv_folds=%d", CV_FOLDS),
  sprintf("crude_B=%d", CRUDE_B),
  sprintf("fitted_S3_k=%d", k3),
  sprintf("fitted_S4_k=%d", p4$k),
  sprintf("fitted_S4_t_low=%.3f", p4$t_low),
  sprintf("fitted_S4_threshold_count=%d", 1L))
writeLines(provenance, file.path(OUT_DIR, "PROVENANCE.txt"))
cat(paste(provenance, collapse = "\n"), "\n")
```

# Appendix -- Session info

```{r session}
sessionInfo()
```
