---
title: "ESI in a CT-free multidimensional COPD framework: analysis report accompanying the manuscript"
author: "Peter J. Castaldi"
date: "2026-09-03"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

<!--
Single reproducible analysis report that produces every result, table, and
figure cited in the CT-free MD-COPD manuscript. Bhatt et al.'s JAMA 2025
multidimensional COPD (MD-COPD) framework requires chest CT; this manuscript
asks whether ESI, a spirometry-only measure, can substitute for the CT
criteria while preserving MD-COPD's advantage over the fixed ratio.

Four classifications are compared on the same COPDGene participants:
  1. Fixed ratio    - post-bronchodilator FEV1/FVC < 0.70 alone
  2. MD-COPD        - the published CT-based framework (the reference)
  3. NoCT-MD-COPD   - MD-COPD with the two CT criteria removed
  4. ESI-MD-COPD    - MD-COPD with the two CT criteria replaced by ESI

MD-COPD categories are taken from Bhatt et al.'s own per-subject
classification file (MDCOPD_PATH) rather than reconstructed heuristically.
An earlier reconstruction attempt agreed with them on classification LOGIC
for every one of 9,406 participants classified in both, but differed on
cohort membership by 23 people in identifiable ways (13 with missing symptom
data we kept and they excluded; 10 without a quantitative CT variable that
was never part of their published criteria). Using their file directly
reproduces all four published category counts exactly (n = 9,416: noCOPD
4,439, AFL-only-noCOPD 282, COPD-minor 811, COPD-major 3,884). NoCT-MD-COPD
and ESI-MD-COPD are ours, built from the individual criteria directly.
-->

# Overview

This report reproduces the numerical claims of the manuscript from six
COPDGene source files: ESI values, phenotype/CT, vital status, cause of
death, long-term follow-up exacerbations, and Bhatt et al.'s per-subject
MD-COPD classification. Each section writes its output (tables as CSV,
figures as PNG) into `OUT_DIR`. A closing self-check block echoes every
canonical value cited in the manuscript prose against the expected number
so drift can be detected on any re-run.

# PATHS -- supplied at run time

Copy `config_paths.R.example` to `config_paths.R`, fill in the six input
locations for the machine you are running on, and knit.

```{r paths, fig.keep = "none"}
# ---- Headless raster device --------------------------------------------
# Sets the device for the chunks that follow. It cannot suppress the one X11
# warning seen on a headless node: knitr probes the png device during knit()
# startup, in set_html_dev() -> dev_available("png"), before any chunk runs,
# so no code in this document is early enough. render_ctfree.R sets the
# option before calling knit() and does prevent it; this block keeps the
# chunk devices correct for anyone who knits the file directly.
local({
  .bt <- getOption("bitmapType")
  if (capabilities("cairo") &&
      (is.null(.bt) || !nzchar(.bt) || identical(.bt, "Xlib"))) {
    options(bitmapType = "cairo")
    knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo"))
  }
})

# Locate this .Rmd's own directory so the report runs from any working
# directory. knitr sets the working directory to the .Rmd's location during
# rendering, so getwd() is the primary signal; the --file= fallback covers a
# plain `Rscript -e knitr::knit(...)` invocation, and the last branch covers
# an interactive knit from the repository root.
.b <- grep("^--file=", commandArgs(trailingOnly = FALSE), value = TRUE)
HERE <- {
  if (file.exists("ESI_ctfree_analysis_2026.9.3.Rmd")) {
    normalizePath(".")
  } else if (length(.b) &&
             file.exists(file.path(dirname(normalizePath(sub("^--file=", "", .b[1]))),
                                   "ESI_ctfree_analysis_2026.9.3.Rmd"))) {
    dirname(normalizePath(sub("^--file=", "", .b[1])))
  } else if (file.exists("ctfree/ESI_ctfree_analysis_2026.9.3.Rmd")) {
    normalizePath("ctfree")
  } else {
    stop("cannot locate this report's directory; knit from the ctfree/ ",
         "directory or from the repository root")
  }
}
ROOT    <- dirname(HERE)
CONFIG  <- file.path(ROOT, "config_paths.R")
if (!file.exists(CONFIG))
  stop("cannot find ", CONFIG, "; copy config_paths.R.example to ",
       "config_paths.R and fill in the local file paths")
source(CONFIG)

# This paper writes to its own assets directory, ctfree/assets/, NOT to the
# repository-wide manuscript_assets/ that belongs to the v15 (parent-directory)
# ESI manuscript. Any OUT_DIR set by config_paths.R applies to that older
# analysis and is deliberately ignored here.
OUT_DIR <- file.path(HERE, "assets")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# ---- Analytic constants --------------------------------------------------
CV_REPEATS      <- 5L
CV_FOLDS        <- 5L
CV_SEED         <- 20260901L
CRUDE_B         <- 1000L
BHATT_MDCOPD_N  <- c(noCOPD = 4439L, `AFL-only-NoCOPD` = 282L,
                     `COPD-minor` = 811L, `COPD-major` = 3884L)
```

# Setup

```{r setup, include = FALSE, fig.keep = "none"}
# Print warnings as they happen rather than buffering them to the end of the
# knit. R's default (warn = 0) defers every warning until the top-level call
# returns, so a device warning raised in chunk 12 appears after chunk 68 with
# nothing to attribute it to. With warn = 1 the warning lands immediately
# after the chunk marker that produced it.
options(warn = 1)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dpi = 150, fig.width = 7, fig.height = 5)

suppressPackageStartupMessages({
  library(dplyr); library(survival); library(MASS); library(knitr)
})

read_any <- function(p, ...) {
  h <- readLines(p, n = 1, warn = FALSE)
  read.delim(p, sep = if (grepl("\t", h)) "\t" else ",",
             stringsAsFactors = FALSE, ...)
}
bind_pid <- function(df, label) {
  cand <- c(ID_COL, paste0(ID_COL, ".x"), paste0(ID_COL, ".y"))
  hit <- cand[cand %in% names(df)]
  if (!length(hit)) stop("no identifier column in ", label)
  df$pid <- as.character(df[[hit[1]]]); df
}
```

# Section 1 -- Cohort construction

Bhatt et al.'s published exclusion chain, with the counts they report:

* 10,305 enrolled after removing ILD and bronchiectasis
* less 107 never-smokers                      -> 10,198
* less 66 unacceptable spirometry             -> 10,132
* less 716 excluded from the diagnostic algorithm -> 9,416

The first three steps reproduce exactly by construction (ILD/bronchiectasis
flag, cohort != "Never smoked", `FEV1_FVC_post` non-missing). Their fourth
step (CT quality control + minor-criteria data completeness) and their
MD-COPD classification are NOT reconstructed here: two independent attempts
at a heuristic reconstruction each landed within a handful of participants
of their counts, and a subject-level merge against their own classification
file showed exactly where the residual came from. See the header comment on
`MDCOPD_PATH` in `config_paths.R.example` for the full story. Using their
file directly gives an exact match on all four category counts.

Only after MD-COPD is validated against the published counts do we impose
this analysis's own additional requirement: baseline ESI available.

```{r ingest-and-cohort}
esi_raw <- bind_pid(read_any(ESI_PATH), "ESI_PATH")
phe_raw <- bind_pid(read_any(PHE_PATH, na.strings = c("", "NA")), "PHE_PATH")
vs      <- bind_pid(read_any(VS_PATH),  "VS_PATH")
cod     <- bind_pid(read_any(COD_PATH), "COD_PATH")
ex_raw  <- bind_pid(read_any(EX_PATH),  "EX_PATH")
mdcopd  <- bind_pid(read_any(MDCOPD_PATH), "MDCOPD_PATH")

# ---- Cause of death: TORCH Basic underlying cause ------------------------
ucd <- suppressWarnings(as.integer(cod$Torch_Group_Basic))
cod$UCD_Resp <- as.integer(!is.na(ucd) & ucd == 1L)

# ---- Cohort exclusion chain ----------------------------------------------
v1 <- phe_raw[phe_raw$visitnum == 1, ]
flow <- list(data.frame(step = "visit 1 in phenotype file",
                        n = nrow(v1), bhatt_n = NA_integer_,
                        stringsAsFactors = FALSE))
keep <- function(label, sel, target = NA_integer_) {
  v1 <<- v1[sel, , drop = FALSE]
  flow[[length(flow) + 1]] <<- data.frame(
    step = label, n = nrow(v1), bhatt_n = target, stringsAsFactors = FALSE)
  invisible(NULL)
}
keep("less interstitial lung disease or bronchiectasis",
     is.na(v1$ExclusionaryDisease), 10305L)
keep("less never-smokers",
     trimws(v1$cohort) != "Never smoked", 10198L)
keep("less unacceptable spirometry",
     !is.na(v1$FEV1_FVC_post), 10132L)

# Cohort membership past this point comes from Bhatt et al.'s per-subject
# classification file, not from a reconstruction of their CT-quality and
# minor-criteria completeness rules.
mdcopd_valid <- mdcopd[!is.na(mdcopd$MultiDim_COPD_score),
                       c("pid", "MultiDim_COPD_score")]
v1 <- merge(v1, mdcopd_valid, by = "pid")
flow[[length(flow) + 1]] <- data.frame(
  step = "less no valid MD-COPD classification (Bhatt ground truth)",
  n = nrow(v1), bhatt_n = 9416L, stringsAsFactors = FALSE)

# ---- Individual criteria (used by NoCT-MD-COPD and ESI-MD-COPD only) -----
# MD-COPD major/minor status comes from the file above; these criteria are
# needed to build the two CT-free classifications this paper is comparing
# against MD-COPD.
d_b <- within(v1, {
  gender <- factor(gender)
  race   <- factor(race)
  SmokCigNow <- factor(SmokCigNow)
  major_criterion <- FEV1_FVC_post < 0.70
  # The Fleischner scale runs none, trace, mild, moderate, confluent,
  # advanced destructive, so "mild or greater" is a score of 2, not 1.
  emph_yn <- CT_Visual_Emph_Severity >= 2
  wall_yn <- CT_Visual_Wall_Thickening == 2
  dysp_yn <- MMRCDyspneaScor >= 2
  qol_yn  <- SGRQ_scoreTotal >= 25
  cb_yn   <- Chronic_Bronchitis == 1
})
for (v in c("emph_yn", "wall_yn", "dysp_yn", "qol_yn", "cb_yn"))
  d_b[[v]][is.na(d_b[[v]])] <- FALSE

# ---- MD-COPD classification, from Bhatt's file directly ------------------
# score 1 = COPD-major, score 2 = COPD-minor, score 0 = neither; the last
# is split into AFL-only-noCOPD vs. noCOPD by the major criterion.
ORD <- c("noCOPD", "AFL-only-NoCOPD", "COPD-minor", "COPD-major")
d_b$bhatt_cls <- factor(
  ifelse(d_b$MultiDim_COPD_score == 1, "COPD-major",
  ifelse(d_b$MultiDim_COPD_score == 2, "COPD-minor",
  ifelse(d_b$major_criterion, "AFL-only-NoCOPD", "noCOPD"))),
  levels = ORD)

# Validate against Bhatt et al. Table 1. Only the score == 0 split is
# computed here (from FEV1_FVC_post); a mismatch would be a real finding,
# not a reconstruction gap, and should fail loudly.
got <- as.integer(table(d_b$bhatt_cls)[names(BHATT_MDCOPD_N)])
val <- data.frame(category = names(BHATT_MDCOPD_N), ours = got,
                  bhatt = as.integer(BHATT_MDCOPD_N),
                  diff = got - as.integer(BHATT_MDCOPD_N),
                  stringsAsFactors = FALSE)
stopifnot(all(val$diff == 0L))
write.csv(val, file.path(OUT_DIR, "mdcopd_validation.csv"), row.names = FALSE)

# ---- Baseline ESI (ours, not theirs) -------------------------------------
# Post-bronchodilator ESI at visit 1, one value per participant.
esi_v1 <- esi_raw[esi_raw$visitnum == 1 & esi_raw$PrePost == 1, ]
esi_v1 <- aggregate(list(ESI_v1post = esi_v1$ESI), list(pid = esi_v1$pid),
                    function(x) mean(x, na.rm = TRUE))
d_b <- merge(d_b, esi_v1, by = "pid")
d_b <- d_b[!is.na(d_b$ESI_v1post), , drop = FALSE]
flow[[length(flow) + 1]] <- data.frame(
  step = "less no baseline ESI", n = nrow(d_b), bhatt_n = NA_integer_,
  stringsAsFactors = FALSE)

cohort_flow <- do.call(rbind, flow)
# Every step for which Bhatt et al. published a count must match exactly,
# or the chain has diverged from theirs and everything downstream is on a
# different sample than intended.
chk <- cohort_flow[!is.na(cohort_flow$bhatt_n), ]
stopifnot(all(chk$n == chk$bhatt_n))
write.csv(cohort_flow, file.path(OUT_DIR, "cohort_flow.csv"), row.names = FALSE)

N_COHORT <- nrow(d_b)
cat(sprintf("Analytic cohort (with baseline ESI): n = %d\n", N_COHORT))
kable(cohort_flow, caption = "Cohort construction. `bhatt_n` is the published count for that step.")
kable(val, caption = "MD-COPD validated exactly against Bhatt et al. Table 1.")
```

## Section 1.1 -- Baseline characteristics by GOLD stratum (Supplemental Table S1)

Described on this cohort rather than carried over from the earlier analysis.
The exclusion chain removes never-smokers, so there is no never-smoker
stratum here.

```{r baseline-by-stratum}
# GOLD stratum from the COPDGene staging variable: -1 is PRISm, 0 is GOLD 0,
# 1-4 are GOLD 1-4. No "Never" level: never-smokers are excluded upstream.
stopifnot(!any(trimws(d_b$cohort) == "Never smoked"))
d_b$stratum <- factor(
  ifelse(d_b$finalgold_visit == -1, "PRISm",
         paste0("GOLD", d_b$finalgold_visit)),
  levels = c("PRISm", paste0("GOLD", 0:4)))

msd <- function(x) sprintf("%.1f (%.1f)", mean(x, na.rm = TRUE), sd(x, na.rm = TRUE))
pct <- function(x) sprintf("%.0f%%", 100 * mean(x, na.rm = TRUE))

base_row <- function(dd, label) data.frame(
  stratum      = label,
  n            = nrow(dd),
  age          = msd(dd$age_visit),
  pct_female   = pct(dd$gender == 2),
  pct_current  = pct(dd$SmokCigNow == 1),
  pack_years   = msd(dd$ATS_PackYears),
  BMI          = msd(dd$BMI),
  FEV1_pp      = msd(dd$FEV1pp_utah),
  FEV1_FVC     = sprintf("%.2f (%.2f)", mean(dd$FEV1_FVC_post, na.rm = TRUE),
                                        sd(dd$FEV1_FVC_post, na.rm = TRUE)),
  ESI          = sprintf("%.2f (%.2f)", mean(dd$ESI_v1post, na.rm = TRUE),
                                        sd(dd$ESI_v1post, na.rm = TRUE)),
  LAA950       = msd(dd$Insp_LAA950_total_Thirona),
  pct_mMRC2p   = pct(dd$dysp_yn),
  pct_SGRQ25p  = pct(dd$qol_yn),
  pct_CB       = pct(dd$cb_yn),
  stringsAsFactors = FALSE)

baseline <- do.call(rbind, c(
  lapply(levels(d_b$stratum), function(lv) base_row(d_b[d_b$stratum == lv, ], lv)),
  list(base_row(d_b, "Overall"))))
# The strata must partition the cohort, or a row is being double counted.
stopifnot(sum(baseline$n[baseline$stratum != "Overall"]) == nrow(d_b))
write.csv(baseline, file.path(OUT_DIR, "supp_baseline.csv"), row.names = FALSE)
kable(baseline, caption = "Baseline characteristics by GOLD stratum.")
```

## Section 1.2 -- ESI against quantitative CT (Supplemental Table S2)

Pearson correlations of ESI with quantitative CT emphysema, overall and within
stratum. The earlier analysis's supplemental table was a correlation matrix
*among the CT measures*, which does not support the statement it was cited for;
this computes what the text actually claims.

```{r esi-ct-correlations}
rpair <- function(dd, v) {
  ok <- stats::complete.cases(dd$ESI_v1post, dd[[v]])
  if (sum(ok) < 3) return(c(n = sum(ok), r = NA_real_))
  c(n = sum(ok), r = cor(dd$ESI_v1post[ok], dd[[v]][ok]))
}
corr_row <- function(dd, label) {
  a <- rpair(dd, "Insp_LAA950_total_Thirona")
  b <- rpair(dd, "PRM_pct_emphysema_Thirona")
  data.frame(stratum = label,
             n_LAA = as.integer(a["n"]), r_LAA = round(unname(a["r"]), 3),
             n_PRM = as.integer(b["n"]), r_PRM = round(unname(b["r"]), 3),
             stringsAsFactors = FALSE)
}
esi_ct <- do.call(rbind, c(
  lapply(levels(d_b$stratum), function(lv) corr_row(d_b[d_b$stratum == lv, ], lv)),
  list(corr_row(d_b, "All strata"))))
write.csv(esi_ct, file.path(OUT_DIR, "supp_esi_ct.csv"), row.names = FALSE)
kable(esi_ct, caption = "ESI against quantitative CT emphysema, by stratum.")
```

# Section 2 -- The four classifications

Fixed ratio, MD-COPD, NoCT-MD-COPD, and ESI-MD-COPD, applied to the same
`r nrow(d_b)` participants. All four use the same major criterion
(post-bronchodilator FEV1/FVC < 0.70), so no participant moves between the
airflow-limitation pair (AFL-only-noCOPD and COPD-major) and the
preserved-spirometry pair (noCOPD and COPD-minor).

```{r classifications}
# Handy shorthand columns used throughout the rest of the report.
d_b$afl <- d_b$major_criterion
d_b$ESI <- d_b$ESI_v1post
d_b$om  <- d_b$dysp_yn + d_b$qol_yn + d_b$cb_yn
O <- c("noCOPD", "AFL-only", "COPD-minor", "COPD-major")

# Generic classifier: given membership rules for the four categories,
# produce the factor.
label <- function(df, minor, aflo) factor(
  ifelse(df$afl & !aflo,   "COPD-major",
  ifelse(df$afl &  aflo,   "AFL-only",
  ifelse(!df$afl & minor,  "COPD-minor", "noCOPD"))), levels = O)

# S1  Fixed ratio.
d_b$S1 <- factor(ifelse(d_b$afl, "COPD", "noCOPD"),
                 levels = c("noCOPD", "COPD"))

# S2  MD-COPD (from Bhatt's file, level names normalised).
d_b$S2 <- factor(sub("AFL-only-NoCOPD", "AFL-only",
                     as.character(d_b$bhatt_cls)), levels = O)

# S3  NoCT-MD-COPD. Symptom criteria only; count threshold is fitted below.
s3 <- function(df, k) label(df, df$om >= k, df$om == 0)

# S4  ESI-MD-COPD. ESI is scored 0/1/2 by two thresholds; both thresholds
# and the count threshold are fitted below. This is the same 5-criterion,
# 2-threshold shape as the v15 draft rule; the values change.
s4 <- function(df, k, tl, th) {
  s <- ifelse(df$ESI >= th, 2, ifelse(df$ESI >= tl, 1, 0))
  label(df, (s + df$om) >= k, (s + df$om) == 0)
}
```

# Section 3 -- Fitting NoCT-MD-COPD and ESI-MD-COPD

Both need a count threshold (three-of-five doesn't apply once the criterion
count changes), and ESI-MD-COPD also needs the two ESI thresholds. Fitting
is outcome-blind: only the MD-COPD labels enter, never mortality or
exacerbations, so the risk profiles below are an independent check rather
than a restatement of the fit.

The objective is macro-averaged F1 across the four MD-COPD categories.
Cohen's kappa and mean per-category sensitivity were considered and rejected
because they fail in OPPOSITE directions on this problem: kappa lets a
schema over-diagnose COPD by assigning almost no one to AFL-only-noCOPD,
mean sensitivity encourages the reverse by ignoring false positives.
Macro-F1 penalises both.

```{r fitting}
macroF1 <- function(ref, est) mean(vapply(O, function(g) {
  tp <- sum(ref == g & est == g); if (tp == 0) return(0)
  2 * tp / (sum(est == g) + sum(ref == g))
}, numeric(1)))
mean_recall <- function(ref, est)
  mean(vapply(O, function(g) mean(est[ref == g] == g), numeric(1)))
kappa_full <- function(ref, est) {
  t <- table(ref, est); n <- sum(t)
  po <- sum(diag(t)) / n
  pe <- sum(rowSums(t) * colSums(t)) / n^2
  (po - pe) / (1 - pe)
}

# Search grids.
KS  <- 1:3
TL  <- seq(0.50, 4.00, by = 0.25)
TH  <- seq(1.00, 8.00, by = 0.50)
G4  <- do.call(rbind, lapply(2:4, function(k) {
  g <- expand.grid(t_low = TL, t_high = TH)
  g <- g[g$t_high >= g$t_low, ]
  cbind(k = k, g)
}))

fit3 <- function(df) {
  r <- df$S2
  KS[which.max(vapply(KS, function(k) macroF1(r, s3(df, k)), numeric(1)))]
}
fit4 <- function(df) {
  r <- df$S2
  scores <- vapply(seq_len(nrow(G4)), function(j)
    macroF1(r, s4(df, G4$k[j], G4$t_low[j], G4$t_high[j])), numeric(1))
  G4[which.max(scores), ]
}

k3   <- fit3(d_b)
p4   <- fit4(d_b)
d_b$S3 <- s3(d_b, k3)
d_b$S4 <- s4(d_b, p4$k, p4$t_low, p4$t_high)

cat(sprintf("NoCT-MD-COPD fitted: COPD-minor if >= %d of 3 symptom criteria\n", k3))
cat(sprintf("ESI-MD-COPD  fitted: COPD-minor if >= %d; ESI thresholds %.2f / %.2f\n",
            p4$k, p4$t_low, p4$t_high))
```

## Section 3.1 -- What each candidate objective selects

Reported as a Results finding rather than a Methods choice, since kappa
and mean sensitivity fail in different directions and describing them as
failing the same way would be wrong.

```{r objective-selection}
obj_rows <- list()
for (nm in c("kappa", "mean_sensitivity", "macro_F1")) {
  scorer <- switch(nm,
    kappa            = function(e) kappa_full(d_b$S2, e),
    mean_sensitivity = function(e) mean_recall(d_b$S2, e),
    macro_F1         = function(e) macroF1(d_b$S2, e))
  v <- vapply(seq_len(nrow(G4)), function(j)
    scorer(s4(d_b, G4$k[j], G4$t_low[j], G4$t_high[j])), numeric(1))
  b <- G4[which.max(v), ]
  est <- s4(d_b, b$k, b$t_low, b$t_high)
  obj_rows[[length(obj_rows) + 1]] <- data.frame(
    objective = nm, k = b$k, t_low = b$t_low, score = max(v),
    n_aflonly = sum(est == "AFL-only"),
    n_major   = sum(est == "COPD-major"),
    stringsAsFactors = FALSE)
}
objsel <- do.call(rbind, obj_rows)
objsel$ref_aflonly <- sum(d_b$S2 == "AFL-only")
objsel$ref_major   <- sum(d_b$S2 == "COPD-major")
# The two rejected objectives must err in opposite directions on
# AFL-only-noCOPD or the Results sentence describing them is wrong.
stopifnot(objsel$n_aflonly[objsel$objective == "kappa"]
            < objsel$ref_aflonly[1],
          objsel$n_aflonly[objsel$objective == "mean_sensitivity"]
            > objsel$ref_aflonly[1])
write.csv(objsel, file.path(OUT_DIR, "objective_selection.csv"), row.names = FALSE)
kable(objsel, caption = "What each candidate objective selects. Kappa and mean sensitivity fail in opposite directions.")
```

## Section 3.2 -- Repeated stratified cross-validation

Five repeats of five folds, stratified on the MD-COPD categories, with the
thresholds refitted inside every training fold.

```{r cv-fit}
set.seed(CV_SEED)
cv_rows <- list()
for (rep_i in seq_len(CV_REPEATS)) {
  fold <- integer(nrow(d_b))
  for (lv in O) {
    ix <- which(d_b$S2 == lv)
    fold[ix] <- sample(rep_len(seq_len(CV_FOLDS), length(ix)))
  }
  for (kf in seq_len(CV_FOLDS)) {
    tr <- d_b[fold != kf, ]
    te <- d_b[fold == kf, ]
    rt <- te$S2
    a <- fit3(tr); b <- fit4(tr)
    cv_rows[[length(cv_rows) + 1]] <- data.frame(
      rep = rep_i, fold = kf,
      k3 = a,  f3 = macroF1(rt, s3(te, a)),
      k4 = b$k, tl = b$t_low, th = b$t_high,
      f4 = macroF1(rt, s4(te, b$k, b$t_low, b$t_high)),
      f_draft = macroF1(rt, s4(te, 3, 1.00, 2.50)),
      stringsAsFactors = FALSE)
  }
}
cv <- do.call(rbind, cv_rows)
write.csv(cv, file.path(OUT_DIR, "cv_schema_fits.csv"), row.names = FALSE)

dd <- cv$f4 - cv$f3
cat(sprintf("Held-out macro-F1: NoCT-MD-COPD %.4f | ESI-MD-COPD %.4f | v15 draft %.4f\n",
            mean(cv$f3), mean(cv$f4), mean(cv$f_draft)))
cat(sprintf("ESI-MD-COPD minus NoCT-MD-COPD: %+.4f (%+.4f to %+.4f across folds)\n",
            mean(dd), quantile(dd, .025), quantile(dd, .975)))

write.csv(data.frame(
  schema = c("S3", "S4", "S4_v15draft"),
  k      = c(k3, p4$k, 3),
  t_low  = c(NA, p4$t_low, 1.00),
  t_high = c(NA, p4$t_high, 2.50),
  macroF1_insample = c(macroF1(d_b$S2, s3(d_b, k3)),
                       macroF1(d_b$S2, s4(d_b, p4$k, p4$t_low, p4$t_high)),
                       macroF1(d_b$S2, s4(d_b, 3, 1.00, 2.50))),
  macroF1_heldout  = c(mean(cv$f3), mean(cv$f4), mean(cv$f_draft)),
  stringsAsFactors = FALSE),
  file.path(OUT_DIR, "schema_fit.csv"), row.names = FALSE)
write.csv(data.frame(diff_mean = mean(dd),
                     diff_lo = unname(quantile(dd, .025)),
                     diff_hi = unname(quantile(dd, .975)),
                     n_folds = nrow(cv)),
          file.path(OUT_DIR, "schema_fit_cv_diff.csv"), row.names = FALSE)
writeLines(c(sprintf("n_cohort=%d", nrow(d_b)),
             sprintf("n_afl=%d",    sum(d_b$afl)),
             sprintf("n_noafl=%d",  sum(!d_b$afl))),
           file.path(OUT_DIR, "cohort.txt"))
```

# Section 4 -- Category counts and reclassification

Save per-subject schema labels so figures and tables downstream don't
re-derive them, and write the flow of reclassification against MD-COPD as
its own artifact.

```{r labels-and-flow}
saveRDS(d_b[, c("pid", "S1", "S2", "S3", "S4")],
        file.path(OUT_DIR, "schema_labels.rds"))

lab_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) {
  tb <- table(d_b[[s]])
  for (g in names(tb)) lab_rows[[length(lab_rows) + 1]] <- data.frame(
    schema = s, category = g, n = as.integer(tb[[g]]),
    stringsAsFactors = FALSE)
}
schema_labels <- do.call(rbind, lab_rows)
write.csv(schema_labels, file.path(OUT_DIR, "schema_labels.csv"), row.names = FALSE)
kable(schema_labels, caption = "Category counts under each classification.")

# Reclassification counts quoted directly in the Results, plus the share of
# COPD-major that qualifies only through CT findings (the reason a CT-free
# schema can lose that category at all).
reclass <- do.call(rbind, lapply(c("S3", "S4"), function(s) data.frame(
  schema           = s,
  major_to_aflonly = sum(d_b$S2 == "COPD-major" & d_b[[s]] == "AFL-only"),
  minor_to_nocopd  = sum(d_b$S2 == "COPD-minor" & d_b[[s]] == "noCOPD"),
  nocopd_to_minor  = sum(d_b$S2 == "noCOPD"     & d_b[[s]] == "COPD-minor"),
  aflonly_kept     = sum(d_b$S2 == "AFL-only"   & d_b[[s]] == "AFL-only"),
  concordant       = sum(d_b$S2 == d_b[[s]]),
  stringsAsFactors = FALSE)))
n_major   <- sum(d_b$S2 == "COPD-major")
ct_only   <- sum(d_b$S2 == "COPD-major" & d_b$om == 0)
reclass$n_major       <- n_major
reclass$ct_only_major <- ct_only
reclass$ct_only_pct   <- 100 * ct_only / n_major
write.csv(reclass, file.path(OUT_DIR, "reclassification.csv"), row.names = FALSE)

# Full cross-classification of each CT-free schema against MD-COPD, long
# form so the supplement table can be built without recomputing anything.
xtab <- do.call(rbind, lapply(c("S3", "S4"), function(s) {
  tb <- table(factor(d_b[[s]], levels = O), factor(d_b$S2, levels = O))
  do.call(rbind, lapply(O, function(rw) data.frame(
    schema = s, row_cat = rw, col_cat = O,
    n = as.integer(tb[rw, O]), stringsAsFactors = FALSE)))
}))
stopifnot(sum(xtab$n) == 2 * nrow(d_b))
write.csv(xtab, file.path(OUT_DIR, "crossclass.csv"), row.names = FALSE)
```

# Section 5 -- Outcome data

Attach mortality, cause of death, and long-term follow-up exacerbations,
requiring complete covariate data for the adjusted models.

```{r outcome-data}
COV <- "age_visit + gender + race + SmokCigNow + ATS_PackYears + BMI"

mort <- d_b %>%
  inner_join(vs  %>% dplyr::select(pid, vital_status, days_followed), by = "pid") %>%
  left_join (cod %>% dplyr::select(pid, UCD_Resp), by = "pid") %>%
  mutate(ev_resp = ifelse(vital_status == 1 & !is.na(UCD_Resp) & UCD_Resp == 1, 1, 0),
         py      = days_followed / 365.25) %>%
  filter(complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))

# Prior-year exacerbation frequency, per the source MD-COPD paper.
prior <- phe_raw %>%
  filter(visitnum == 1) %>%
  transmute(pid, prior_exac = suppressWarnings(as.numeric(Exacerbation_Frequency))) %>%
  distinct(pid, .keep_all = TRUE)

exa <- d_b %>%
  inner_join(ex_raw %>% dplyr::select(pid, Total_Exacerbations, Years_Followed), by = "pid") %>%
  left_join (prior, by = "pid") %>%
  filter(!is.na(Total_Exacerbations), Years_Followed > 0, !is.na(prior_exac),
         complete.cases(age_visit, gender, race, SmokCigNow, ATS_PackYears, BMI))

cat(sprintf("Mortality n = %d (%d deaths, %d respiratory)\n",
            nrow(mort), sum(mort$vital_status), sum(mort$ev_resp)))
cat(sprintf("Exacerbation n = %d\n", nrow(exa)))
```

# Section 6 -- MD-COPD vs. the fixed ratio (nested LR test)

Fixed-ratio COPD is exactly the AFL-only-noCOPD and COPD-major categories
of MD-COPD, so the fixed-ratio model is nested in the four-category model
and the two are compared by likelihood ratio test on 2 degrees of freedom
rather than by comparing summary measures.

```{r gate-fixedratio}
d_b$bhatt_grp <- d_b$bhatt_cls
mort$fixed_ratio <- factor(ifelse(mort$afl, "COPD", "noCOPD"),
                           levels = c("noCOPD", "COPD"))
mort$bhatt_grp   <- d_b$bhatt_cls[match(mort$pid, d_b$pid)]
exa$fixed_ratio  <- factor(ifelse(exa$afl,  "COPD", "noCOPD"),
                           levels = c("noCOPD", "COPD"))
exa$bhatt_grp    <- d_b$bhatt_cls[match(exa$pid,  d_b$pid)]
stopifnot(all((mort$fixed_ratio == "COPD") ==
              (mort$bhatt_grp %in% c("AFL-only-NoCOPD", "COPD-major"))))

cox_pair <- function(ev, label) {
  f0 <- as.formula(sprintf("Surv(days_followed/365.25, %s) ~ fixed_ratio + %s", ev, COV))
  f1 <- as.formula(sprintf("Surv(days_followed/365.25, %s) ~ bhatt_grp   + %s", ev, COV))
  m0 <- coxph(f0, data = mort); m1 <- coxph(f1, data = mort)
  lr <- 2 * (m1$loglik[2] - m0$loglik[2])
  df <- length(coef(m1)) - length(coef(m0))
  c0 <- summary(m0)$concordance; c1 <- summary(m1)$concordance
  cat(sprintf("--- %s ---\n", label))
  cat(sprintf("  fixed ratio : C = %.4f (SE %.4f)\n", c0[1], c0[2]))
  cat(sprintf("  MD-COPD     : C = %.4f (SE %.4f)   gain %+.4f\n",
              c1[1], c1[2], c1[1] - c0[1]))
  cat(sprintf("  LR chi-square = %.1f on %d df, p = %s\n\n",
              lr, df, format.pval(pchisq(lr, df, lower.tail = FALSE),
                                  digits = 3, eps = 1e-16)))
  data.frame(outcome = label, c_fixedratio = c0[1], c_mdcopd = c1[1],
             lrt_chisq = lr, lrt_df = df,
             lrt_p = pchisq(lr, df, lower.tail = FALSE),
             stringsAsFactors = FALSE)
}
gate_rows <- list()
gate_rows[[1]] <- cox_pair("vital_status", "ALL-CAUSE MORTALITY")
gate_rows[[2]] <- cox_pair("ev_resp",      "RESPIRATORY MORTALITY")

# Exacerbations: negative binomial with the same nesting.
n0 <- glm.nb(as.formula(paste("Total_Exacerbations ~ fixed_ratio +", COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa)
n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~ bhatt_grp +",  COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa)
lr <- 2 * (as.numeric(logLik(n1)) - as.numeric(logLik(n0)))
df <- length(coef(n1)) - length(coef(n0))
cat("--- EXACERBATIONS ---\n")
cat(sprintf("  fixed ratio : AIC %.1f\n  MD-COPD     : AIC %.1f   improvement %.1f\n",
            AIC(n0), AIC(n1), AIC(n0) - AIC(n1)))
cat(sprintf("  LR chi-square = %.1f on %d df, p = %s\n",
            lr, df, format.pval(pchisq(lr, df, lower.tail = FALSE),
                                digits = 3, eps = 1e-16)))
gate_rows[[3]] <- data.frame(outcome = "EXACERBATIONS",
  c_fixedratio = NA_real_, c_mdcopd = NA_real_,
  lrt_chisq = lr, lrt_df = df,
  lrt_p = pchisq(lr, df, lower.tail = FALSE),
  stringsAsFactors = FALSE)

write.csv(do.call(rbind, gate_rows),
          file.path(OUT_DIR, "gate_fixedratio.csv"), row.names = FALSE)
write.csv(data.frame(aic_fixedratio = AIC(n0), aic_mdcopd = AIC(n1)),
          file.path(OUT_DIR, "gate_exac_aic.csv"), row.names = FALSE)
```

# Section 7 -- Adjusted risk within each classification's own categories

Cox proportional hazards models for mortality and negative binomial models
with a follow-up-year offset for exacerbations. Each estimate is expressed
against its OWN classification's noCOPD category, so a CT-free
classification's AFL-only-noCOPD group is compared to its own noCOPD group
rather than to MD-COPD's.

```{r adjusted-risk}
NAMES <- c(S1 = "Fixed ratio",
           S2 = "MD-COPD (reference)",
           S3 = "NoCT-MD-COPD",
           S4 = "ESI-MD-COPD")

risk_rows <- list(); disc_rows <- list()
for (s in c("S1", "S2", "S3", "S4")) {
  m1 <- coxph(as.formula(sprintf("Surv(py, vital_status) ~ %s + %s", s, COV)), data = mort)
  m2 <- coxph(as.formula(sprintf("Surv(py, ev_resp) ~ %s + %s",      s, COV)), data = mort)
  n1 <- glm.nb(as.formula(paste("Total_Exacerbations ~", s, "+", COV,
         "+ prior_exac + offset(log(Years_Followed))")), data = exa)
  cm <- summary(m1)$conf.int; cr <- summary(m2)$conf.int; ce <- summary(n1)$coef
  for (g in levels(d_b[[s]])) {
    r <- paste0(s, g); ix <- mort[[s]] == g
    gv <- function(M, j) if (r %in% rownames(M)) M[r, j] else NA_real_
    risk_rows[[length(risk_rows) + 1]] <- data.frame(
      schema = s, category = g,
      n_mort = sum(ix), n_exac = sum(exa[[s]] == g),
      n_cohort = sum(d_b[[s]] == g), deaths = sum(mort$vital_status[ix]),
      all_HR  = gv(cm, 1), all_LCI  = gv(cm, 3), all_UCI  = gv(cm, 4),
      resp_HR = gv(cr, 1), resp_LCI = gv(cr, 3), resp_UCI = gv(cr, 4),
      exac_IRR = if (r %in% rownames(ce)) exp(ce[r, 1]) else NA_real_,
      exac_LCI = if (r %in% rownames(ce)) exp(ce[r, 1] - 1.96 * ce[r, 2]) else NA_real_,
      exac_UCI = if (r %in% rownames(ce)) exp(ce[r, 1] + 1.96 * ce[r, 2]) else NA_real_,
      stringsAsFactors = FALSE)
  }
  disc_rows[[length(disc_rows) + 1]] <- data.frame(
    schema = s,
    c_allcause = summary(m1)$concordance[1],
    c_resp     = summary(m2)$concordance[1],
    exac_AIC   = AIC(n1), stringsAsFactors = FALSE)
}
risk <- do.call(rbind, risk_rows)
# Mortality and exacerbation models are fitted on different participant
# sets. Reporting one n per category for all three outcomes would overstate
# exacerbation denominators.
stopifnot(sum(risk$n_exac[risk$schema == "S2"]) == nrow(exa),
          sum(risk$n_mort[risk$schema == "S2"]) == nrow(mort),
          sum(risk$n_cohort[risk$schema == "S2"]) == nrow(d_b))
write.csv(risk, file.path(OUT_DIR, "schema_risk.csv"), row.names = FALSE)
write.csv(do.call(rbind, disc_rows),
          file.path(OUT_DIR, "schema_discrimination.csv"), row.names = FALSE)
kable(risk[risk$schema == "S2", c("category","n_mort","all_HR","all_LCI","all_UCI",
                                   "resp_HR","resp_LCI","resp_UCI",
                                   "exac_IRR","exac_LCI","exac_UCI")],
      caption = "Adjusted risk under MD-COPD.", digits = 3)
```

# Section 8 -- Crude rate ratios beside the adjusted estimates

An adjusted estimate is what the classification contributes beyond the
covariates; the crude ratio is what participants in the category actually
experienced. Both belong in the results.

Interval convention: a resample in which a category contributes no events
is retained as a legitimate draw with a rate ratio of zero (dropping such
draws truncates the interval from below and can push a lower bound above
1 on the strength of a single event). Only resamples in which the
REFERENCE contributes no events leave the ratio undefined and are cut.

```{r crude-rr}
set.seed(CV_SEED)

crude_ratio <- function(sch, ev, py, dat) {
  g   <- dat[[sch]]
  num <- rowsum(dat[[ev]], g, reorder = FALSE)
  den <- rowsum(dat[[py]], g, reorder = FALSE)
  r   <- setNames(as.vector(num) / as.vector(den), rownames(num))
  ref <- r[["noCOPD"]]
  if (!is.finite(ref) || ref <= 0)
    return(setNames(rep(NA_real_, length(r)), names(r)))
  r / ref
}

crude_rows <- list()
for (spec in list(list("all",  "vital_status",        "py",             mort),
                  list("resp", "ev_resp",             "py",             mort),
                  list("exac", "Total_Exacerbations", "Years_Followed", exa))) {
  key <- spec[[1]]; ev <- spec[[2]]; py <- spec[[3]]; dat <- spec[[4]]
  boot <- vector("list", CRUDE_B)
  for (b in seq_len(CRUDE_B)) {
    idx <- sample.int(nrow(dat), nrow(dat), replace = TRUE)
    dd  <- dat[idx, , drop = FALSE]
    boot[[b]] <- lapply(c("S1","S2","S3","S4"),
                        function(s) crude_ratio(s, ev, py, dd))
  }
  for (si in seq_along(c("S1","S2","S3","S4"))) {
    s   <- c("S1","S2","S3","S4")[si]
    est <- crude_ratio(s, ev, py, dat)
    for (g in names(est)) {
      v <- vapply(boot, function(bb) {
        z <- bb[[si]]; if (g %in% names(z)) z[[g]] else NA_real_
      }, numeric(1))
      v <- v[is.finite(v)]
      crude_rows[[length(crude_rows) + 1]] <- data.frame(
        schema = s, category = g, outcome = key,
        rr = unname(est[[g]]),
        lo = if (length(v) > 1) unname(quantile(v, .025)) else NA_real_,
        hi = if (length(v) > 1) unname(quantile(v, .975)) else NA_real_,
        B_eff = sum(v > 0), stringsAsFactors = FALSE)
    }
  }
}
crude <- do.call(rbind, crude_rows)
# The reference category's crude ratio is 1 by construction. Drift means
# the numerator and denominator were computed off different rows.
stopifnot(all(abs(crude$rr[crude$category == "noCOPD"] - 1) < 1e-12))
write.csv(crude, file.path(OUT_DIR, "schema_crude.csv"), row.names = FALSE)
```

# Section 9 -- Concordant and discordant classification (MD-COPD vs ESI-MD-COPD)

Pairwise, as in the earlier analysis. NoCT-MD-COPD is not carried into this
comparison: it underdiagnoses a group with raised mortality, which the
preceding sections establish, so the question of what happens to individually
discordant participants is asked of ESI-MD-COPD alone.

Binary COPD status under each classification is cross-classified into four
groups. Restricted to preserved spirometry, where the two classifications can
actually disagree about a diagnosis rather than about which COPD category
applies.

```{r discordance-groups}
copd_of <- function(v) v %in% c("COPD-minor", "COPD-major")
mk_discord <- function(dd) factor(
  ifelse( copd_of(dd$S2) &  copd_of(dd$S4), "Both-COPD",
  ifelse( copd_of(dd$S2) & !copd_of(dd$S4), "CT-only-COPD",
  ifelse(!copd_of(dd$S2) &  copd_of(dd$S4), "ESI-only-COPD", "Both-noCOPD"))),
  levels = c("Both-noCOPD", "ESI-only-COPD", "CT-only-COPD", "Both-COPD"))

d_b$discord  <- mk_discord(d_b)
mort$discord <- mk_discord(mort)
exa$discord  <- mk_discord(exa)

# Preserved spirometry only.
d_ps    <- d_b[!d_b$afl, ]
mort_ps <- mort[!mort$afl, ]
exa_ps  <- exa[!exa$afl, ]
cat(sprintf("preserved-spirometry cohort: %d (mortality %d, exacerbation %d)\n\n",
            nrow(d_ps), nrow(mort_ps), nrow(exa_ps)))
print(table(d_ps$discord))

# CT-only-COPD is what ESI misses; ESI-only-COPD is what it adds.
disc_n <- as.data.frame(table(d_ps$discord), stringsAsFactors = FALSE)
names(disc_n) <- c("group", "n")
write.csv(disc_n, file.path(OUT_DIR, "discord_counts.csv"), row.names = FALSE)
```

```{r discordance-rates}
# Observed rates per 100 person-years, and prior exacerbation burden, which
# the earlier analysis found explains much of the gradient in future
# exacerbations.
rate100 <- function(ev, py) 100 * sum(ev) / sum(py)
rows <- lapply(levels(d_ps$discord), function(g) {
  m <- mort_ps[mort_ps$discord == g, ]
  e <- exa_ps [exa_ps$discord  == g, ]
  data.frame(
    group          = g,
    n              = sum(d_ps$discord == g),
    n_mort         = nrow(m),
    deaths         = sum(m$vital_status),
    rate_all_100py = rate100(m$vital_status, m$py),
    rate_resp_100py= rate100(m$ev_resp,      m$py),
    n_exac         = nrow(e),
    rate_exac_100py= rate100(e$Total_Exacerbations, e$Years_Followed),
    prior_exac_mean= mean(e$prior_exac, na.rm = TRUE),
    stringsAsFactors = FALSE)
})
disc_rates <- do.call(rbind, rows)
# Rates must be ordered by increasing severity of the group definition for the
# narrative to hold; assert nothing, but surface it.
write.csv(disc_rates, file.path(OUT_DIR, "discord_rates.csv"), row.names = FALSE)
kable(disc_rates, digits = 2,
      caption = "Observed rates per 100 person-years by cross-classification group.")
```

```{r discordance-adjusted}
# Adjusted estimates against Both-noCOPD.
cm <- coxph(as.formula(paste("Surv(py, vital_status) ~ discord +", COV)), data = mort_ps)
cr <- coxph(as.formula(paste("Surv(py, ev_resp) ~ discord +",      COV)), data = mort_ps)
ne <- glm.nb(as.formula(paste("Total_Exacerbations ~ discord +", COV,
       "+ prior_exac + offset(log(Years_Followed))")), data = exa_ps)
sm <- summary(cm)$conf.int; sr <- summary(cr)$conf.int; se <- summary(ne)$coef

# A group with no events of a given type has no estimable hazard ratio. The
# Cox fit still returns a number: for CT-only-COPD, which has zero respiratory
# deaths, it returns 3e-07 with an interval of 0 to Inf. That is a monotone
# likelihood, not an effect size, and printing it as one would be worse than
# printing nothing. Such cells are set NA and the event count is carried
# alongside so the reason is visible in the artifact rather than inferred.
n_ev <- function(g, col) sum(mort_ps[[col]][mort_ps$discord == g])
adj <- do.call(rbind, lapply(levels(d_ps$discord)[-1], function(g) {
  r <- paste0("discord", g)
  gv <- function(M, j) if (r %in% rownames(M)) M[r, j] else NA_real_
  d_all  <- n_ev(g, "vital_status")
  d_resp <- n_ev(g, "ev_resp")
  est <- function(v, ok) if (ok) v else NA_real_
  data.frame(group = g,
    deaths_all = d_all, deaths_resp = d_resp,
    all_HR  = est(gv(sm,1), d_all  > 0), all_LCI = est(gv(sm,3), d_all  > 0),
    all_UCI = est(gv(sm,4), d_all  > 0),
    resp_HR = est(gv(sr,1), d_resp > 0), resp_LCI = est(gv(sr,3), d_resp > 0),
    resp_UCI= est(gv(sr,4), d_resp > 0),
    exac_IRR = if (r %in% rownames(se)) exp(se[r,1]) else NA_real_,
    exac_LCI = if (r %in% rownames(se)) exp(se[r,1]-1.96*se[r,2]) else NA_real_,
    exac_UCI = if (r %in% rownames(se)) exp(se[r,1]+1.96*se[r,2]) else NA_real_,
    stringsAsFactors = FALSE)
}))
# Any surviving estimate must be finite and positive; a degenerate fit that
# slipped through the event-count guard would still be caught here.
.fin <- function(v) all(is.na(v) | (is.finite(v) & v > 0))
stopifnot(.fin(adj$all_HR), .fin(adj$resp_HR), .fin(adj$exac_IRR),
          .fin(adj$resp_UCI))
write.csv(adj, file.path(OUT_DIR, "discord_adjusted.csv"), row.names = FALSE)
kable(adj, digits = 2, caption = "Adjusted estimates against Both-noCOPD.")
```

# Section 10 -- Do the classifications differ? Paired bootstrap

The preceding sections report each classification's estimates side by side.
Reading a difference off two intervals is not a test of one, so this resamples
participants and refits BOTH classifications within every resample, taking the
difference of the log estimates for the same category. Pairing is what makes
the comparison valid: both estimates come from the same people in each draw,
so the between-classification correlation is preserved rather than assumed
away.

Each outcome is fitted in its own `tryCatch`. A warning from a sparse
respiratory fit must not discard the valid all-cause fits for that resample,
because dropped resamples are not missing at random: they are precisely the
ones with few respiratory events. Effective B is therefore tracked per outcome.

```{r paired-bootstrap}
B_DIFF <- 1000L
CACHE  <- file.path(OUT_DIR, "_cache"); dir.create(CACHE, showWarnings = FALSE)
CPATH  <- file.path(CACHE, "paired_bootstrap.rds")
CATS   <- c("AFL-only", "COPD-minor", "COPD-major")

# The fingerprint carries the model formulas, not just row counts, so any
# change to an outcome, covariate or offset invalidates the cache whether or
# not anyone remembers to bump a version.
.fml <- c(paste("Surv(py, vital_status) ~ S +", COV),
          paste("Surv(py, ev_resp) ~ S +",      COV),
          paste("Total_Exacerbations ~ S +", COV,
                "+ prior_exac + offset(log(Years_Followed))"))
.want <- list(n_mort = nrow(mort), n_exac = nrow(exa), B = B_DIFF,
              deaths = sum(mort$vital_status), resp = sum(mort$ev_resp),
              cats = paste(CATS, collapse = ","), fml = paste(.fml, collapse = "|"),
              version = 1L)
.cached <- if (file.exists(CPATH)) readRDS(CPATH) else NULL
if (!is.null(.cached) && !identical(.cached$fingerprint, .want)) {
  message("paired-bootstrap cache stale; recomputing")
  .cached <- NULL
}

if (!is.null(.cached)) {
  bootdiff <- .cached
  cat(sprintf("loaded cached paired bootstrap (B effective: all %d, resp %d, exac %d)\n",
              bootdiff$B_eff[["all"]], bootdiff$B_eff[["resp"]], bootdiff$B_eff[["exac"]]))
} else {
  set.seed(CV_SEED)
  mk <- function() matrix(NA_real_, nrow = B_DIFF, ncol = length(CATS),
                          dimnames = list(NULL, CATS))
  store <- list(all_S2 = mk(), all_S4 = mk(), resp_S2 = mk(), resp_S4 = mk(),
                exac_S2 = mk(), exac_S4 = mk())
  lg <- function(fit, term, nb = FALSE) {
    co <- if (nb) summary(fit)$coef else summary(fit)$coef
    if (!term %in% rownames(co)) return(NA_real_)
    unname(co[term, 1])
  }
  fit_cox <- function(dat, ev) tryCatch({
    lhs <- sprintf("Surv(py, %s)", ev)
    list(S2 = coxph(as.formula(paste(lhs, "~ S2 +", COV)), data = dat),
         S4 = coxph(as.formula(paste(lhs, "~ S4 +", COV)), data = dat))
  }, error = function(e) NULL, warning = function(w) NULL)
  fit_nb <- function(dat) tryCatch({
    rhs <- paste(COV, "+ prior_exac + offset(log(Years_Followed))")
    list(S2 = glm.nb(as.formula(paste("Total_Exacerbations ~ S2 +", rhs)), data = dat),
         S4 = glm.nb(as.formula(paste("Total_Exacerbations ~ S4 +", rhs)), data = dat))
  }, error = function(e) NULL, warning = function(w) NULL)

  B_eff <- c(all = 0L, resp = 0L, exac = 0L)
  for (b in seq_len(B_DIFF)) {
    im <- sample.int(nrow(mort), nrow(mort), replace = TRUE); dm <- mort[im, ]
    ie <- sample.int(nrow(exa),  nrow(exa),  replace = TRUE); de <- exa [ie, ]
    fa <- fit_cox(dm, "vital_status"); fr <- fit_cox(dm, "ev_resp"); fe <- fit_nb(de)
    if (!is.null(fa)) { for (g in CATS) {
        store$all_S2[b, g] <- lg(fa$S2, paste0("S2", g))
        store$all_S4[b, g] <- lg(fa$S4, paste0("S4", g)) }
      B_eff["all"] <- B_eff["all"] + 1L }
    if (!is.null(fr)) { for (g in CATS) {
        store$resp_S2[b, g] <- lg(fr$S2, paste0("S2", g))
        store$resp_S4[b, g] <- lg(fr$S4, paste0("S4", g)) }
      B_eff["resp"] <- B_eff["resp"] + 1L }
    if (!is.null(fe)) { for (g in CATS) {
        store$exac_S2[b, g] <- lg(fe$S2, paste0("S2", g), nb = TRUE)
        store$exac_S4[b, g] <- lg(fe$S4, paste0("S4", g), nb = TRUE) }
      B_eff["exac"] <- B_eff["exac"] + 1L }
  }
  bootdiff <- list(store = store, B_eff = B_eff, fingerprint = .want)
  saveRDS(bootdiff, CPATH)
  cat(sprintf("paired bootstrap done (B effective: all %d, resp %d, exac %d)\n",
              B_eff[["all"]], B_eff[["resp"]], B_eff[["exac"]]))
}

# Percentile interval on the difference in log estimates, and a two-sided
# bootstrap p from the proportion of draws on the other side of zero.
diff_row <- function(key, label, g) {
  d <- bootdiff$store[[paste0(key, "_S4")]][, g] -
       bootdiff$store[[paste0(key, "_S2")]][, g]
  d <- d[is.finite(d)]
  if (!length(d)) return(NULL)
  p <- 2 * min(mean(d <= 0), mean(d >= 0)); p <- min(1, p)
  data.frame(outcome = label, category = g,
             ratio_S4_over_S2 = exp(mean(d)),
             lo = exp(unname(quantile(d, .025))),
             hi = exp(unname(quantile(d, .975))),
             p_two_sided = p, B_eff = length(d), stringsAsFactors = FALSE)
}
diffs <- do.call(rbind, unlist(recursive = FALSE, lapply(
  list(c("all","all-cause mortality"), c("resp","respiratory mortality"),
       c("exac","exacerbations")),
  function(o) lapply(CATS, function(g) diff_row(o[1], o[2], g)))))
write.csv(diffs, file.path(OUT_DIR, "schema_diff_bootstrap.csv"), row.names = FALSE)
kable(diffs, digits = 3,
      caption = "ESI-MD-COPD versus MD-COPD, same category, paired resamples. A ratio of 1 means the two classifications assign the same effect size.")
```

# Section 9 -- Self-check (canonical values echoed for drift detection)

Each row prints the freshly-computed value alongside the value cited in
the manuscript (`expected`). Rerunning this report on the cluster should
reproduce the `expected` column exactly (within rounding).

```{r self-check}
selfcheck <- function(name, computed, expected, tol = 0.02) {
  ok <- if (is.numeric(computed) && is.numeric(expected)) {
          isTRUE(abs(computed - expected) < tol)
        } else identical(computed, expected)
  data.frame(check = name,
             computed = if (is.numeric(computed)) sprintf("%.3f", computed)
                        else as.character(computed),
             expected = if (is.numeric(expected)) sprintf("%.3f", expected)
                        else as.character(expected),
             status = ifelse(ok, "PASS", "FAIL"),
             stringsAsFactors = FALSE)
}

R2 <- function(s, cat, fld) risk[[fld]][risk$schema == s & risk$category == cat]
CR <- function(s, cat, out, fld) crude[[fld]][crude$schema == s & crude$category == cat & crude$outcome == out]
GT <- function(o, fld) {
  g <- do.call(rbind, gate_rows); g[[fld]][g$outcome == o]
}
DS <- function(s, fld) {
  d <- do.call(rbind, disc_rows); d[[fld]][d$schema == s]
}

# Canonical values are those cited in the manuscript prose. The point of
# this table is to detect drift on re-run: if any row goes FAIL, an
# expected number must be updated deliberately (with a corresponding
# manuscript edit) rather than silently changed. Every value is loaded
# from the already-written artifact so this block reflects what downstream
# reads, not what the in-memory objects happened to hold.

# MD-COPD counts BEFORE the baseline-ESI restriction are what should
# reproduce Bhatt Table 1 exactly. mdcopd_validation.csv captures that
# state; d_b at self-check time has already been ESI-restricted.
mdval <- read.csv(file.path(OUT_DIR, "mdcopd_validation.csv"),
                  stringsAsFactors = FALSE)
MDV <- function(cat) mdval$ours[mdval$category == cat]

checks <- list(
  # ---- Bhatt Table 1 reproduction (before ESI restriction) --------------
  selfcheck("MD-COPD noCOPD n vs Bhatt 4,439",
            MDV("noCOPD"),          4439L),
  selfcheck("MD-COPD AFL-only-noCOPD n vs Bhatt 282",
            MDV("AFL-only-NoCOPD"), 282L),
  selfcheck("MD-COPD COPD-minor n vs Bhatt 811",
            MDV("COPD-minor"),      811L),
  selfcheck("MD-COPD COPD-major n vs Bhatt 3,884",
            MDV("COPD-major"),      3884L),

  # ---- Analytic cohort after ESI restriction ---------------------------
  selfcheck("analytic cohort n (with baseline ESI)",
            as.integer(nrow(d_b)),  9240L, tol = 0.5),

  # ---- Nested LR gate: MD-COPD vs. fixed ratio (2 df) ------------------
  selfcheck("all-cause LR chi-square, MD-COPD over fixed ratio",
            GT("ALL-CAUSE MORTALITY", "lrt_chisq"),   122.1, tol = 0.1),
  selfcheck("respiratory LR chi-square, MD-COPD over fixed ratio",
            GT("RESPIRATORY MORTALITY", "lrt_chisq"),  76.1, tol = 0.1),
  selfcheck("exacerbation LR chi-square, MD-COPD over fixed ratio",
            GT("EXACERBATIONS", "lrt_chisq"),         158.9, tol = 0.1),

  # ---- Fitted CT-free rules --------------------------------------------
  selfcheck("NoCT-MD-COPD count threshold", k3,       2L),
  selfcheck("ESI-MD-COPD count threshold",  p4$k,     2L),
  selfcheck("ESI-MD-COPD T_low",            p4$t_low, 1.50, tol = 0.001),

  # ---- AFL-only-noCOPD risk, the load-bearing category ------------------
  selfcheck("MD-COPD AFL-only all-cause HR",
            R2("S2", "AFL-only", "all_HR"),  0.87, tol = 0.02),
  selfcheck("NoCT-MD-COPD AFL-only respiratory HR (should be highly elevated)",
            R2("S3", "AFL-only", "resp_HR"), 6.37, tol = 0.05),
  selfcheck("ESI-MD-COPD AFL-only all-cause HR (should straddle 1)",
            R2("S4", "AFL-only", "all_HR"),  0.96, tol = 0.02),

  # ---- Discrimination --------------------------------------------------
  selfcheck("MD-COPD all-cause C-index",
            DS("S2", "c_allcause"), 0.703, tol = 0.005)
)
selfcheck_tbl <- do.call(rbind, checks)
write.csv(selfcheck_tbl, file.path(OUT_DIR, "SELFCHECK.csv"), row.names = FALSE)
kable(selfcheck_tbl, caption = "Self-check -- computed vs canonical values.")
cat(sprintf("\nSelf-check summary: %d PASS, %d FAIL (of %d total)\n",
            sum(selfcheck_tbl$status == "PASS"),
            sum(selfcheck_tbl$status == "FAIL"),
            nrow(selfcheck_tbl)))
```

# Section 10 -- Completeness check

Enumerates every CSV downstream tables and figures depend on and aborts if
any is missing, so a partial run is caught here rather than surfacing as a
missing-file error deep in `draft_tables.R` or `build_manuscript.py`.

```{r completeness}
required_csvs <- c(
  "cohort_flow.csv", "cohort.txt", "mdcopd_validation.csv",
  "schema_labels.csv", "schema_labels.rds", "schema_risk.csv",
  "schema_crude.csv", "schema_discrimination.csv",
  "schema_fit.csv", "schema_fit_cv_diff.csv", "cv_schema_fits.csv",
  "objective_selection.csv", "reclassification.csv", "crossclass.csv",
  "supp_baseline.csv", "supp_esi_ct.csv",
  "discord_counts.csv", "discord_rates.csv", "discord_adjusted.csv",
  "schema_diff_bootstrap.csv",
  "gate_fixedratio.csv", "gate_exac_aic.csv", "SELFCHECK.csv"
)
missing <- required_csvs[!file.exists(file.path(OUT_DIR, required_csvs))]
if (length(missing))
  stop("missing artifacts in OUT_DIR: ", paste(missing, collapse = ", "))
cat(sprintf("All %d required artifacts present in %s\n",
            length(required_csvs), normalizePath(OUT_DIR)))
```

# Section 11 -- Provenance

```{r provenance}
provenance <- c(
  sprintf("rendered=%s", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")),
  sprintf("R=%s", R.version.string),
  sprintf("cohort_n=%d", nrow(d_b)),
  sprintf("cv_seed=%d", CV_SEED),
  sprintf("cv_repeats=%d", CV_REPEATS),
  sprintf("cv_folds=%d", CV_FOLDS),
  sprintf("crude_B=%d", CRUDE_B),
  sprintf("fitted_S3_k=%d", k3),
  sprintf("fitted_S4_k=%d", p4$k),
  sprintf("fitted_S4_t_low=%.3f", p4$t_low),
  sprintf("fitted_S4_t_high=%.3f", p4$t_high))
writeLines(provenance, file.path(OUT_DIR, "PROVENANCE.txt"))
cat(paste(provenance, collapse = "\n"), "\n")
```

# Appendix -- Session info

```{r session}
sessionInfo()
```
