---
title: "Why the ESI classification rule has been refitted"
subtitle: "Supporting detail for the CT-free MD-COPD manuscript"
output:
  html_document:
    theme: null
    highlight: null
    self_contained: true
    css: !expr NULL
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = "")
options(stringsAsFactors = FALSE)
# knitr sets the working directory to this file's own location, so the paths
# are derived from it directly rather than through _locate.R, whose commandArgs
# lookup would resolve against the render script and the pre-knit directory.
HERE   <- normalizePath(if (file.exists("_locate.R")) "." else "ctfree")
ROOT   <- dirname(HERE)
ASSETS <- file.path(HERE, "assets")
V15    <- file.path(ROOT, "manuscript_assets")

need <- function(p, f) { x <- file.path(p, f)
  if (!file.exists(x)) stop("missing artifact: ", x, "\nrun ctfree/analysis.R first")
  x }
t6  <- read.csv(need(V15, "Table_6.csv"), check.names = FALSE)
ag  <- readLines(need(V15, "Table_Agreement_stats.txt"))
xc  <- read.csv(need(ASSETS, "crossclass.csv"))
fit <- read.csv(need(ASSETS, "schema_fit.csv"))
rec <- read.csv(need(ASSETS, "reclassification.csv"))
rsk <- read.csv(need(ASSETS, "schema_risk.csv"))
stat <- function(k) as.numeric(sub("^[^=]+=", "", ag[grep(paste0("^", k, "="), ag)]))

CT <- c("noCOPD", "AFL-only-NoCOPD", "COPD-minor", "COPD-major")
M <- as.matrix(t6[, -1]); rownames(M) <- CT; colnames(M) <- CT
N <- sum(M)
O <- c("noCOPD", "AFL-only", "COPD-minor", "COPD-major")
K <- matrix(xc$n[xc$schema == "S4"], 4, 4, byrow = TRUE, dimnames = list(O, O))
pct <- function(x) sprintf("%.1f%%", 100 * x)
cm  <- function(x) format(x, big.mark = ",")
kb  <- function(d, ...) knitr::kable(d, format = "html", align = "lrrrrr",
                                     table.attr = 'class="t"', ...)
```

<style>
/* Explicit background and color: an attachment opened in an unknown client
   must not inherit a dark theme and become unreadable. */
html{background:#fff}
body{font-family:Helvetica,Arial,sans-serif;line-height:1.5;max-width:52em;
     margin:2em auto;padding:0 1em;color:#222;background:#fff}
h1{font-size:1.5em;border-bottom:2px solid #333;padding-bottom:.3em}
h2{font-size:1.15em;margin-top:2em;color:#111}
table.t{border-collapse:collapse;margin:1em 0;font-size:.92em}
table.t th{background:#f0f0f0;border-bottom:2px solid #999;padding:.4em .7em;text-align:right}
table.t th:first-child{text-align:left}
table.t td{border-bottom:1px solid #ddd;padding:.35em .7em;text-align:right}
table.t td:first-child{text-align:left}
.bad{color:#b00;font-weight:600}
.good{color:#060;font-weight:600}
.note{background:#f7f7f7;border-left:3px solid #999;padding:.6em 1em;margin:1.2em 0}
</style>

## 1. The agreement we reported, and what it measured

We reported **`r pct(sum(diag(M))/N)` agreement** and **kappa `r sprintf("%.2f", stat("kappa"))`**
between the CT-based MD-COPD classification and our ESI-based version. Those
figures are correct. They were computed at the binary level, COPD versus no
COPD, which the Methods stated. Agreement within each of the four diagnostic
categories was never reported.

## 2. Agreement within each category

Rows are the CT-based classification, columns the ESI-based one.

```{r xtab}
d <- as.data.frame.matrix(addmargins(M))
names(d)[5] <- "total"; d <- cbind(`CT-based` = c(CT, "total"), d)
kb(d, row.names = FALSE)
```

```{r diag}
dg <- data.frame(Category = CT,
                 Concordant = as.integer(diag(M)),
                 `CT-based n` = as.integer(rowSums(M)),
                 Agreement = pct(diag(M) / rowSums(M)), check.names = FALSE)
kb(dg[order(-diag(M)/rowSums(M)), ], row.names = FALSE)
```

The two large categories agree almost perfectly. The two that MD-COPD adds to
conventional spirometry do not: our rule reproduced
<span class="bad">`r pct(M["COPD-minor","COPD-minor"]/sum(M["COPD-minor",]))`</span>
of COPD-minor and
<span class="bad">`r pct(M["AFL-only-NoCOPD","AFL-only-NoCOPD"]/sum(M["AFL-only-NoCOPD",]))`</span>
of AFL-only-noCOPD.

<div class="note">
**Why the overall figure looks so good.** noCOPD and COPD-major are
`r pct((sum(M["noCOPD",])+sum(M["COPD-major",]))/N)` of the cohort and supply
`r pct((M["noCOPD","noCOPD"]+M["COPD-major","COPD-major"])/sum(diag(M)))` of all
concordance. Collapsing to a binary also merges COPD-minor with COPD-major and
AFL-only-noCOPD with noCOPD, so both disagreements above sit inside the merge
and cannot show.
</div>

## 3. Why the thresholds were never fitted for this

The ESI cut-points came from a classification tree fitted to predict the
**visual CT structural score**, not the MD-COPD category. The ten-candidate grid
that followed was scored on **binary** COPD agreement. Neither step included
per-category agreement, so the two categories where the rule performs worst
were outside what was optimized. This is a limit of scope in the original
fitting rather than an error in it.

## 4. The refit

Both thresholds were refitted over the full parameter space, using only the
CT-based labels, with no mortality or exacerbation data, and validated by five
repeats of five-fold cross-validation.

```{r fitted}
f <- fit[fit$schema %in% c("S4_v15draft", "S4"), ]
f <- f[order(f$schema != "S4_v15draft"), ]
kb(data.frame(
  Rule = c("Original: ESI ≥ 1.00, ≥ 3 of 4, two-threshold scoring",
           "Revised: ESI ≥ 1.25, ≥ 2 of 4"),
  `Held-out macro-F1` = sprintf("%.3f", f$macroF1_heldout), check.names = FALSE),
  row.names = FALSE)
```

The count threshold moves from three to two because dropping the two CT
criteria and adding one ESI criterion changes the scale of the count: three was
calibrated for five criteria, not four. The revised rule is also simpler, a
single ESI threshold in place of the 0/1/2 scoring, which we found was doing
almost no work.

```{r gain}
g <- data.frame(
  Category = c("COPD-minor", "AFL-only-noCOPD"),
  Original = c(sprintf("%d of %d (%s)", M["COPD-minor","COPD-minor"],
                       sum(M["COPD-minor",]), pct(M["COPD-minor","COPD-minor"]/sum(M["COPD-minor",]))),
               sprintf("%d of %d (%s)", M["AFL-only-NoCOPD","AFL-only-NoCOPD"],
                       sum(M["AFL-only-NoCOPD",]), pct(M["AFL-only-NoCOPD","AFL-only-NoCOPD"]/sum(M["AFL-only-NoCOPD",])))),
  Revised = c(sprintf("%d of %d (%s)", K["COPD-minor","COPD-minor"],
                      sum(K[, "COPD-minor"]), pct(K["COPD-minor","COPD-minor"]/sum(K[,"COPD-minor"]))),
              sprintf("%d of %d (%s)", K["AFL-only","AFL-only"],
                      sum(K[, "AFL-only"]), pct(K["AFL-only","AFL-only"]/sum(K[,"AFL-only"])))))
kb(g, row.names = FALSE)
```

## 5. What still falls short

AFL-only-noCOPD remains the weakest category at
<span class="bad">`r pct(K["AFL-only","AFL-only"]/sum(K[,"AFL-only"]))`</span>,
and the manuscript says so. It is `r cm(sum(K[, "AFL-only"]))` participants,
`r pct(sum(K[,"AFL-only"])/N)` of the cohort, and the interval on any estimate
within it is correspondingly wide.

<div class="note">
**Why this matters clinically, and why ESI is needed at all.** For
`r sprintf("%.1f%%", rec$ct_only_pct[1])` of COPD-major, a CT finding is the only
minor criterion met. Drop the CT criteria without replacing them and
<span class="bad">`r cm(rec$major_to_aflonly[rec$schema=="S3"])`</span>
participants leave COPD-major for AFL-only-noCOPD. With ESI in their place that
falls to <span class="good">`r cm(rec$major_to_aflonly[rec$schema=="S4"])`</span>.
The category receiving them carries
`r sprintf("%.1f", rsk$resp_HR[rsk$schema=="S3" & rsk$category=="AFL-only"])`
times the respiratory mortality of its reference when CT is simply dropped, and
no significant excess when ESI replaces it.
</div>

<p style="color:#777;font-size:.85em;margin-top:2.5em">
Every value above is read at render time from the analysis artifacts in
<code>ctfree/assets/</code> and <code>manuscript_assets/</code>. Regenerate with
<code>Rscript ctfree/analysis.R</code> then
<code>Rscript ctfree/render_massimo_report.R</code>.
</p>
